### 2023-09-06 ms/ms.qmd
initial commit all files

---
title: Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)
format:
  pdf:
    keep-tex: true  
    include-in-header: 
      - preamble.tex
    toc: false
header-includes: |
  \addtokomafont{disposition}{\rmfamily}
  \RedeclareSectionCommand[
    font=\normalfont\Large]{section}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\bfseries]{subsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\itshape]{subsubsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize]{paragraph}
mainfont: Times New Roman
bibliography: stomata-ilima.bib
csl: https://www.zotero.org/styles/american-journal-of-botany
---

\begin{center}
Genevieve Triplett$^1$, Thomas N. Buckley$^2$, Christopher D. Muir$^{1,3,*}$
\end{center}

$^1$ School of Life Sciences, University of Hawaiʻi Mānoa, Honolulu, HI, USA 96822

$^2$ School of Life Sciences, University of California, Davis, CA, USA 95616

$^3$ Department of Botany, University of Wisconsin, Madison, WI, USA 53706

$^*$ Corresponding author: Christopher D. Muir, cdmuir@wisc.edu

Manuscript received _______; revision accepted _______.

Running head: Amphistomy advantage in ʻilima

\newpage

```{=html}

<!-- TO DO,

1. at end of INTRODUCTION, address Tom's comments

"suggest you rigorously define sun leaves and shade leaves here, and dispense with the quotes."

Brief Communications—Brief Communications are short (c 3000–4000 word) research articles reporting exciting, significant new findings. They include no more than 4 figures and tables, combined. Manuscripts, and their abstracts, should be organized as described for Research articles. Authors must adhere to the same membership and page charge requirements as for regular Research Articles.

a. Title and Running Heads
The manuscript title for research papers should be specific and informative, conveying the key findings of the research in an active voice. Provide a title written with sentence-style capitalization. Below the title, list authors: each author’s first name, middle initial, surname. On the next line, give affiliation and unabbreviated address. If authors have different affiliations and addresses, add a superscript number after each author’s name to indicate the footnoted address. Include another footnote superscript number to indicate the author for correspondence.

Include the line: Manuscript received _______; revision accepted _______.

After this text, add a short title no longer than 65 characters (including spaces) to be used as a running head in the article.

b. Abstract
AJB requires structured abstracts for Research Articles and Brief Communications. The abstract is 250 words or fewer, written in the following structured format:
• Premise of the study (why the work was done, what major questions of plant biology are addressed, and why it is important to the broad AJB readership)
• Methods
• Key results
• Conclusions (what major points should the reader take from this article)

The abstract may be all a reader sees initially, so it should be written to capture the interest of the general botanical community as well as specialists. Avoid references; if essential, cite parenthetically with journal name, volume number, pages, and year.
-->
```
```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(brms)
library(dplyr)
library(ggdist)
library(glue)
library(readr)
library(stringr)
library(tidyr)

licor = read_rds("../processed-data/licor.rds")

# Function to print estimates and CIs
pretty_est = function(estimate, lower, upper, ...) {
  .e = prettyNum(estimate, ...)
  .l = prettyNum(lower, ...)
  .u = prettyNum(upper, ...)
  # glue("{.e}~[{.l}-{.u}]")
  glue("{.e}~[\\numrange{{{.l}}}{{{.u}}}]")
}

# From 04_fit-licor.R
fit_licor = read_rds("../objects/fit_licor.rds")
dim_fit_licor = dim(as.array(fit_licor$fit))

# From 05_fit-traits.R
fit_stomata = read_rds("../objects/fit_stomata.rds")
fit_thickness = read_rds("../objects/fit_thickness.rds")

dim_fit_stomata = dim(as.array(fit_stomata$fit))
dim_fit_thickness = dim(as.array(fit_thickness$fit))

habitat_thickness = fit_thickness |>
  as_draws_df(c("b_Intercept", "b_site_typemontane")) |>
  transmute(
    coastal = exp(b_Intercept),
    montane = exp(b_Intercept + b_site_typemontane),
    diff = coastal - montane
  ) |>
  point_interval(.width = 0.95, .point = median, .interval = qi)

thickness_diff = pretty_est(habitat_thickness$diff,
                     habitat_thickness$diff.lower,
                     habitat_thickness$diff.upper,
                     digits = 2L)

# From 08_plot-habitat-aa.R
# Estimated difference in AA between habitats
habitat_aa = read_rds("../objects/habitat_aa.rds") |>
  pivot_wider(names_from = site_type, values_from = AA) |>
  mutate(AA_diff = coastal - montane, times_increase = coastal / montane) |>
  point_interval(coastal, montane, AA_diff, times_increase,
                 .width = 0.95, .point = median, .interval = qi)

AA_coastal = pretty_est(habitat_aa$coastal,
                        habitat_aa$coastal.lower,
                        habitat_aa$coastal.upper,
                        digits = 2L)

AA_montane = pretty_est(habitat_aa$montane,
                        habitat_aa$montane.lower,
                        habitat_aa$montane.upper,
                        digits = 2L)

AA_diff = pretty_est(habitat_aa$AA_diff,
                     habitat_aa$AA_diff.lower,
                     habitat_aa$AA_diff.upper,
                     digits = 2L)

habitat_aa1 = read_rds("../objects/habitat_aa1.rds") |>
  pivot_wider(names_from = site_type, values_from = AA) |>
  mutate(AA_diff = coastal - montane, times_increase = coastal / montane) |>
  point_interval(coastal, montane, AA_diff, times_increase,
                 .width = 0.95, .point = median, .interval = qi)

AA_diff1 = pretty_est(habitat_aa1$AA_diff,
                     habitat_aa1$AA_diff.lower,
                     habitat_aa1$AA_diff.upper,
                     digits = 2L)

# From 09_plot-thickness-aa.R
coef_thickness_aa = read_rds("../objects/coef_thickness_aa.rds") |>
  point_interval(slope, .width = 0.95, .point = median, .interval = qi)

thickness_aa = pretty_est(
  coef_thickness_aa$slope,
  coef_thickness_aa$.lower,
  coef_thickness_aa$.upper,
  digits = 2L
)

# From 10_calculate-gmaxratio.R
gmaxratio_sitetype = read_rds("../objects/gmaxratio_sitetype.rds") |>
  select(-lower_gmax, -upper_gmax) |>
  pivot_wider(names_from = "habitat", values_from = "gmax_ratio") |>
  mutate(d = coastal - montane) |>
  point_interval(d, .width = 0.95, .point = median, .interval = qi) |>
  with(pretty_est(d, .lower, .upper, digits = 2L))

# From 12_plot-gmaxratio-aa.R
coef_gmaxratio_aa = read_rds("../objects/coef_gmaxratio_aa.rds") |>
  point_interval(slope, .width = 0.95, .point = median, .interval = qi)

gmaxratio_aa = pretty_est(
  coef_gmaxratio_aa$slope,
  coef_gmaxratio_aa$.lower,
  coef_gmaxratio_aa$.upper,
  digits = 2L
)

# From 13_plot-habitat-Ags.R
fit_habitat_Ags = read_rds("../objects/fit_habitat_Ags.rds")

# Calculation amount of gsw extrapolation
gsw_extrap = licor |>
  summarize(min = min(gsw), max = max(gsw), .by = c("site_code", "aperture")) |>
  pivot_wider(names_from = aperture, values_from = min:max) |>
  rowwise() |>
  mutate(gsw = ifelse(
    `max_two-sided` > `max_one-sided`,
    mean(c(`max_one-sided`, `min_two-sided`)),
    mean(c(`min_one-sided`, `max_two-sided`))
  )) |>
  transmute(site_code = site_code, d_gsw = (`min_two-sided` - `max_one-sided`) / 2) |>
  filter(d_gsw > 0) |>
  pull(d_gsw) %>%
  sprintf("%.2e", .) |>
  str_replace("^([0-9].[0-9]{2})e-0([0-9])", "\\1\\\\times 10^{-\\2}") |>
  str_c(collapse = "~\\text{and}~") %>%
  paste0("$", ., "~\\text{mol}~\\text{m}^{-2}~\\text{s}^{-1}$")

```

# ABSTRACT --- {#abstract}

\noindent \textbf{Premise of the study}  The adaptive significance of stomata on both upper and lower leaf surfaces, called amphistomy, is unresolved. A widespread association between amphistomy and open, sunny habitats suggests the adaptive benefit of amphistomy may be greatest in these contexts, but this hypothesis has not been tested experimentally. Understanding why amphistomy evolves can inform  its potential as a target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

\textbf{Key results}  Coastal ʻilima leaves benefit 4.58 times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.

\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts. The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic basis of this observation is an area for future research.

**Keywords:** amphstomy, leaf, light, Hawaiʻi, leaf, *Sida fallax*, stomata

# INTRODUCTION --- {#introduction}

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts (\autoref{fig:compare-circuit-diagrams}), yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through evapotranspiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

Where are amphistomatous leaves found and why aren't they more common? Among terrestrial flowering plants, amphistomatous leaves are rarely found on woody plants and shade-tolerant herbs, but they are common in annual and perennial herbs from sunny habitats [@salisbury_i_1928; @parkhurst_adaptive_1978; @mott_adaptive_1982; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. Even in resupinate leaves where the abaxial surface faces up toward the sky, stomata develop on the lower adaxial surface [@lyshede_comparative_2002]. Exceptions to this general pattern include some arid woody plants which typically have vertically oriented, isobilateral leaves [@wood_physiology_1934; @jordan_using_2014; @de_boer_apparent_2016; @drake_two_2019] and floating/amphibious leaves of aquatic plants [@kaul_anatomical_1976; @doll_callitriche_2021]. The dearth of amphistomatous leaves should be quite surprising and has been described as one of the most important unsolved problems in the study of leaf structure-function relations despite some recent progress [@grubb_leaf_1977; @grubb_leaf_2020].

Amphistomatous leaves should be common because, all else being equal, a leaf with a given number of stomata per area could increase its photosynthetic rate simply by apportioning approximately half its stomata to each surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @gutschick_photosynthesis_1984-1]. The key difference between a hypo- and amphistomatous leaf, holding all other factors constant, is that an amphistomatous leaf has two parallel diffusion paths through the internal airspace to any given chloroplast. Those airspaces pose a resistance for CO$_2$ diffusion, so CO$_2$ concentration drops as it approaches chloroplasts. Shorter pathways mean a smaller drop in CO$_2$ concentration. Thus, chloroplasts in amphistomatous leaves experience higher CO$_2$ concentrations than in hypostomatous leaves, thereby increasing photosynthesis. The airspace resistance (or its inverse, the airspace conductance, $g_\mathrm{ias}$) is rarely measured directly and there is disagreement between empirical [@parkhurst_intercellular_1990; @morison_lateral_2005; @evans_resistances_2009; @tomas_importance_2013; @earles_beyond_2018; @santrucek_partitioning_2019; @nobel_physicochemical_2020; @harwood_understanding_2021; @marquez_assessing_2023] and theoretical models [@tholen_mechanistic_2011; @ho_three-dimensional_2016; @theroux-rancourt_maximum_2021]. The $g_\mathrm{ias}$ in thin, porous leaves may be so large as to be inconsequential given much lower conductances for other components of the diffusion pathway, whereas the $g_\mathrm{ias}$ of thick leaves with little airspace may greatly hinder CO$_2$ diffusion to chloroplasts. Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$.

Amphistomy should also enhance photosynthesis when leaf boundary layer resistance is high, because apportioning total flux between two boundary layers rather than one results in a smaller CO$_2$ concentration drop between the atmosphere and stomata. A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019].

The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light. @mott_adaptive_1982 proposed that "plants with a high photosynthetic capacity, living in full-sun environments, and experiencing rapidly fluctuating or continuously available soil water" would benefit most, in terms of increased carbon gain, from having amphistomatous leaves. As described above, herbs from sunny habitats are often amphistomatous. Most variation in stomatal density ratio ($\mathrm{SR}$, the ratio of stomatal density between the upper and lower surfaces) among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light. Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop \~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991]. Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop \~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975]. Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of adaxial stomata increases from \~10-20% under low light to \~30-40% under high light [@james_leaf_2001]. In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], resulting in lower $g_\mathrm{ias}$ [@parkhurst_adaptive_1978]. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has very open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that takes into account differences in both stomatal density and size between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to test the relationship between light habitat and $\mathrm{AA}$. Many lineages have adapted to different light habitats after colonization and leaf anatomical traits such as $\mathrm{SR}$ and thickness vary within and among closely related species. It is hypothesized that the common ancestor in many Hawaiian clades was a "weedy" species with high dispersal ability adapted to open habitats [@carlquist_biota_1966]. Colonization was followed by adaptive radiation into higher elevation, montane, closed, forested habitats. Consequently, adaptation to sun and shade is a common axis of phenotypic variation among Hawaiian plants such as lobeliads [@givnish_adaptive_2004; @montgomery_adaptive_2008; @givnish_origin_2009; @givnish_common-garden_2014; @scoffoni_light-induced_2015], *Bidens* [@carlquist_biota_1966; @knope_dispersal_2020], *Scaevola* [@robichaux_evolution_1984; @mckown_trait_2016], *Euphorbia* [@sporck_hawaiian_2011], and *Plantago* [@dunbar-co_leaf_2009].

Here we focus on variation within an indigenous plant species *Sida fallax* Walp. (Malvaceae), known in the Hawaiian language as ʻilima. ʻIlima is found from sea level to elevations $>1000$ mas on multiple Hawaiian islands. Coastal populations are morphologically different from montane populations (\autoref{fig:study-system}). Coastal regions of Hawaiʻi are characterized by high sun exposure, warmer temperatures, high winds, salinity, and variation in water availability. Coastal populations of ʻilima tend to be short and prostrate which likely helps them to withstand the windy environment (\autoref{fig:study-system}a). The leaves of these populations are covered on both surfaces in dense, soft hairs that give the leaves a silvery green appearance (\autoref{fig:study-system}b), which helps mitigate water loss by reflecting solar radiation, thereby lowering leaf temperature [@ehleringer_pubescence_1978]. Montane regions, on the other hand, provide very different challenges. Many other tall species grow on the slopes of these wet mountainous regions, which makes light competition a factor that plants may need to adapt to. Possibly due to this, montane populations are erect and shrub- or tree-like, capable of growing meters tall with strong, woody stems. These individuals have smooth, green foliage with serrated edges. Montane populations exhibit traits that help them to compete for light availability (Raunch et al. 1997). This montane morphology is not found in *S. fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022]. The variable morphology between coastal and montane populations has even led some scientists to argue that it makes up a species complex, rather than a single species (Raunch et al. 1997).

The objectives of our study are to test whether 1) 'sun' leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than 'shade' leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?

```{=html}
<!-- TEXT FROM PROPOSAL THAT MAY HAVE SOMETHING USEFUL:

*Knowing the adaptive significance of amphistomy will advance our understanding of leaf structure-function relationships, aid reconstructing paleoenvironments, and guide crop optimization*. Stomata are studied in disparate fields such as cell biology, ecosystem ecology, and paleontology [@hetherington_role_2003; @bergmann_stomatal_2007; @berry_stomata:_2010; @mcelwain_paleoecology_2017]. The intense focus on stomata makes it all the more surprising that we know relatively little about the adaptive significance of where stomata are on the leaf! Amphistomy is deeply integrated with leaf traits associated with high photosynthetic capacity and leaf water transport capacity [@smith_leaf_1997; @drake_two_2019; @xiong_one_2020]. Knowing the upper limits of photosynthesis will help breed more productive crops [@long_can_2006] and predict future terrestrial carbon sinks at higher atmospheric CO$_2$ [@keenan_terrestrial_2018]. In paleontology, stomata visible on fossil leaves help reconstruct paleo-CO$_2$ concentrations based on density/index [@konrad_modelling_2008; @franks_co2-forced_2009; @mcelwain_paleoecology_2017; but see @jordan_links_2020], but stomatal density ratio ($\mathrm{SR}$) may add information about whether plants lived in open or closed vegetation [@salisbury_i_1928; @jordan_using_2014; @bucher_stomatal_2017]. Finally, $\mathrm{SR}$ tends to increase (higher proportion of stomata on upper surface) during domestication [@milla_shifts_2013], but we do not know why. Our work will help infer how changes in $\mathrm{SR}$ and other leaf anatomical traits affect photosynthesis and water use in crops. This is particularly promising because we may learn ways to increase leaf photosynthesis without increasing water loss by altering $\mathrm{SR}$ and/or the regulation of stomatal aperture on each leaf surface.

To redress these gaps in our knowledge, in Aim 1 we will model the benefits and costs of amphistomy under low- and high-light conditions and in Aims 2-3 we will test model predictions using the wild tomatoes as model clade (*Solanum* spp.; Solanaceae). Specifically, we will experimentally manipulate whether gas exchange proceeds through one or both leaf surfaces, holding all else constant. This will enable us to measure how much amphistomy increases photosynthesis compared to an otherwise identical hypostomatous leaf (Aim 2). In Aim 3, we will test whether amphistomy is costly in shade leaves because it causes dehydration of the palisade mesophyll under high evaporative demand. The result of this project will determine whether high light, photosynthetic capacity, and adequate hydration are sufficient to explain the distribution of amphistomatous leaves in nature, as Mott and colleagues hypothesized 40 years ago.

The dearth of amphistomatous leaves implies a fitness cost associated with stomata on the upper leaf surface in particular [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984-1; @muir_is_2019]. Two nonmutually exclusive hypotheses for the cost of upper stomata have received the most attention, but neither has been tested experimentally. We will test one hypothesis, which is that amphistomy hinders photosynthesis by causing dehydration of the palisade mesophyll under high evaporative demand [@buckley_sites_2017; @drake_two_2019]. This cost of amphistomy could be greater in shade leaves because they dehydrate severely when exposed to intense sunflecks, warm temperatures, and/or low humidity, whereas sun leaves have constitutive traits that mitigate the risk of dehydration. Sensitivity to dehydration would be more consequential if it affected the palisade mesophyll, which is adjacent to the illuminated upper surface and is thus more photosynthetically active [@smith_leaf_1997; @evans_profiles_2003; @earles_embracing_2019] than the spongy mesophyll, which is adjacent to the lower surface (\autoref{fig:compare-circuit-diagrams}). An alternative hypothesis with some empirical support is that stomata on the upper surface are more likely to be colonized by bacterial or fungal pathogens, presumably because a propagule is more likely to land on the upper surface [@gutschick_photosynthesis_1984-1; @mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. The current proposal does not address this pathogen susceptibility hypothesis because PI Muir is seeking funding for this work elsewhere. Although not the focus of our aims, our modeling and data can address two less well theorized alternative hypotheses that upper stomata compete for space on the epidermis [@de_boer_optimal_2016; @haworth_allocation_2018; @muir_how_2023] or alter palisade mesophyll structure [@dow_disruption_2017; @lundgren_mesophyll_2019; @baillie_developmental_2020].

*Despite the fact that we do not know the adaptive significance of amphistomy, the distribution of hypo- and amphistomatous leaves is nonrandom and most likely caused by natural selection.* Consistent relationships between stomatal traits, growth form, and light environment strongly suggest convergent evolution, a hallmark of adaptive evolution. Hypo- and amphistomatous leaves are two ends of a continuum in $\mathrm{SR}$. Hypostomatous leaves have a $\mathrm{SR}$ close to 0; fully amphistomatous leaves with equal densities on each leaf have a $\mathrm{SR}$ close to 0.5. Herbaceous plants from high-light habitats tend to be amphistomatous and have higher $\mathrm{SR}$ values than herbaceous plants from shaded habitats or woody species [@salisbury_i_1928; @parkhurst_adaptive_1978; @mott_adaptive_1982; @parkhurst_intercellular_1990; @peat_comparative_1994; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. A second line of evidence for adaptation is that the distribution of $\mathrm{SR}$ among species is inconsistent with the expected distribution for a ratio trait evolving neutrally; instead, a nonneutral bimodal distribution evolves repeatedly in land plants [@muir_making_2015]. Finally, amphistomatous leaves are strongly coordinated with other traits involved with high photosynthetic capacity [@beerling_evolutionary_1996; @xiong_one_2020] and domestication [@milla_shifts_2013], suggesting that amphistomy is favored by selection in some circumstances. What specific fitness costs and benefits underlie these adaptive patterns?

*The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light*. To reiterate, herbaceous leaves in high light habitats tend to be amphistomatous. Most variation in $\mathrm{SR}$ among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light. Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop ~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991]. Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop ~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975]. Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of upper stomata increases from ~10-20% under low light to ~30-40% under high light [@james_leaf_2001]. In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.

*We don’t know whether amphistomy is an adaptation to high light intensity because no one has shown experimentally that amphistomy* per se *increases CO$_2$ assimilation more under high light than under low light*. Other factors correlated with light habitat could drive selection on $\mathrm{SR}$ and result in a spurious association between light habitat and $\mathrm{SR}$. Even if light intensity is the primary causal selective agent for $\mathrm{SR}$, its effect on selection for higher/lower $\mathrm{SR}$ could be the result of 1) *facultative* responses to light intensity and/or 2) *constitutive* differences between shade and sun leaves (\autoref{fig:causal-graph}). The *facultative* hypothesis is that amphistomy increases photosynthesis under high light intensity because of high CO$_2$ demand and/or CO$_2$ supply is constrained by low $g_\text{ias}$. Under this hypothesis, the advantage of amphistomy depends on the immediate light intensity, not the light environment in which plants develop. The nonmutually exclusive *constitutive* hypothesis is that developmental plasticity results in 'sun' and 'shade' leaves with manifold differences in anatomy and biochemistry such as thicker leaves, denser cell packing, or increased Rubisco concentration [@terashima_why_2001; @lambers_plant_2008]. Sun leaves benefit more from amphistomy than shade leaves under all light intensities because they have lower $g_\text{ias}$ and/or greater CO$_2$ demand. The proposed research will quantify the relative contributions of facultative and constitutive effects on the photosynthetic advantage of amphistomy. We will describe predictions of these hypotheses in more detail below.

MOVE THIS TO INTRO Parkhurst (1978) introduced the concept of “amphistomy advantage” to quantify the photosynthetic advantage of an amphistomatous leaf over an otherwise identical hypostomatous leaf. An important limitation of his approach is that he assumed photosynthesis was simply proportional to the total diffusive conductance from the atmosphere to the sites of carboxylation; that tacitly assumes CO$_2$ concentration at the sites of carboxylation is zero, which in turn infinite local did not model assumes CO$_2$ demand is locally infinite throughout the leaf. In reality, photosynthesis at any given point in the leaf depends on the local concentrations of CO$_2$, chlorophyll, and Rubisco, and the local irradiance. Thus, some mesophyll layers can draw down chloroplastic CO$_2$ concentration to a low level and potentially assimilate a lot of carbon when CO$_2$ is adequately supplied, whereas other layers have much weaker CO$_2$ demand. By not incorporating a CO$_2$ demand function, Parkhurst assumed that any increase in CO$_2$ supply was matched by a concomitant increase in CO$_2$ assimilation. We now know that increasing assimilation depends on both CO$_2$ supply and demand. If supply increases, but the demand function is flat because of low light intensity and/or low photosynthetic capacity, then $A$ won’t change much (demand-limited). Conversely, if the demand function is steep because there is excess light and/or enzymatic capacity, then increased CO$_2$ supply will yield a large increase in $A$ (supply-limited).

 -->
```
```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}
```
# MATERIALS AND METHODS --- {#methods}

## Plant sampling --- {#plant-sampling}

We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting [@yorkston_interfertility_2006] and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites that appeared to be cultivated. We visited sites between August and November 2022. For logistical reasons, the sites on Hawaiʻi were sampled in one three-day trip. We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site. For safety and conservation reasons, transects were along a trail or road. We did not sample small individuals if there was risk removing leaves would cause mortality. From each plant, we collected two fully expanded leaves for traits measurements. We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.

```{r, locations, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-locations
#| tbl-cap: "ʻIlima study sites for this study."
#| 
read_rds("../processed-data/site.rds") |>
  filter(site_code != "nnpl") |>
  mutate(Island = case_when(
    island == "oahu" ~ "Oʻahu",
    island == "hawaii" ~ "Hawaiʻi"
  )) |>
  arrange(site_type, site) |>
  select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
         Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
  knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
               booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "1.5cm") |>
  kableExtra::column_spec(3, width = "1.5cm") |>
  kableExtra::column_spec(4, width = "1.5cm") |>
  kableExtra::column_spec(5, width = "1.5cm") |>
  kableExtra::column_spec(6, width = "1.5cm")

```

## Leaf traits --- {#leaf-traits}

### Stomata --- {#stomata}

We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves. For pubescent leaves (usually coastal), we dried and pressed leaves for $\approx 1$ week [@hill_temperature_2014], carefully scraped trichomes off with a razor blade, and rehydrated the leaf. For glabrous leaves, we used fresh leaves. We applied clear nail polish to both leaf surfaces of fresh or rehydrated leaves in the middle of the lamina away from major veins. After nail polish dried, we mounted impressions on a microscope slide using transparent tape [@mott_amphistomy_1991]. We digitized a portion of each leaf surface impression using a brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted all stomata and divided by the visible leaf area (0.890 mm$^2$) to estimate density and measured guard cell length from five randomly chosen stomata per field using ImageJ [@schneider_nih_2012].

### Leaf thickness --- {#leaf-thickness}

We cut thin sections using two razor blades taped together and cutting the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.

### Gas exchange measurements --- {#gas-exchange-measurements}

At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA). We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$. We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification. For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values, holding other environmental variables constant. We refer to this as an $A \textendash g_\text{sw}$ curve. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of "pseudohypostomatous" leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm). We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

$$\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})$$

The log-response ratio is easy to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. Because the propafilm reduces transmission, we used a higher incident $\mathrm{PPFD}$ for pseudohypo leaves (higher by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm). We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

## Data analysis --- {#data-analyis}

### Objective 1: Do coastal leaves have greater amphistomy advantage than montane leaves? ---

It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$. To overcome this, we fit $A \textendash g_\text{sw}$ curves using a flexible generalized additive model (GAM) to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$. In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\mathrm{AA}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. We fit GAMs using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] for Bayesian statistical inference with MCMC sampling in *Stan* [@stan_development_team_stan_2023]. We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023]. We sampled from posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$.

### Objective 2a: Are coastal leaves thicker than montane leaves? ---

We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with site-type as a fixed effect and individual plant and site as random effects. We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023]. We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations. We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population. We used this approach most of the variation in leaf thickness was among sites (see Results) and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation. We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive.

### Objective 2b: Is $g_{\mathrm{smax,ratio}}$ closer to 0.5 in coastal leaves than montane leaves? ---

We tested whether $g_\mathrm{smax,ratio}$ varied between coastal and montane populations and among individuals within populations using a Bayesian mutliresponse, mixed-effects model. The modeled response variables are stomatal count and guard cell length on each surface. Counts were modeled as negative binomially distributed variable from a latent stomatal density and a parameter $\phi$ to estimate overdispersion in counts relative to a Poisson model. For all traits, the explanatory variables were site-type as a fixed effect and leaf within individual plant, individual plant, and site as random effects. We used the *R* package **brms** version `r fit_stomata$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_stomata$version$cmdstan` and **cmdstanr** version `r fit_stomata$version$cmdstanr` [@gabry_cmdstanr_2023]. We interpolated missing adaxial guard cell lengths from `r length(which(fit_stomata$data$upper_number_of_stomata == 0))` out of `r nrow(fit_stomata$data)` samples with zero adaxial stomata using the "mi" function in **brms** package. We sampled the posterior distribution from `r dim_fit_stomata[2]` chains with `r dim_fit_stomata[1]` iterations each after `r dim_fit_stomata[1]` warmup iterations. From each posterior sample, we calculated $g_{\mathrm{smax,ratio}}$ as:

$$g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},$$ where $g_\mathrm{smax,lower}$ and $g_\mathrm{smax,upper}$ are maximum stomatal conductance to water vapor at $T_\mathrm{leaf} = 25 ^ \circ$ C on the lower and upper surface, respectively. The maximum stomatal conductance is calculated from stomatal density and length, assuming that stomata are fully open, following @sack_developmental_2016.

<!-- They defined a biophysical and morphological constant as: -->

<!-- \begin{align*} -->

<!--   b = & D_\text{wv} / v \\ -->

<!--   m = & \frac{\pi c ^ 2}{j ^ {0.5} (4 h j + \pi c)} -->

<!-- \end{align*} -->

<!-- $b$ is the diffusion coefficient of water vapor in air ($D_\text{wv}$) divided by the kinematic viscosity of dry air ($v$). $D_\text{wv} = 2.49 \times 10^{-5} ~ \text{m}^2 ~ \text{s}^{-1}$ and $v = 2.24 \times 10 ^ {-2} ~ \text{m}^3 ~ \text{mol} ^ {-1}$ at $25 ^ \circ$ [@monteith_principles_2013]. For kidney-shaped guard cells, $c = h = j = 0.5$. -->

We estimated the relationship between leaf $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ measured from a single leaf per population. We propagated uncertainty about in $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$.

# RESULTS --- {#results}

## Amphistomy advantage is greater in coastal leaves ---

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants. $\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual curves). Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$.  The difference between coastal and montane leaves is lower,  $`r AA_diff1`$, but still greater than 0 if we remove two sites where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves. Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}).

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}
```
## Leaf thickness is associated with amphistomy advantage between but not within habitats ---

Coastal ʻilima leaves are $`r thickness_diff`$ $\mu\text{m}$ thicker than their montane counterparts. Although coastal leaves are thicker and have greater $\mathrm{AA}$, there is little relationship between leaf thickness and $\mathrm{AA}$ within habitats (\autoref{fig:traits-aa}A; $\text{slope}~= `r thickness_aa`$).

## $g_\mathrm{smax,ratio}$ is not associated with amphistomy advantage ---

Coastal and montane leaves have similar average $g_\mathrm{smax,ratio}$, the ratio of adaxial (upper) to total $g_\mathrm{smax}$, the anatomical maximum stomatal conductance to water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have $`r gmaxratio_sitetype`$ higher $g_\mathrm{smax,ratio}$ than montane leaves, but the 95% confidence intervals overlap 0 difference. The $g_\mathrm{smax,ratio}$ is somewhat bimodal among sites. Some sites in both habitats have leaves with $g_\mathrm{smax,ratio} < 0.07$ and others with $g_\mathrm{smax,ratio} > 0.2$ (\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in montane sites where those on the Big Island of Hawaiʻi all have low $g_\mathrm{smax,ratio}$ whereas those on Oʻahu have relatively high $g_\mathrm{smax,ratio}$. There is no relationship between $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ in either habitat (\autoref{fig:traits-aa}B; $\text{slope}~= `r gmaxratio_aa`$) in our sample.

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) $g_\mathrm{smax,ratio}$ and (B) leaf thickness among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi., and amphistomy advantage. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}
```

# DISCUSSION --- {#discussion}

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the ‘amphistomy advantage’ ($\mathrm{AA}$) holding all else constant. Taking advantage of the of the steep climatic gradients in the Hawaiian archipegalo, we applied this new method to show for the first time that sun leaves benefit more from amphistomy than shade leaves in ʻilima (*Sida fallax*). While this result has not yet been validated in other speciesthere are limits to how much we can generalize to all land plants, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

A second natural hypothesis is that amphistomatous leaves with few adaxial (upper) stomata benefit less than those with similar densities on both surfaces. We predicted that leaves with $g_\mathrm{smax,ratio}$ closer to 0.5 would have higher $\mathrm{AA}$ based on biophysical models [@gutschick_photosynthesis_1984]. The logic is that a small number of stomata on the upper surface are insufficient to supply the entire upper mesophyll due to limited lateral diffusion [@morison_lateral_2005]. Our results do not support this hypothesis. Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-AA}B). Among coastal sites, the site with the lowest $g_\mathrm{smax,ratio}$ had the lowest $\mathrm{AA}$, but there was little variation in $g_\mathrm{smax,ratio}$ among coastal leaves in our sample. We therefore cannot rule out that a larger sample of coastal leaves with greater variance in $g_\mathrm{smax,ratio}$ might support this hypothesis.

Two major implications from our study are that 1) photosynthesis in hypostomatous leaves is likely limited by CO$_2$ concentration drawdown within leaf airspaces; and 2) amphistomy *per se* contributes to, but is not wholly responsible for, higher photosynthetic rates among amphistomatous leaves. The amphistomy advantage we observe in ʻilima leaves implies decreased CO$_2$ supply in pseudohypostomatous leaves because of concentration drawdowns in the leaf airspace. Limited diffusion through the airspace has long been hypothesized to depress photosynthesis in hypostomatous leaves [@parkhurst_diffusion_1994], with empirical support from helox studies [@parkhurst_intercellular_1990]. However, these studies relied on interspecific comparisons of amphi- and hypostomatous leaves that differ systematically in many traits that affect gas exchange and photosynthesis [@xiong_one_2020]. Our experimental approach overcomes this limitation and implies that the drop in CO$_2$ concentration from substomatal cavities to the upper surface depresses photosynthesis.

Among land plants grown in a common garden, amphistomatous leaves have on average nearly $2\times$ higher area-based photosynthetic rates [@xiong_one_2020], naively implying an $\mathrm{AA} \approx \text{log}~2 = 0.69$. This is much higher than our estimate of `r round(habitat_aa$coastal, 2)` among coastal ʻilima leaves. The most likely explanation is that amphistomy is not the only cause of high photosynthetic rate. Indeed, species adapted to open, high light habitats with amphistomatous leaves also have higher concentrations of Rubisco, overall stomatal conductance, and photosynthetic capacity [@smith_leaf_1997; @xiong_one_2020]. For a leaf with high photosynthetic capacity that is well illuminated and hydrated, the major limitation becomes CO$_2$. Under these conditions, amphistomy may substantially increase photosynthesis, as we observe in coastal ʻilima leaves. Selection on increased photosynthesis under similar conditions may explain why crops leaves tend to increase stomatal density ratio during domestication [@milla_shifts_2013].

Two limitations of this study are the small sample sample sizes and experimental design that precludes distinguishing genetic from environmental differences in leaf traits. Understanding the mechanistic basis of higher $\mathrm{A}A$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be useful understand genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.

# CONCLUSIONS --- {#conclusions}

This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common. By developing a straightforward gas exchange method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant. In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest. We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.

# Acknowledgments --- {#acknowledgments}

<!-- add tom buckley as coauthor or acknowledge help developing method -->

The authors thank Kasey Barton for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.

# Author Contributions --- {#author-ontributions}

GT and CDM contributed equally to all stages of this project; TNB contributed to development of the method and helped edit the manuscript.

# Data Availability Statement ---

Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo with a DOI and stable URL upon publication. Raw data will be deposited on Dryad with a DOI and stable URL upon publication.

# Supporting Information ---

Additional supporting information may be found online in the Supporting Information section at the end of the article

* Appendix S1: Supplemental figures and table

<!-- Placeholders for supplemental figures -->

```{=tex}
\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\renewcommand\theequation{S\arabic{equation}}    
\setcounter{figure}{0}    
\setcounter{table}{0}    
\setcounter{equation}{0}
```
```{=tex}
\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:licor}
\end{figure}
```
```{=tex}
\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:habitat-Ags}
\end{figure}
```
```{=tex}
\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:habitat-gmaxratio}
\end{figure}
```
```{=tex}
\begin{table}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{tab:sum-data}
\end{table}
```

# Literature Cited --- {#literature-cited}

### 2023-09-06 ms/ms.tex
initial commit all files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{hyperref}

% customize figure references
\renewcommand{\figureautorefname}{Fig.}

% toggle line numbers
\usepackage{lineno}
\linenumbers

% toggle spacing
\usepackage{setspace}
% \singlespacing
\doublespacing

% Hawaiian language characters
% https://tex.stackexchange.com/questions/424535/how-to-type-a-proper-hawai%CA%BBian-%CA%BBokina
\usepackage[utf8]{inputenc}
\usepackage{newunicodechar,graphicx}

\DeclareRobustCommand{\okina}{%
  \raisebox{\dimexpr\fontcharht\font`A-\height}{%
    \scalebox{0.8}{`}%
  }%
}
\newunicodechar{ʻ}{\okina}

% make better hyphen for numeric ranges, \numrange{}{}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\addtokomafont{disposition}{\rmfamily}
\RedeclareSectionCommand[
  font=\normalfont\Large]{section}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\bfseries]{subsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize\itshape]{subsubsection}
\RedeclareSectionCommand[
  font=\normalfont\normalsize]{paragraph}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (Sida fallax)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Amphistomy increases leaf photosynthesis more in coastal than
montane plants of Hawaiian ʻilima (\emph{Sida fallax})}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi

\begin{center}
Genevieve Triplett$^1$, Thomas N. Buckley$^2$, Christopher D. Muir$^{1,3,*}$
\end{center}

\(^1\) School of Life Sciences, University of Hawaiʻi Mānoa, Honolulu,
HI, USA 96822

\(^2\) School of Life Sciences, University of California, Davis, CA, USA
95616

\(^3\) Department of Botany, University of Wisconsin, Madison, WI, USA
53706

\(^*\) Corresponding author: Christopher D. Muir, cdmuir@wisc.edu

Manuscript received \_\_\_\_\_\_\_; revision accepted \_\_\_\_\_\_\_.

Running head: Amphistomy advantage in ʻilima

\newpage

\hypertarget{abstract}{%
\section{ABSTRACT ---}\label{abstract}}

\noindent \textbf{Premise of the study} The adaptive significance of
stomata on both upper and lower leaf surfaces, called amphistomy, is
unresolved. A widespread association between amphistomy and open, sunny
habitats suggests the adaptive benefit of amphistomy may be greatest in
these contexts, but this hypothesis has not been tested experimentally.
Understanding why amphistomy evolves can inform its potential as a
target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf with the same total
stomatal conductance but with gas exchange blocked through the upper
(adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated
\(\mathrm{AA}\) and related physiological and anatomical traits in 12
populations, six coastal (open, sunny) and six montane (closed, shaded),
of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4.58 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and that leaves with more
similar conductances at each surface benefit more than leaves with few
stomata on the upper surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy \emph{per se} increases photosynthesis, consistent with the
hypothesis that parallel pathways through upper and lower mesophyll
increase the supply of CO\(_2\) to chloroplasts. The prevalence of
amphistomatous leaves in open, sunny habitats can partially be explained
the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic
basis of this observation is an area for future research.

\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, leaf, \emph{Sida
fallax}, stomata

\hypertarget{introduction}{%
\section{INTRODUCTION ---}\label{introduction}}

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts
(\autoref{fig:compare-circuit-diagrams}), yet paradoxically this
seemingly simple adaptation is uncommon in nature and we don't know why.
Understanding variation in stomatal traits like amphistomy is imperative
because these tiny pores play an outsized ecological role in the global
carbon and water cycles (Hetherington and Woodward, 2003; Berry et al.,
2010). A widely applicable, accurate representation of how stomata
mediate the relationship between CO\(_2\) gained through photosynthesis
and water lost through evapotranspiration is essential to predict future
climate using Earth Systems Models (Jarvis, 1976; Ball et al., 1987;
Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997). Optimality
models accurately predict the major cause of water loss, stomatal
conductance (\(g_\mathrm{sw}\)), by assuming plants maximize carbon gain
minus a cost of water (Cowan and Farquhar, 1977; Givnish, 1986; Medlyn
et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks et al., 2018;
Deans et al., 2020; Franklin et al., 2020; Wang et al., 2020; Harrison
et al., 2021). Despite the success of optimality modeling in predicting
\(g_\mathrm{sw}\), the same modeling approach has so far failed to
explain the rarity of amphistomatous leaves (Muir, 2019). \textbf{This
gap between theory and observations strongly implies that we remain
ignorant about some key benefits and costs associated with stomata.}

Where are amphistomatous leaves found and why aren't they more common?
Among terrestrial flowering plants, amphistomatous leaves are rarely
found on woody plants and shade-tolerant herbs, but they are common in
annual and perennial herbs from sunny habitats (Salisbury, 1928;
Parkhurst, 1978; Mott et al., 1982; Peat and Fitter, 1994; Gibson, 1996;
Jordan et al., 2014; Muir, 2015, 2018; Bucher et al., 2017). Even in
resupinate leaves where the abaxial surface faces up toward the sky,
stomata develop on the lower adaxial surface (Lyshede, 2002). Exceptions
to this general pattern include some arid woody plants which typically
have vertically oriented, isobilateral leaves (Wood, 1934; Jordan et
al., 2014; Boer et al., 2016; Drake et al., 2019) and
floating/amphibious leaves of aquatic plants (Kaul, 1976; Doll et al.,
2021). The dearth of amphistomatous leaves should be quite surprising
and has been described as one of the most important unsolved problems in
the study of leaf structure-function relations despite some recent
progress (Grubb, 1977, 2020).

Amphistomatous leaves should be common because, all else being equal, a
leaf with a given number of stomata per area could increase its
photosynthetic rate simply by apportioning approximately half its
stomata to each surface (Parkhurst, 1978; Gutschick, 1984a, b). The key
difference between a hypo- and amphistomatous leaf, holding all other
factors constant, is that an amphistomatous leaf has two parallel
diffusion paths through the internal airspace to any given chloroplast.
Those airspaces pose a resistance for CO\(_2\) diffusion, so CO\(_2\)
concentration drops as it approaches chloroplasts. Shorter pathways mean
a smaller drop in CO\(_2\) concentration. Thus, chloroplasts in
amphistomatous leaves experience higher CO\(_2\) concentrations than in
hypostomatous leaves, thereby increasing photosynthesis. The airspace
resistance (or its inverse, the airspace conductance,
\(g_\mathrm{ias}\)) is rarely measured directly and there is
disagreement between empirical (Parkhurst and Mott, 1990; Morison et
al., 2005; Evans et al., 2009; Tomás et al., 2013; Earles et al., 2018;
Šantrůček et al., 2019; Nobel, 2020; Harwood et al., 2021; Márquez et
al., 2023) and theoretical models (Tholen and Zhu, 2011; Ho et al.,
2016; Théroux-Rancourt et al., 2021). The \(g_\mathrm{ias}\) in thin,
porous leaves may be so large as to be inconsequential given much lower
conductances for other components of the diffusion pathway, whereas the
\(g_\mathrm{ias}\) of thick leaves with little airspace may greatly
hinder CO\(_2\) diffusion to chloroplasts. Amphistomy should confer the
largest photosynthetic benefit in leaves with intrinsically low
\(g_\mathrm{ias}\).

Amphistomy should also enhance photosynthesis when leaf boundary layer
resistance is high, because apportioning total flux between two boundary
layers rather than one results in a smaller CO\(_2\) concentration drop
between the atmosphere and stomata. A similar effect has been validated
with a computer model and measurements for transpiration: amphistomatous
leaves lose somewhat more water for the same vapor pressure deficit and
total \(g_\mathrm{sw}\) (Foster and Smith, 1986), but the additional
carbon gain should be enough to offset this cost under most realistic
conditions (Muir, 2019).

The most promising adaptive hypothesis is that amphistomy is important
for maximizing photosynthetic rate under high light. Mott et al. (1982)
proposed that ``plants with a high photosynthetic capacity, living in
full-sun environments, and experiencing rapidly fluctuating or
continuously available soil water'' would benefit most, in terms of
increased carbon gain, from having amphistomatous leaves. As described
above, herbs from sunny habitats are often amphistomatous. Most
variation in stomatal density ratio (\(\mathrm{SR}\), the ratio of
stomatal density between the upper and lower surfaces) among species is
assumed to be genetic, but there is also putatively adaptive plasticity
in response to light. Leaves of \emph{Ambrosia cordifolia}, a desert
perennial herb, are hypostomatous under low light (photosynthetic photon
flux density, PPFD = 110 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
but develop \textasciitilde20\% of their stomata on the upper surface
under high light (1700 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
(Mott and Michaelson, 1991). Similarly, \emph{Solanum lycopersicum}
leaves are hypostomatous when grown in the shade but develop
\textasciitilde20\% of their stomata on the upper surface grown under
high light-intensity (Gay and Hurd, 1975). Adult leaves of
\emph{Eucalyptus globulus} are amphistomatous, but the proportion of
adaxial stomata increases from \textasciitilde10-20\% under low light to
\textasciitilde30-40\% under high light (James and Bell, 2001). In
summary, both genetic and plastic responses to evince a widespread
association between light and \(\mathrm{SR}\).

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
leaves have a stomatal density ratio closer to 0.5 (same density on each
leaf surface), this will confer a greater advantage than an
amphistomatous leaf with most stomata on one surface. In other words,
amphistomy doesn't make much difference if one leaf surface has very
open stomata on it. We therefore predict that sun leaves will have
greater \(\mathrm{AA}\) possibly because they have thicker leaves and/or
\(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that takes into account differences in both stomatal density and size
between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to
test the relationship between light habitat and \(\mathrm{AA}\). Many
lineages have adapted to different light habitats after colonization and
leaf anatomical traits such as \(\mathrm{SR}\) and thickness vary within
and among closely related species. It is hypothesized that the common
ancestor in many Hawaiian clades was a ``weedy'' species with high
dispersal ability adapted to open habitats (Carlquist, 1966).
Colonization was followed by adaptive radiation into higher elevation,
montane, closed, forested habitats. Consequently, adaptation to sun and
shade is a common axis of phenotypic variation among Hawaiian plants
such as lobeliads (Givnish et al., 2004; Montgomery and Givnish, 2008;
Givnish et al., 2009; Givnish and Montgomery, 2014; Scoffoni et al.,
2015), \emph{Bidens} (Carlquist, 1966; Knope et al., 2020),
\emph{Scaevola} (Robichaux and Pearcy, 1984; McKown et al., 2016),
\emph{Euphorbia} (Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al.,
2009).

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that
help them to compete for light availability (Raunch et al.~1997). This
montane morphology is not found in \emph{S. fallax} populations on other
Pacific Islands (Pejhanmehr, 2022). The variable morphology between
coastal and montane populations has even led some scientists to argue
that it makes up a species complex, rather than a single species (Raunch
et al.~1997).

The objectives of our study are to test whether 1) `sun' leaves of
coastal ʻilima plants will have greater \(\mathrm{AA}\) than `shade'
leaves of montane plants; and if so, is this because 2a) coastal plants
have thicker leaves than montane plants and/or 2b) coastal plants have a
\(g_{\mathrm{smax,ratio}}\) closer to 0.5?

\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}

\hypertarget{methods}{%
\section{MATERIALS AND METHODS ---}\label{methods}}

\hypertarget{plant-sampling}{%
\subsection{Plant sampling ---}\label{plant-sampling}}

We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting (Yorkston and Daehler, 2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
in one three-day trip. We haphazardly sampled eight plants distributed
evenly between the highest and lowest elevation plants along a transect
at each site. For safety and conservation reasons, transects were along
a trail or road. We did not sample small individuals if there was risk
removing leaves would cause mortality. From each plant, we collected two
fully expanded leaves for traits measurements. We sampled stomatal
traits on all leaves; leaf thickness on one leaf from three randomly
selected plants per site; and, due to limited time, a single leaf from a
single plant at the middle of each transect for gas exchange
measurements.

\hypertarget{tbl-locations}{}
\begin{table}
\caption{\label{tbl-locations}ʻIlima study sites for this study. }\tabularnewline

\centering
\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
\toprule
Site & Island & Habitat & Latitude & Longitude & Elevation (mas)\\
\midrule
\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.710} & \cellcolor{gray!6}{-157.982} & \cellcolor{gray!6}{4}\\
Kaloko beach & Oʻahu & coastal & 21.293 & -157.661 & 4\\
\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{19.676} & \cellcolor{gray!6}{-156.024} & \cellcolor{gray!6}{6}\\
Kaʻena Point & Oʻahu & coastal & 21.574 & -158.278 & 4\\
\cellcolor{gray!6}{Makapuʻu beach} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.313} & \cellcolor{gray!6}{-157.661} & \cellcolor{gray!6}{3}\\
\addlinespace
Puakō petroglyph park & Hawaiʻi & coastal & 19.957 & -155.858 & 8\\
\cellcolor{gray!6}{Hawaiʻi loa ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.294} & \cellcolor{gray!6}{-157.727} & \cellcolor{gray!6}{352}\\
Hāloa ʻĀina & Hawaiʻi & montane & 19.552 & -155.793 & 1567\\
\cellcolor{gray!6}{Kaʻohe game management area} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{19.817} & \cellcolor{gray!6}{-155.616} & \cellcolor{gray!6}{1946}\\
Koaiʻa tree sanctuary & Hawaiʻi & montane & 20.048 & -155.737 & 970\\
\addlinespace
\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.305} & \cellcolor{gray!6}{-157.779} & \cellcolor{gray!6}{248}\\
Waʻahila ridge & Oʻahu & montane & 21.314 & -157.793 & 357\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{leaf-traits}{%
\subsection{Leaf traits ---}\label{leaf-traits}}

\hypertarget{stomata}{%
\subsubsection{Stomata ---}\label{stomata}}

We estimated the stomatal density and size on ab- and adaxial leaf
surfaces from all leaves. For pubescent leaves (usually coastal), we
dried and pressed leaves for \(\approx 1\) week (Hill et al., 2014),
carefully scraped trichomes off with a razor blade, and rehydrated the
leaf. For glabrous leaves, we used fresh leaves. We applied clear nail
polish to both leaf surfaces of fresh or rehydrated leaves in the middle
of the lamina away from major veins. After nail polish dried, we mounted
impressions on a microscope slide using transparent tape (Mott and
Michaelson, 1991). We digitized a portion of each leaf surface
impression using a brightfield microscope (Leica DM2000, Wetzlar,
Germany). We counted all stomata and divided by the visible leaf area
(0.890 mm\(^2\)) to estimate density and measured guard cell length from
five randomly chosen stomata per field using ImageJ (Schneider et al.,
2012).

\hypertarget{leaf-thickness}{%
\subsubsection{Leaf thickness ---}\label{leaf-thickness}}

We cut thin sections using two razor blades taped together and cutting
the leaf in a petri dish of water, wet-mounted sections onto a slide,
and took digital micrographs using a brightfield microscope, as
described above.

\hypertarget{gas-exchange-measurements}{%
\subsubsection{Gas exchange measurements
---}\label{gas-exchange-measurements}}

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
but with modification. For each leaf, we measured the photosynthetic
rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
range of \(g_\text{sw}\) values, holding other environmental variables
constant. We refer to this as an \(A \textendash g_\text{sw}\) curve. We
compared the \(A \textendash g_\text{sw}\) curve of the untreated leaf
to the photosynthetic rate of ``pseudohypostomatous'' leaf
(\(A_\mathrm{hypo}\)), which is the same leaf but with gas exchange
through the upper surface blocked by a neutral density plastic
(propafilm). We refer to the untreated and partially blocked leaves as
``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
calculated as the log-response ratio of \(A\) compared at the same total
\(g_\text{sw}\):

\[\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})\]

The log-response ratio is easy to interpret because values above 0
indicate a photosynthetic advantage of amphistomy, whereas values less
than 0 indicate a disadvantage. The response ratio is also better than
the absolute difference because it indicates a proportional change in
\(A\), which facilitates comparisons across leaves and environments with
different baseline photosynthetic rates. The irradiance of the light
source in the pseudohypo leaf is higher because the propafilm reduces
transmission. Because the propafilm reduces transmission, we used a
higher incident \(\mathrm{PPFD}\) for pseudohypo leaves (higher by a
factor 1/0.91, the inverse of the measured transmissivity of the
propafilm). We also set the stomatal conductance ratio, for purposes of
calculating boundary layer conductance, to 0 for pseudohypo leaves
following manufacturer directions.

To control for order effects, we alternated between starting with amphi
(untreated) or pseudohypo (treated) leaf measurements. In the field, we
acclimated the focal leaf to high light
(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
high relative humidity (\(\mathrm{RH} = 70\%\)),
\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
closure without photosynthetic downregulation. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process with the
other treatment (amphi or pseudohypo).

\hypertarget{data-analyis}{%
\subsection{Data analysis ---}\label{data-analyis}}

\hypertarget{objective-1-do-coastal-leaves-have-greater-amphistomy-advantage-than-montane-leaves}{%
\subsubsection{Objective 1: Do coastal leaves have greater amphistomy
advantage than montane leaves?
---}\label{objective-1-do-coastal-leaves-have-greater-amphistomy-advantage-than-montane-leaves}}

It is not feasible to record \(A_\mathrm{amphi}\) and
\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
we fit \(A \textendash g_\text{sw}\) curves using a flexible generalized
additive model (GAM) to interpolate modeled \(A\) for amphi and
pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12 leaves, the
minimum \(g_\text{sw}\) of the amphi curve was smaller than the maximum
\(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves overlapped for
a range of \(g_\text{sw}\) values). In those cases, we estimated
\(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle of the
range of overlap. In 2 of 12 leaves, the \(A \textendash g_\text{sw}\)
curves did not quite overlap because the minimum \(g_\text{sw}\) of the
amphi curve was slightly greater than the maximum \(g_\text{sw}\) of the
pseudohypo curve. In those cases, we estimated \(\mathrm{AA}\) by
extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. We fit GAMs using the \emph{R} package \textbf{brms} version
2.20.1 (Bürkner, 2017) for Bayesian statistical inference with MCMC
sampling in \emph{Stan} (Stan Development Team, 2023). We used CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023) to interface with \emph{R} version 4.3.1 (R Core Team, 2023). We
sampled from posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated parameters and
confidence intervals as the median and 95\% quantile intervals of the
posterior, respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).

\hypertarget{objective-2a-are-coastal-leaves-thicker-than-montane-leaves}{%
\subsubsection{Objective 2a: Are coastal leaves thicker than montane
leaves?
---}\label{objective-2a-are-coastal-leaves-thicker-than-montane-leaves}}

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with site-type as a fixed effect
and individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.32.2
and \textbf{cmdstanr} version 0.5.3 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated the relationship between
population average leaf thickness and \(\mathrm{AA}\) measured from a
single individual per population. We used this approach most of the
variation in leaf thickness was among sites (see Results) and the plant
selected for gas exchange measurements was not always among the plants
randomly selected for leaf thickness, precluding individual level
correlation. We propagated uncertainty about in \(\mathrm{AA}\) and leaf
thickness by integrating over the entire posterior distribution sample
for each variable. The key prediction is that the effect of leaf
thickness on \(\mathrm{AA}\) is positive.

\hypertarget{objective-2b-is-g_mathrmsmaxratio-closer-to-0.5-in-coastal-leaves-than-montane-leaves}{%
\subsubsection{\texorpdfstring{Objective 2b: Is
\(g_{\mathrm{smax,ratio}}\) closer to 0.5 in coastal leaves than montane
leaves?
---}{Objective 2b: Is g\_\{\textbackslash mathrm\{smax,ratio\}\} closer to 0.5 in coastal leaves than montane leaves? ---}}\label{objective-2b-is-g_mathrmsmaxratio-closer-to-0.5-in-coastal-leaves-than-montane-leaves}}

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were site-type as a fixed effect and leaf
within individual plant, individual plant, and site as random effects.
We used the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner,
2017) to fit the model in \emph{Stan} (Stan Development Team, 2023) with
CmdStan version 2.32.2 and \textbf{cmdstanr} version 0.5.3 (Gabry and
Češnovar, 2023). We interpolated missing adaxial guard cell lengths from
6 out of 185 samples with zero adaxial stomata using the ``mi'' function
in \textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations. From each
posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:

\[g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},\]
where \(g_\mathrm{smax,lower}\) and \(g_\mathrm{smax,upper}\) are
maximum stomatal conductance to water vapor at
\(T_\mathrm{leaf} = 25 ^ \circ\) C on the lower and upper surface,
respectively. The maximum stomatal conductance is calculated from
stomatal density and length, assuming that stomata are fully open,
following Sack and Buckley (2016).

We estimated the relationship between leaf \(g_\mathrm{smax,ratio}\) and
\(\mathrm{AA}\) measured from a single leaf per population. We
propagated uncertainty about in \(\mathrm{AA}\) and
\(g_\mathrm{smax,ratio}\) by integrating over the entire posterior
distribution sample for each variable. The key prediction is that the
effect of \(g_\mathrm{smax,ratio}\) on \(\mathrm{AA}\) is positive until
\(g_\mathrm{smax,ratio} < 0.5\).

\hypertarget{results}{%
\section{RESULTS ---}\label{results}}

\hypertarget{amphistomy-advantage-is-greater-in-coastal-leaves}{%
\subsection{Amphistomy advantage is greater in coastal leaves
---}\label{amphistomy-advantage-is-greater-in-coastal-leaves}}

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.078}{0.15}]\) and
\(0.027~[\numrange{-0.0033}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.04}{0.14}]\).
The difference between coastal and montane leaves is lower,
\(0.054~[\numrange{-0.012}{0.11}]\), but still greater than 0 if we
remove two sites where we extrapolated slightly beyond fitted
\(A \textendash g_\mathrm{sw}\) curves. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}).

\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}

\hypertarget{leaf-thickness-is-associated-with-amphistomy-advantage-between-but-not-within-habitats}{%
\subsection{Leaf thickness is associated with amphistomy advantage
between but not within habitats
---}\label{leaf-thickness-is-associated-with-amphistomy-advantage-between-but-not-within-habitats}}

Coastal ʻilima leaves are \(91~[\numrange{26}{164}]\) \(\mu\text{m}\)
thicker than their montane counterparts. Although coastal leaves are
thicker and have greater \(\mathrm{AA}\), the relationship between leaf
thickness and \(\mathrm{AA}\) within habitats is actually slightly
negative (\autoref{fig:traits-aa}A;
\(\text{slope}~= -0.11~[\numrange{-0.28}{0.033}]\)), the opposite
direction our hypothesis predicts.

\hypertarget{g_mathrmsmaxratio-is-not-associated-with-amphistomy-advantage}{%
\subsection{\texorpdfstring{\(g_\mathrm{smax,ratio}\) is not associated
with amphistomy advantage
---}{g\_\textbackslash mathrm\{smax,ratio\} is not associated with amphistomy advantage ---}}\label{g_mathrmsmaxratio-is-not-associated-with-amphistomy-advantage}}

Coastal and montane leaves have similar average
\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
\(0.057~[\numrange{-0.16}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
montane leaves, but the 95\% confidence intervals overlap 0 difference.
The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
and others with \(g_\mathrm{smax,ratio} > 0.2\)
(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
montane sites where those on the Big Island of Hawaiʻi all have low
\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
\(g_\mathrm{smax,ratio}\). There is no relationship between
\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
(\autoref{fig:traits-aa}B;
\(\text{slope}~= 0.11~[\numrange{-0.032}{0.25}]\)) in our sample.

\begin{figure}[H]
  \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) $g_\mathrm{smax,ratio}$ and (B) leaf thickness among ʻilima (*Sida fallax*) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi., and amphistomy advantage. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}

\hypertarget{discussion}{%
\section{DISCUSSION ---}\label{discussion}}

Amphistomy is a seemingly simple way that leaves can increase carbon
gain without significant additional water loss, yet it is rare in nature
and we do not know why. The strong association between amphiostomy and
sunny, open habitats suggests that amphistomy may benefit sun leaves
more than shade leaves, but progress has been limited by the lack of
evidence that amphiostomy \emph{per se} affects improves photosynthesis
in a given leaf. By experimentally blocking gas exchange through the
upper surface in a controlled environment, we directly compared an
amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
This allows us to quantity the `amphistomy advantage' (\(\mathrm{AA}\))
holding all else constant. Taking advantage of the of the steep climatic
gradients in the Hawaiian archipegalo, we applied this new method to
show for the first time that sun leaves benefit more from amphistomy
than shade leaves in ʻilima (\emph{Sida fallax}). While this result has
not yet been validated in other speciesthere are limits to how much we
can generalize to all land plants, our results indicate that part of the
reason amphistomatous leaves are found most commonly in high light
habitats is that the adaptive benefit is greater in this environment.

If \(\mathrm{AA}\) is typically greater in sun leaves than shade leaves,
it could partially explain the distribution of amphi- and hypostoamtous
leaves, but the precise mechanism(s) require further study. One
hypothesis is that the internal airspace conductance,
\(g_\mathrm{ias}\), from stomata to mesophyll cell walls is lower in
thicker sun leaves (Parkhurst, 1978). All else being equal, a leaf with
lower \(g_\mathrm{ias}\) will benefit more from amphistomy. Our results
partially support this hypothesis. Coastal ʻilima leaves with high
\(\mathrm{AA}\) (\autoref{fig:habitat-aa}) are thicker than montane
leaves, but the relationship between \(\mathrm{AA}\) and leaf thickness
within habitats is actually slightly negative
(\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and
montane leaves differ in many respects besides thickness, we do not have
enough data to conclude that leaf thickness explains the variation in
\(\mathrm{AA}\) between habitats. Alternatively, other biochemical or
anatomical differences between coastal and montane leaves may explain
why \(\mathrm{AA}\) is greater in coastal leaves. The negative
relationship between leaf thickness and \(\mathrm{AA}\) could be
explained if thicker leaves compensated by having a more porous
mesophyll and/or less tortuous airspaces (Théroux-Rancourt et al.,
2021).

A second natural hypothesis is that amphistomatous leaves with few
adaxial (upper) stomata benefit less than those with similar densities
on both surfaces. We predicted that leaves with
\(g_\mathrm{smax,ratio}\) closer to 0.5 would have higher
\(\mathrm{AA}\) based on biophysical models (Gutschick, 1984a). The
logic is that a small number of stomata on the upper surface are
insufficient to supply the entire upper mesophyll due to limited lateral
diffusion (Morison et al., 2005). Our results do not support this
hypothesis. Montane leaves from Big Island sites had low
\(g_\mathrm{smax,ratio}\) and low \(\mathrm{AA}\) whereas low montane
leaves on Oʻahu had high \(g_\mathrm{smax,ratio}\), but similarly low
\(\mathrm{AA}\) (\autoref{fig:traits-AA}B). Among coastal sites, the
site with the lowest \(g_\mathrm{smax,ratio}\) had the lowest
\(\mathrm{AA}\), but there was little variation in
\(g_\mathrm{smax,ratio}\) among coastal leaves in our sample. We
therefore cannot rule out that a larger sample of coastal leaves with
greater variance in \(g_\mathrm{smax,ratio}\) might support this
hypothesis.

Two major implications from our study are that 1) photosynthesis in
hypostomatous leaves is likely limited by CO\(_2\) concentration
drawdown within leaf airspaces; and 2) amphistomy \emph{per se}
contributes to, but is not wholly responsible for, higher photosynthetic
rates among amphistomatous leaves. The amphistomy advantage we observe
in ʻilima leaves implies decreased CO\(_2\) supply in
pseudohypostomatous leaves because of concentration drawdowns in the
leaf airspace. Limited diffusion through the airspace has long been
hypothesized to depress photosynthesis in hypostomatous leaves
(Parkhurst, 1994), with empirical support from helox studies (Parkhurst
and Mott, 1990). However, these studies relied on interspecific
comparisons of amphi- and hypostomatous leaves that differ
systematically in many traits that affect gas exchange and
photosynthesis (Xiong and Flexas, 2020). Our experimental approach
overcomes this limitation and implies that the drop in CO\(_2\)
concentration from substomatal cavities to the upper surface depresses
photosynthesis.

Among land plants grown in a common garden, amphistomatous leaves have
on average nearly \(2\times\) higher area-based photosynthetic rates
(Xiong and Flexas, 2020), naively implying an
\(\mathrm{AA} \approx \text{log}~2 = 0.69\). This is much higher than
our estimate of 0.12 among coastal ʻilima leaves. The most likely
explanation is that amphistomy is not the only cause of high
photosynthetic rate. Indeed, species adapted to open, high light
habitats with amphistomatous leaves also have higher concentrations of
Rubisco, overall stomatal conductance, and photosynthetic capacity
(Smith et al., 1997; Xiong and Flexas, 2020). For a leaf with high
photosynthetic capacity that is well illuminated and hydrated, the major
limitation becomes CO\(_2\). Under these conditions, amphistomy may
substantially increase photosynthesis, as we observe in coastal ʻilima
leaves. Selection on increased photosynthesis under similar conditions
may explain why crops leaves tend to increase stomatal density ratio
during domestication (Milla et al., 2013).

Two limitations of this study are the small sample sample sizes and
experimental design that precludes distinguishing genetic from
environmental differences in leaf traits. Understanding the mechanistic
basis of higher \(\mathrm{A}A\) in sun leaves would require much larger
sample sizes. Sun leaves tend to be thicker, more densely packed with
mesophyll cells, and have greater photosynthetic capacity and higher
stomatal conductance, among other traits (Lambers et al., 2008). Each of
these factors and others potentially modulate \(\mathrm{AA}\).
Quantifying the contribution of all these factors requires larger
samples and additional measurements that are beyond the scope of this
study, but exciting avenues for future research on leaf
structure-function relations. Although many morphological traits that
distinguish coastal and montane ʻilima populations persist in a common
environment (Yorkston and Daehler, 2006), we cannot distinguish between
genetic effects and plastic responses to habitat as causes of difference
in \(\mathrm{AA}\) because we measured naturally occurring plants
\emph{in situ}. While disentangling genetic and plastic contributions is
not necessarily important for understanding the distribution of
amphistomatous leaves, it would be useful understand genetic and
environmental contributions to trait variation. A reciprocal transplant
would be able to determine the genetic and environmental contributions,
as well their interaction, to trait variance in nature.

\hypertarget{conclusions}{%
\section{CONCLUSIONS ---}\label{conclusions}}

This study reports the first direct experimental evidence that having
stomata open on both leaf surfaces, amphistomy, increases photosynthesis
for a given total stomatal conductance, particularly in leaves from the
type of open, sunny habitats where this trait is most common. By
developing a straightforward gas exchange method to block gas exchange
through the upper surface, we directly compared the photosynthetic rate
of a leaf with gas exchange through both surfaces or just one, holding
all other factors constant. In doing so, we found that coastal leaves of
the indigenous Hawaiian ʻilima (\emph{Sida fallax}) enjoyed a greater
photosynthetic benefit from amphistomy than nearby montane leaves living
in more closed forest. We cannot yet ascribe the difference in
amphistomy advantage between coastal and montane leaves to particular
physiological or anatomical variation, but this is a promising area for
future research.

\hypertarget{acknowledgments}{%
\section{Acknowledgments ---}\label{acknowledgments}}

The authors thank Kasey Barton for feedback on an earlier version of
this manuscript, the University of Hawaiʻi honors council for guidance
to GT, Tawn Speetjens for access to one montane site, startup funds from
the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
2307341 to TNB.

\hypertarget{author-ontributions}{%
\section{Author Contributions ---}\label{author-ontributions}}

GT and CDM contributed equally to all stages of this project; TNB
contributed to development of the method and helped edit the manuscript.

\hypertarget{data-availability-statement}{%
\section{Data Availability Statement
---}\label{data-availability-statement}}

Custom scripts are available on a GitHub repository
(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
with a DOI and stable URL upon publication. Raw data will be deposited
on Dryad with a DOI and stable URL upon publication.

\hypertarget{supporting-information}{%
\section{Supporting Information ---}\label{supporting-information}}

Additional supporting information may be found online in the Supporting
Information section at the end of the article

\begin{itemize}
\tightlist
\item
  Appendix S1: Supplemental figures and table
\end{itemize}

\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\renewcommand\theequation{S\arabic{equation}}    
\setcounter{figure}{0}    
\setcounter{table}{0}    
\setcounter{equation}{0}

\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:licor}
\end{figure}

\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:habitat-Ags}
\end{figure}

\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:habitat-gmaxratio}
\end{figure}

\begin{table}
  \captionsetup{labelformat=empty}
  \caption{}

\end{table}

\hypertarget{literature-cited}{%
\section*{Literature Cited ---}\label{literature-cited}}
\addcontentsline{toc}{section}{Literature Cited ---}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-noauthor_inaturalist_2022}{}}%
Anon. 2022. \href{https://www.inaturalist.org}{{iNaturalist}}.

\leavevmode\vadjust pre{\hypertarget{ref-biggins_model_1987}{}}%
Ball, J. T., I. E. Woodrow, and J. A. Berry. 1987.
\href{https://doi.org/10.1007/978-94-017-0519-6_48}{A {Model}
{Predicting} {Stomatal} {Conductance} and its {Contribution} to the
{Control} of {Photosynthesis} under {Different} {Environmental}
{Conditions}}. \emph{In} J. Biggins {[}ed.{]}, Progress in
{Photosynthesis} {Research}, 221--224. Springer Netherlands, Dordrecht.

\leavevmode\vadjust pre{\hypertarget{ref-berry_stomata:_2010}{}}%
Berry, J. A., D. J. Beerling, and P. J. Franks. 2010.
\href{https://doi.org/10.1016/j.pbi.2010.04.013}{Stomata: Key players in
the earth system, past and present}. \emph{Current Opinion in Plant
Biology} 13: 232--239.

\leavevmode\vadjust pre{\hypertarget{ref-de_boer_apparent_2016}{}}%
Boer, H. J. de, P. L. Drake, E. Wendt, C. A. Price, E.-D. Schulze, N. C.
Turner, D. Nicolle, and E. J. Veneklaas. 2016.
\href{https://doi.org/10.1104/pp.16.01313}{Apparent overinvestment in
leaf venation relaxes leaf morphological constraints on photosynthesis
in arid habitats}. \emph{Plant Physiology} 172: 2286--2299.

\leavevmode\vadjust pre{\hypertarget{ref-bucher_stomatal_2017}{}}%
Bucher, S. F., K. Auerswald, C. Grün-Wenzel, S. I. Higgins, J. Garcia
Jorge, and C. Römermann. 2017.
\href{https://doi.org/10.1016/j.flora.2017.02.011}{Stomatal traits
relate to habitat preferences of herbaceous species in a temperate
climate}. \emph{Flora} 229: 107--115.

\leavevmode\vadjust pre{\hypertarget{ref-burkner_brms_2017}{}}%
Bürkner, P.-C. 2017.
\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
\emph{stan}}. \emph{Journal of Statistical Software} 80.

\leavevmode\vadjust pre{\hypertarget{ref-carlquist_biota_1966}{}}%
Carlquist, S. 1966. \href{https://doi.org/10.2307/2406147}{The {Biota}
of {Long}-{Distance} {Dispersal}. {II}. {Loss} of {Dispersibility} in
{Pacific} {Compositae}}. \emph{Evolution} 20: 30.

\leavevmode\vadjust pre{\hypertarget{ref-collatz_physiological_1991}{}}%
Collatz, G. J., J. T. Ball, C. Grivet, and J. A. Berry. 1991.
\href{https://doi.org/10.1016/0168-1923(91)90002-8}{Physiological and
environmental regulation of stomatal conductance, photosynthesis and
transpiration: A model that includes a laminar boundary layer}.
\emph{Agricultural and Forest Meteorology} 54: 107--136.

\leavevmode\vadjust pre{\hypertarget{ref-cowan_stomatal_1977}{}}%
Cowan, I. R., and G. D. Farquhar. 1977. Stomatal function in relation to
leaf metabolism and environment. \emph{Symposia of the Society for
Experimental Biology} 31: 471--505.

\leavevmode\vadjust pre{\hypertarget{ref-deans_optimization_2020}{}}%
Deans, R. M., T. J. Brodribb, F. A. Busch, and G. D. Farquhar. 2020.
\href{https://doi.org/10.1038/s41477-020-00760-6}{Optimization can
provide the fundamental link between leaf photosynthesis, gas exchange
and water relations}. \emph{Nature Plants} 6: 1116--1125.

\leavevmode\vadjust pre{\hypertarget{ref-doll_callitriche_2021}{}}%
Doll, Y., H. Koga, and H. Tsukaya. 2021.
\href{https://doi.org/10.1080/15592324.2021.1978201}{\emph{Callitriche}
as a potential model system for evolutionary studies on the dorsiventral
distribution of stomata}. \emph{Plant Signaling \& Behavior} 16:
1978201.

\leavevmode\vadjust pre{\hypertarget{ref-drake_two_2019}{}}%
Drake, P. L., H. J. de Boer, S. J. Schymanski, and E. J. Veneklaas.
2019. \href{https://doi.org/10.1111/nph.15652}{Two sides to every leaf:
Water and {CO}\(_{\textrm{2}}\) transport in hypostomatous and
amphistomatous leaves}. \emph{New Phytologist} 222: 1179--1187.

\leavevmode\vadjust pre{\hypertarget{ref-dunbar-co_leaf_2009}{}}%
Dunbar-Co, S., M. J. Sporck, and L. Sack. 2009. Leaf trait
diversification and design in seven rare taxa of the {Hawaiian}
\emph{p}lantago radiation. \emph{International Journal of Plant
Sciences} 170: 61--75.

\leavevmode\vadjust pre{\hypertarget{ref-earles_beyond_2018}{}}%
Earles, J. M., G. Theroux-Rancourt, A. B. Roddy, M. E. Gilbert, A. J.
McElrone, and C. R. Brodersen. 2018.
\href{https://doi.org/10.1104/pp.18.00550}{Beyond porosity: {3D} leaf
intercellular airspace traits that impact mesophyll conductance}.
\emph{Plant Physiology} 178: 148--162.

\leavevmode\vadjust pre{\hypertarget{ref-ehleringer_pubescence_1978}{}}%
Ehleringer, J. R., and O. Björkman. 1978.
\href{https://doi.org/10.1007/BF00349805}{Pubescence and leaf spectral
characteristics in a desert shrub, {Encelia} farinosa}. \emph{Oecologia}
36: 151--162.

\leavevmode\vadjust pre{\hypertarget{ref-evans_resistances_2009}{}}%
Evans, J. R., R. Kaldenhoff, B. Genty, and I. Terashima. 2009.
\href{https://doi.org/10.1093/jxb/erp117}{Resistances along the
{CO}\(_{\textrm{2}}\) diffusion pathway inside leaves}. \emph{Journal of
Experimental Botany} 60: 2235--2248.

\leavevmode\vadjust pre{\hypertarget{ref-foster_influence_1986}{}}%
Foster, J. R., and W. K. Smith. 1986. Influence of stomatal distribution
on transpiration in low-wind environments. \emph{Plant, Cell \&
Environment} 9: 751--759.

\leavevmode\vadjust pre{\hypertarget{ref-franklin_organizing_2020}{}}%
Franklin, O., S. P. Harrison, R. Dewar, C. E. Farrior, Å. Brännström, U.
Dieckmann, S. Pietsch, et al. 2020.
\href{https://doi.org/10.1038/s41477-020-0655-x}{Organizing principles
for vegetation dynamics}. \emph{Nature Plants} 6: 444--453.

\leavevmode\vadjust pre{\hypertarget{ref-franks_comparing_2018}{}}%
Franks, P. J., G. B. Bonan, J. A. Berry, D. L. Lombardozzi, N. M.
Holbrook, N. Herold, and K. W. Oleson. 2018.
\href{https://doi.org/10.1111/gcb.14445}{Comparing optimal and empirical
stomatal conductance models for application in {Earth} system models}.
\emph{Global Change Biology} 24: 5708--5723.

\leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
Gabry, J., and R. Češnovar. 2023.
\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
{R} {Interface} to '{CmdStan}'}.

\leavevmode\vadjust pre{\hypertarget{ref-gay_influence_1975}{}}%
Gay, A. P., and R. G. Hurd. 1975.
\href{https://doi.org/10.1111/j.1469-8137.1975.tb01368.x}{The influence
of light on stomatal density in the tomato}. \emph{New Phytologist} 75:
37--46.

\leavevmode\vadjust pre{\hypertarget{ref-gibson_structure-function_1996}{}}%
Gibson, A. C. 1996.
\href{http://public.eblib.com/choice/PublicFullRecord.aspx?p=6495247}{Structure-{Function}
{Relations} of {Warm} {Desert} {Plants}}. Springer Berlin / Heidelberg,
Berlin, Heidelberg.

\leavevmode\vadjust pre{\hypertarget{ref-givnish_economy_1986}{}}%
Givnish, T. J. ed. 1986. On the economy of plant form and function.
Cambridge University Press, Cambridge.

\leavevmode\vadjust pre{\hypertarget{ref-givnish_origin_2009}{}}%
Givnish, T. J., K. C. Millam, A. R. Mast, T. B. Paterson, T. J. Theim,
A. L. Hipp, J. M. Henss, et al. 2009.
\href{https://doi.org/10.1098/rspb.2008.1204}{Origin, adaptive radiation
and diversification of the {Hawaiian} lobeliads ({Asterales}:
{Campanulaceae})}. \emph{Proceedings of the Royal Society B: Biological
Sciences} 276: 407--416.

\leavevmode\vadjust pre{\hypertarget{ref-givnish_common-garden_2014}{}}%
Givnish, T. J., and R. A. Montgomery. 2014.
\href{https://doi.org/10.1098/rspb.2013.2944}{Common-garden studies on
adaptive radiation of photosynthetic physiology among {Hawaiian}
lobeliads}. \emph{Proceedings of the Royal Society B: Biological
Sciences} 281: 20132944--20132944.

\leavevmode\vadjust pre{\hypertarget{ref-givnish_adaptive_2004}{}}%
Givnish, T. J., R. A. Montgomery, and G. Goldstein. 2004.
\href{https://doi.org/10.3732/ajb.91.2.228}{Adaptive radiation of
photosynthetic physiology in the {Hawaiian} lobeliads: Light regimes,
static light responses, and whole-plant compensation points}.
\emph{American Journal of Botany} 91: 228--246.

\leavevmode\vadjust pre{\hypertarget{ref-grubb_leaf_2020}{}}%
Grubb, P. J. 2020. Leaf structure and function. \emph{In} A. Dobson, D.
Tilman, and R. D. Holt {[}eds.{]}, Unsolved {Problems} in {Ecology},
124--144. Princeton University Press, Princeton.

\leavevmode\vadjust pre{\hypertarget{ref-grubb_leaf_1977}{}}%
Grubb, P. J. 1977. Leaf structure and function. \emph{In} R. Duncan, and
M. Weston-Smith {[}eds.{]}, The encyclopedia of ignorance, 317--330.
Pergamon, Oxford.

\leavevmode\vadjust pre{\hypertarget{ref-gutschick_photosynthesis_1984}{}}%
Gutschick, V. P. 1984a. Photosynthesis model for {C}\(_{\textrm{3}}\)
leaves incorporating {CO}\(_{\textrm{2}}\) transport, propagation of
radiation, and biochemistry 1. Kinetics and their parameterization.
\emph{Photosynthetica} 18: 549--568.

\leavevmode\vadjust pre{\hypertarget{ref-gutschick_photosynthesis_1984-1}{}}%
Gutschick, V. P. 1984b. Photosynthesis model for {C}\(_{\textrm{3}}\)
leaves incorporating {CO}\(_{\textrm{2}}\) transport, propagation of
radiation, and biochemistry 2. Ecological and agricultural utility.
\emph{Photosynthetica} 18: 569--595.

\leavevmode\vadjust pre{\hypertarget{ref-harrison_ecoevolutionary_2021}{}}%
Harrison, S. P., W. Cramer, O. Franklin, I. C. Prentice, H. Wang, Å.
Brännström, H. Boer, et al. 2021.
\href{https://doi.org/10.1111/nph.17558}{Eco‐evolutionary optimality as
a means to improve vegetation and land‐surface models}. \emph{New
Phytologist} 231: 2125--2141.

\leavevmode\vadjust pre{\hypertarget{ref-harwood_understanding_2021}{}}%
Harwood, R., G. Théroux‐Rancourt, and M. M. Barbour. 2021.
\href{https://doi.org/10.1111/pce.14079}{Understanding airspace in
leaves: {3D} anatomy and directional tortuosity}. \emph{Plant, Cell \&
Environment} 44: 2455--2465.

\leavevmode\vadjust pre{\hypertarget{ref-hetherington_role_2003}{}}%
Hetherington, A. M., and F. I. Woodward. 2003.
\href{https://doi.org/10.1038/nature01843}{The role of stomata in
sensing and driving environmental change}. \emph{Nature} 424: 901--908.

\leavevmode\vadjust pre{\hypertarget{ref-hill_temperature_2014}{}}%
Hill, K. E., G. R. Guerin, R. S. Hill, and J. R. Watling. 2014.
\href{https://doi.org/10.1071/BT14204}{Temperature influences stomatal
density and maximum potential water loss through stomata of {Dodonaea}
viscosa subsp. Angustissima along a latitude gradient in southern
{Australia}}. \emph{Australian Journal of Botany} 62: 657.

\leavevmode\vadjust pre{\hypertarget{ref-ho_three-dimensional_2016}{}}%
Ho, Q. T., H. N. C. Berghuijs, R. Watté, P. Verboven, E. Herremans, X.
Yin, M. A. Retta, et al. 2016.
\href{https://doi.org/10.1111/pce.12590}{Three-dimensional microscale
modelling of {CO}\(_{\textrm{2}}\) transport and light propagation in
tomato leaves enlightens photosynthesis: 3-{D} modelling of
photosynthesis in leaves}. \emph{Plant, Cell \& Environment} 39: 50--61.

\leavevmode\vadjust pre{\hypertarget{ref-james_leaf_2001}{}}%
James, S. A., and D. T. Bell. 2001.
\href{https://doi.org/10.1071/BT99044}{Leaf morphological and anatomical
characteristics of heteroblastic {Eucalyptus} globulus ssp. Globulus
({Myrtaceae})}. \emph{Australian Journal of Botany} 49: 259.

\leavevmode\vadjust pre{\hypertarget{ref-jarvis_interpretation_1976}{}}%
Jarvis, P. G. 1976. \href{https://doi.org/10.1098/rstb.1976.0035}{The
interpretation of the variations in leaf water potential and stomatal
conductance found in canopies in the field}. \emph{Philosophical
Transactions of the Royal Society of London. B, Biological Sciences}
273: 593--610.

\leavevmode\vadjust pre{\hypertarget{ref-jordan_using_2014}{}}%
Jordan, G. J., R. J. Carpenter, and T. J. Brodribb. 2014.
\href{https://doi.org/10.1016/j.palaeo.2013.12.035}{Using fossil leaves
as evidence for open vegetation}. \emph{Palaeogeography,
Palaeoclimatology, Palaeoecology} 395: 168--175.

\leavevmode\vadjust pre{\hypertarget{ref-kaul_anatomical_1976}{}}%
Kaul, R. B. 1976.
\href{https://doi.org/10.1016/0304-3770(76)90022-X}{Anatomical
observations on floating leaves}. \emph{Aquatic Botany} 2: 215--234.

\leavevmode\vadjust pre{\hypertarget{ref-knope_dispersal_2020}{}}%
Knope, M. L., V. A. Funk, M. A. Johnson, W. L. Wagner, E. M. Datlof, G.
Johnson, D. J. Crawford, et al. 2020.
\href{https://doi.org/10.1111/jse.12704}{Dispersal and adaptive
radiation of \emph{bidens} ({Compositae}) across the remote
archipelagoes of {Polynesia}}. \emph{Journal of Systematics and
Evolution} 58: 805--822.

\leavevmode\vadjust pre{\hypertarget{ref-lambers_plant_2008}{}}%
Lambers, H., F. S. Chapin, and T. L. Pons. 2008.
\href{https://doi.org/10.1007/978-0-387-78341-3}{Plant {Physiological}
{Ecology}}. Springer New York, New York, NY.

\leavevmode\vadjust pre{\hypertarget{ref-leuning_critical_1995}{}}%
Leuning, R. 1995.
\href{https://doi.org/10.1111/j.1365-3040.1995.tb00370.x}{A critical
appraisal of a combined stomatal-photosynthesis model for {C3} plants}.
\emph{Plant, Cell and Environment} 18: 339--355.

\leavevmode\vadjust pre{\hypertarget{ref-lin_optimal_2015}{}}%
Lin, Y.-S., B. E. Medlyn, R. A. Duursma, I. C. Prentice, H. Wang, S.
Baig, D. Eamus, et al. 2015.
\href{https://doi.org/10.1038/nclimate2550}{Optimal stomatal behaviour
around the world}. \emph{Nature Climate Change} 5: 459--464.

\leavevmode\vadjust pre{\hypertarget{ref-lyshede_comparative_2002}{}}%
Lyshede, O. B. 2002. Comparative and functional leaf anatomy of selected
{Alstroemeriaceae} of mainly {Chilean} origin. \emph{Botanical Journal
of the Linnean Society} 140: 261--272.

\leavevmode\vadjust pre{\hypertarget{ref-marquez_assessing_2023}{}}%
Márquez, D. A., H. Stuart‐Williams, L. A. Cernusak, and G. D. Farquhar.
2023. \href{https://doi.org/10.1111/nph.18784}{Assessing the
{\textless{}}span style="font-variant:small-caps;"{\textgreater{}} {CO}
\(_{\textrm{2}}\) {\textless{}}/span{\textgreater{}} concentration at
the surface of photosynthetic mesophyll cells}. \emph{New Phytologist}
238: 1446--1460.

\leavevmode\vadjust pre{\hypertarget{ref-mckown_trait_2016}{}}%
McKown, A. D., M. E. Akamine, and L. Sack. 2016.
\href{https://doi.org/10.1007/s00442-016-3640-3}{Trait convergence and
diversification arising from a complex evolutionary history in
{Hawaiian} species of {Scaevola}}. \emph{Oecologia} 181: 1083--1100.

\leavevmode\vadjust pre{\hypertarget{ref-medlyn_reconciling_2011}{}}%
Medlyn, B. E., R. A. Duursma, D. Eamus, D. S. Ellsworth, I. C. Prentice,
C. V. M. Barton, K. Y. Crous, et al. 2011.
\href{https://doi.org/10.1111/j.1365-2486.2010.02375.x}{Reconciling the
optimal and empirical approaches to modelling stomatal conductance}.
\emph{Global Change Biology} 17: 2134--2144.

\leavevmode\vadjust pre{\hypertarget{ref-milla_shifts_2013}{}}%
Milla, R., N. de Diego-Vico, and N. Martín-Robles. 2013.
\href{https://doi.org/10.1093/jxb/ert147}{Shifts in stomatal traits
following the domestication of plant species}. \emph{Journal of
Experimental Botany} 64: 3137--3146.

\leavevmode\vadjust pre{\hypertarget{ref-montgomery_adaptive_2008}{}}%
Montgomery, R. A., and T. J. Givnish. 2008.
\href{https://doi.org/10.1007/s00442-007-0936-3}{Adaptive radiation of
photosynthetic physiology in the {Hawaiian} lobeliads: Dynamic
photosynthetic responses}. \emph{Oecologia} 155: 455--467.

\leavevmode\vadjust pre{\hypertarget{ref-morison_lateral_2005}{}}%
Morison, J. I. L., E. Gallouët, T. Lawson, G. Cornic, R. Herbin, and N.
R. Baker. 2005. \href{https://doi.org/10.1104/pp.105.062950}{Lateral
diffusion of {CO}\(_{\textrm{2}}\) in leaves is not sufficient to
support photosynthesis}. \emph{Plant Physiology} 139: 254--266.

\leavevmode\vadjust pre{\hypertarget{ref-mott_adaptive_1982}{}}%
Mott, K. A., A. C. Gibson, and J. W. O'Leary. 1982.
\href{https://doi.org/10.1111/1365-3040.ep11611750}{The adaptive
significance of amphistomatic leaves}. \emph{Plant, Cell \& Environment}
5: 455--460.

\leavevmode\vadjust pre{\hypertarget{ref-mott_amphistomy_1991}{}}%
Mott, K. A., and O. Michaelson. 1991. Amphistomy as an adaptation to
high light intensity in \emph{{Ambrosia} cordifolia} ({Compositae}).
\emph{American Journal of Botany} 78: 76--79.

\leavevmode\vadjust pre{\hypertarget{ref-muir_is_2019}{}}%
Muir, C. D. 2019. \href{https://doi.org/10.1093/icb/icz085}{Is
{Amphistomy} an {Adaptation} to {High} {Light}? {Optimality} {Models} of
{Stomatal} {Traits} along {Light} {Gradients}}. \emph{Integrative and
Comparative Biology} 59: 571--584.

\leavevmode\vadjust pre{\hypertarget{ref-muir_light_2018}{}}%
Muir, C. D. 2018. Light and growth form interact to shape stomatal ratio
among {British} angiosperms. \emph{New Phytologist} 218: 242--252.

\leavevmode\vadjust pre{\hypertarget{ref-muir_making_2015}{}}%
Muir, C. D. 2015. \href{https://doi.org/10.1098/rspb.2015.1498}{Making
pore choices: Repeated regime shifts in stomatal ratio}.
\emph{Proceedings of the Royal Society B: Biological Sciences} 282:
20151498.

\leavevmode\vadjust pre{\hypertarget{ref-nobel_physicochemical_2020}{}}%
Nobel, P. S. 2020. Physicochemical and environmental plant physiology.
Fifth edition. Academic Press, an imprint of Elsevier, London, United
Kingdom ; San Diego, CA.

\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_diffusion_1994}{}}%
Parkhurst, D. F. 1994.
\href{http://www.jstor.org/stable/2557929}{Diffusion of
{CO}\(_{\textrm{2}}\) and other gases inside leaves}. \emph{New
Phytologist} 126: 449--479.

\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_adaptive_1978}{}}%
Parkhurst, D. F. 1978. \href{https://doi.org/10.2307/2259142}{The
adaptive significance of stomatal occurrence on one or both surfaces of
leaves}. \emph{The Journal of Ecology} 66: 367.

\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_intercellular_1990}{}}%
Parkhurst, D. F., and K. A. Mott. 1990.
\href{https://doi.org/10.1104/pp.94.3.1024}{Intercellular diffusion
limits to {CO}\(_{\textrm{2}}\) uptake in leaves: Studies in air and
helox}. \emph{Plant Physiology} 94: 1024--1032.

\leavevmode\vadjust pre{\hypertarget{ref-peat_comparative_1994}{}}%
Peat, H. J., and A. H. Fitter. 1994. A comparative study of the
distribution and density of stomata in the {British} flora.
\emph{Biological Journal of the Linnean Society} 52: 377--393.

\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_phylogenetic_2022}{}}%
Pejhanmehr, M. 2022. Phylogenetic affinities among
{\textless{}}i{\textgreater{}}{Sida}{\textless{}}{\textbackslash{}}i{\textgreater{}}
species and allied genera ({Malvaceae}: {Malveae}), and examination of
{\textless{}}i{\textgreater{}}{Sida}
fallax{\textless{}}{\textbackslash{}}i{\textgreater{}} within the
{Hawaiian} {Islands} and throughout the {Pacific}. PhD thesis.
University of Hawai'i at Mānoa.

\leavevmode\vadjust pre{\hypertarget{ref-poorter_metaanalysis_2019}{}}%
Poorter, H., Ü. Niinemets, N. Ntagkas, A. Siebenkäs, M. Mäenpää, S.
Matsubara, and T. L. Pons. 2019.
\href{https://doi.org/10.1111/nph.15754}{A meta‐analysis of plant
responses to light intensity for 70 traits ranging from molecules to
whole plant performance}. \emph{New Phytologist}: nph.15754.

\leavevmode\vadjust pre{\hypertarget{ref-r_core_team_r:_2023}{}}%
R Core Team. 2023. \href{http://www.R-project.org/}{R: {A} {Language}
and {Environment} for {Statistical} {Computing}}. R Foundation for
Statistical Computing, Vienna, Austria.

\leavevmode\vadjust pre{\hypertarget{ref-robichaux_evolution_1984}{}}%
Robichaux, R. H., and R. W. Pearcy. 1984.
\href{https://doi.org/10.1002/j.1537-2197.1984.tb12492.x}{Evolution of
{C}\(_{\textrm{3}}\) and {C}\(_{\textrm{4}}\) plants along an
environmental moisture gradient: Patterns of photosynthetic
differentiation in {Hawaiian}
{\textless{}}i{\textgreater{}}{Scaevola}{\textless{}}{\textbackslash{}}i{\textgreater{}}
and
{\textless{}}i{\textgreater{}}{Euphorbia}{\textless{}}{\textbackslash{}}i{\textgreater{}}
species}. \emph{American Journal of Botany} 71: 121--129.

\leavevmode\vadjust pre{\hypertarget{ref-sack_developmental_2016}{}}%
Sack, L., and T. N. Buckley. 2016.
\href{https://doi.org/10.1104/pp.16.00476}{The developmental basis of
stomatal density and flux}. \emph{Plant Physiology} 171: 2358--2363.

\leavevmode\vadjust pre{\hypertarget{ref-salisbury_i_1928}{}}%
Salisbury, E. J. 1928. \href{https://doi.org/10.1098/rstb.1928.0001}{I.
{On} the causes and ecological significance of stomatal frequency, with
special reference to the woodland flora}. \emph{Philosophical
Transactions of the Royal Society of London. Series B, Containing Papers
of a Biological Character} 216: 1--65.

\leavevmode\vadjust pre{\hypertarget{ref-santrucek_partitioning_2019}{}}%
Šantrůček, J., L. Schreiber, J. Macková, M. Vráblová, J. Květoň, P.
Macek, and J. Neuwirthová. 2019. Partitioning of mesophyll conductance
for {CO}\(_{\textrm{2}}\) into intercellular and cellular components
using carbon isotope composition of cuticles from opposite leaf sides.
\emph{Photosynthesis Research}: 1--19.

\leavevmode\vadjust pre{\hypertarget{ref-schneider_nih_2012}{}}%
Schneider, C. A., W. S. Rasband, and K. W. Eliceiri. 2012.
\href{https://doi.org/10.1038/nmeth.2089}{{NIH} {Image} to {ImageJ}: 25
years of image analysis}. \emph{Nature Methods} 9: 671--675.

\leavevmode\vadjust pre{\hypertarget{ref-scoffoni_light-induced_2015}{}}%
Scoffoni, C., J. Kunkle, J. Pasquet-Kok, C. Vuong, A. J. Patel, R. A.
Montgomery, T. J. Givnish, and L. Sack. 2015.
\href{https://doi.org/10.1111/nph.13346}{Light-induced plasticity in
leaf hydraulics, venation, anatomy, and gas exchange in ecologically
diverse {Hawaiian} lobeliads}. \emph{New Phytologist} 207: 43--58.

\leavevmode\vadjust pre{\hypertarget{ref-sellers_modeling_1997}{}}%
Sellers, P. J., R. E. Dickinson, D. A. Randall, A. K. Betts, F. G. Hall,
J. A. Berry, G. J. Collatz, et al. 1997.
\href{https://doi.org/10.1126/science.275.5299.502}{Modeling the
exchanges of energy, water, and carbon between continents and the
atmosphere}. \emph{Science} 275: 502--509.

\leavevmode\vadjust pre{\hypertarget{ref-smith_leaf_1997}{}}%
Smith, W. K., T. C. Vogelmann, E. H. DeLucia, D. T. Bell, and K. A.
Shepherd. 1997. Leaf form and photosynthesis. \emph{BioScience} 11:
785--793.

\leavevmode\vadjust pre{\hypertarget{ref-sporck_hawaiian_2011}{}}%
Sporck, M. J. 2011. The {Hawaiian} {C}\(_{\textrm{4}}\) \emph{euphorbia}
adaptive radiation: An ecophysiological approach to understanding leaf
trait variation. Ph.\{D\}. University of Hawaii.

\leavevmode\vadjust pre{\hypertarget{ref-stan_development_team_stan_2023}{}}%
Stan Development Team. 2023. \href{https://mc-stan.org}{Stan {Modeling}
{Language} {Users} {Guide} and {Reference} {Manual}}.

\leavevmode\vadjust pre{\hypertarget{ref-theroux-rancourt_maximum_2021}{}}%
Théroux-Rancourt, G., A. B. Roddy, J. M. Earles, M. E. Gilbert, M. A.
Zwieniecki, C. K. Boyce, D. Tholen, et al. 2021.
\href{https://doi.org/10.1098/rspb.2020.3145}{Maximum
{CO}\(_{\textrm{2}}\) diffusion inside leaves is limited by the scaling
of cell size and genome size}. \emph{Proceedings of the Royal Society B:
Biological Sciences} 288: 20203145.

\leavevmode\vadjust pre{\hypertarget{ref-tholen_mechanistic_2011}{}}%
Tholen, D., and X.-G. Zhu. 2011.
\href{https://doi.org/10.1104/pp.111.172346}{The mechanistic basis of
internal conductance: {A} theoretical analysis of mesophyll cell
photosynthesis and {CO}\(_{\textrm{2}}\) diffusion}. \emph{Plant
Physiology} 156: 90--105.

\leavevmode\vadjust pre{\hypertarget{ref-tomas_importance_2013}{}}%
Tomás, M., J. Flexas, L. Copolovici, J. Galmés, L. Hallik, H. Medrano,
M. Ribas-Carbó, et al. 2013.
\href{https://doi.org/10.1093/jxb/ert086}{Importance of leaf anatomy in
determining mesophyll diffusion conductance to {CO2} across species:
Quantitative limitations and scaling up by models}. \emph{Journal of
Experimental Botany} 64: 2269--2281.

\leavevmode\vadjust pre{\hypertarget{ref-wang_towards_2017}{}}%
Wang, H., I. C. Prentice, T. F. Keenan, T. W. Davis, I. J. Wright, W. K.
Cornwell, B. J. Evans, and C. Peng. 2017.
\href{https://doi.org/10.1038/s41477-017-0006-8}{Towards a universal
model for carbon dioxide uptake by plants}. \emph{Nature Plants} 3:
734--741.

\leavevmode\vadjust pre{\hypertarget{ref-wang_theoretical_2020}{}}%
Wang, Y., J. S. Sperry, W. R. L. Anderegg, M. D. Venturas, and A. T.
Trugman. 2020. \href{https://doi.org/10.1111/nph.16572}{A theoretical
and empirical assessment of stomatal optimization modeling}. \emph{New
Phytologist} 227: 311--325.

\leavevmode\vadjust pre{\hypertarget{ref-wood_physiology_1934}{}}%
Wood, J. G. 1934. The physiology of xerophytism in {Australian} plants:
The stomatal frequencies, transpiration and osmotic pressures of
sclerophyll and tomentose-succulent leaved plants. \emph{Journal of
Ecology} 22: 69--87.

\leavevmode\vadjust pre{\hypertarget{ref-xiong_one_2020}{}}%
Xiong, D., and J. Flexas. 2020.
\href{https://doi.org/10.1111/nph.16801}{From one side to two sides: The
effects of stomatal distribution on photosynthesis}. \emph{New
Phytologist} 228: 1754--1766.

\leavevmode\vadjust pre{\hypertarget{ref-yorkston_interfertility_2006}{}}%
Yorkston, M., and C. C. Daehler. 2006.
\href{https://doi.org/10.1086/499615}{Interfertility between {Hawaiian}
ecotypes of \emph{{Sida} fallax} ({Malvaceae}) and evidence of a hybrid
disadvantage}. \emph{International Journal of Plant Sciences} 167:
221--230.

\end{CSLReferences}

\end{document}

### 2023-09-06 ms/si.qmd
initial commit all files

---
format:
  pdf:
    keep-tex: true  
    include-in-header: 
      - preamble.tex
    toc: false
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
mainfont: Times New Roman
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(readr)

# From 15_archive-data.R
sum_data = read_rds("../objects/sum_data.csv")

```

```{=tex}
\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\renewcommand\theequation{S\arabic{equation}}    
\setcounter{figure}{0}    
\setcounter{table}{0}    
\setcounter{equation}{0}
```

```{r, sum-data, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
#| label: sum-data
#| tbl-cap: "Average traits values among ʻilima leaves at each site. $\\mathrm{SD}_\\text{abaxial}$ is the stomatal density per mm$^2$ on the abaxial (lower) surface; $\\mathrm{SD}_\\text{adaxial}$ is the stomatal density per mm$^2$ on the adaxial (upper) surface; $\\mathrm{GCL}_\\text{abaxial}$ is the guard cell length in $\\mu$m on the abaxial (lower) surface; $\\mathrm{GCL}_\\text{adaxial}$ is the guard cell length in $\\mu$m on the adaxial (upper) surface; Leaf thickness is the length from upper cuticle to lower cuticle in $\\mu$m; $A$ is the photosynthetic rate in $\\mu \\text{mol CO}_2~\\text{m}^{-2}~\\text{s}^{-1}$; $g_\\mathrm{sw}$ is the stomatal conductance to water vapor in $\\text{mol}~\\text{m}^{-2}~\\text{s}^{-1}$."
#| 
sum_data |>
  knitr::kable(format = "latex", digits = c(rep(0, 3), rep(2, 6), 3),
               booktabs = TRUE, escape = FALSE, longtable = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "1.5cm") |>
  kableExtra::column_spec(3, width = "1.5cm") |>
  kableExtra::landscape()

```

\newpage

```{=tex}
\begin{figure}
  \includegraphics{../figures/licor.pdf}
  \caption{See next page.}
  \label{fig:licor}
\end{figure}
```

\newpage

\setcounter{figure}{\numexpr\value{figure}-1\relax}

```{=tex}
\begin{figure}
  \caption{(Continued from previous page.) Individual $A \textendash g_\mathrm{sw}$ curves used to estimate $\mathrm{AA}$. For each leaf, one per site, we measured $A$ ($y$-axis) over a range of $g_\mathrm{sw}$ ($x$-axis) on the same leaf with two treatments: `amphi' (circles, solid line) leaves were untreated; `pseudohypo' (squares, dashed line) leaves had no conductance through the upper (adaxial) surface. In all coastal (orange) and montane (blue) leaves, we fit generalized additive models and 95\% confidence ribbons to estimate $\mathrm{AA}$ at a $g_\mathrm{sw}$ where the curves overlap (vertical black line). In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap. $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
  \label{fig:licor}
\end{figure}
```

\newpage

```{=tex}
\begin{figure}
  \includegraphics{../figures/habitat-Ags.pdf}
  \caption{The photsynthetic rate (left facet) and stomatal conductance to water vapor (right facet) of montane (blue) and coastal (orange) ʻilima leaves. Each point-interval is the median posterior estimate plus 95\% confidence interval of trait value for that habitat. Smaller points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants, one per site. $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
  \label{fig:habitat-Ags}
\end{figure}
```

\newpage

```{=tex}
\begin{figure}
  \includegraphics{../figures/habitat-gmaxratio.pdf}
  \caption{The $g_\mathrm{smax,ratio}$ ($y$-axis) of montane (blue) and coastal (orange) ʻilima leaves. Each point-interval is the median posterior estimate plus 95\% confidence interval of $g_\mathrm{smax,ratio}$ for that site. Sites are arranged by habitat and ascending $g_\mathrm{smax,ratio}$ within habitat. Smaller, transparent points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants. $g_\mathrm{smax,ratio}$, the ratio of anatomical maximum stomatal conductance to water vapor on the the adaxial surface over the total.}
  \label{fig:habitat-gmaxratio}
\end{figure}
```

### 2023-09-06 ms/si.tex
initial commit all files

--- 2023-09-06 ms/ms.tex
@@
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \KOMAoption{captions}{tableheading}
-\addtokomafont{disposition}{\rmfamily}
-\RedeclareSectionCommand[
-  font=\normalfont\Large]{section}
-\RedeclareSectionCommand[
-  font=\normalfont\normalsize\bfseries]{subsection}
-\RedeclareSectionCommand[
-  font=\normalfont\normalsize\itshape]{subsubsection}
-\RedeclareSectionCommand[
-  font=\normalfont\normalsize]{paragraph}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (Sida fallax)},
   colorlinks=true,
@@
 
-\title{Amphistomy increases leaf photosynthesis more in coastal than
-montane plants of Hawaiian ʻilima (\emph{Sida fallax})}
 \author{}
@@
 \begin{document}
-\maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi
-
-\begin{center}
-Genevieve Triplett$^1$, Thomas N. Buckley$^2$, Christopher D. Muir$^{1,3,*}$
-\end{center}
-
-\(^1\) School of Life Sciences, University of Hawaiʻi Mānoa, Honolulu,
-HI, USA 96822
-
-\(^2\) School of Life Sciences, University of California, Davis, CA, USA
-95616
-
-\(^3\) Department of Botany, University of Wisconsin, Madison, WI, USA
-53706
-
-\(^*\) Corresponding author: Christopher D. Muir, cdmuir@wisc.edu
-
-Manuscript received \_\_\_\_\_\_\_; revision accepted \_\_\_\_\_\_\_.
-
-Running head: Amphistomy advantage in ʻilima
-
-\newpage
-
-\hypertarget{abstract}{%
-\section{ABSTRACT ---}\label{abstract}}
-
-\noindent \textbf{Premise of the study} The adaptive significance of
-stomata on both upper and lower leaf surfaces, called amphistomy, is
-unresolved. A widespread association between amphistomy and open, sunny
-habitats suggests the adaptive benefit of amphistomy may be greatest in
-these contexts, but this hypothesis has not been tested experimentally.
-Understanding why amphistomy evolves can inform its potential as a
-target for crop improvement and paleoenvironment reconstruction.
-
-\textbf{Methods} We developed a new method to quantify ``amphistomy
-advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
-amphistomatous leaf to that of the same leaf with the same total
-stomatal conductance but with gas exchange blocked through the upper
-(adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated
-\(\mathrm{AA}\) and related physiological and anatomical traits in 12
-populations, six coastal (open, sunny) and six montane (closed, shaded),
-of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).
-
-\textbf{Key results} Coastal ʻilima leaves benefit 4.58 times more from
-amphistomy compared to their montane counterparts. Our evidence was
-equivocal with respect to two hypotheses -- that coastal leaves benefit
-more because they are thicker and therefore have lower CO\(_2\)
-conductance through the internal airspace, and that leaves with more
-similar conductances at each surface benefit more than leaves with few
-stomata on the upper surface.
-
-\textbf{Conclusions} This is the first direct experimental evidence that
-amphistomy \emph{per se} increases photosynthesis, consistent with the
-hypothesis that parallel pathways through upper and lower mesophyll
-increase the supply of CO\(_2\) to chloroplasts. The prevalence of
-amphistomatous leaves in open, sunny habitats can partially be explained
-the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic
-basis of this observation is an area for future research.
-
-\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, leaf, \emph{Sida
-fallax}, stomata
-
-\hypertarget{introduction}{%
-\section{INTRODUCTION ---}\label{introduction}}
-
-Amphistomy, the presence of stomata on both lower and upper surfaces of
-broad leaves, should increase carbon gain by reducing the average
-diffusion pathlength between stomata and chloroplasts
-(\autoref{fig:compare-circuit-diagrams}), yet paradoxically this
-seemingly simple adaptation is uncommon in nature and we don't know why.
-Understanding variation in stomatal traits like amphistomy is imperative
-because these tiny pores play an outsized ecological role in the global
-carbon and water cycles (Hetherington and Woodward, 2003; Berry et al.,
-2010). A widely applicable, accurate representation of how stomata
-mediate the relationship between CO\(_2\) gained through photosynthesis
-and water lost through evapotranspiration is essential to predict future
-climate using Earth Systems Models (Jarvis, 1976; Ball et al., 1987;
-Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997). Optimality
-models accurately predict the major cause of water loss, stomatal
-conductance (\(g_\mathrm{sw}\)), by assuming plants maximize carbon gain
-minus a cost of water (Cowan and Farquhar, 1977; Givnish, 1986; Medlyn
-et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks et al., 2018;
-Deans et al., 2020; Franklin et al., 2020; Wang et al., 2020; Harrison
-et al., 2021). Despite the success of optimality modeling in predicting
-\(g_\mathrm{sw}\), the same modeling approach has so far failed to
-explain the rarity of amphistomatous leaves (Muir, 2019). \textbf{This
-gap between theory and observations strongly implies that we remain
-ignorant about some key benefits and costs associated with stomata.}
-
-Where are amphistomatous leaves found and why aren't they more common?
-Among terrestrial flowering plants, amphistomatous leaves are rarely
-found on woody plants and shade-tolerant herbs, but they are common in
-annual and perennial herbs from sunny habitats (Salisbury, 1928;
-Parkhurst, 1978; Mott et al., 1982; Peat and Fitter, 1994; Gibson, 1996;
-Jordan et al., 2014; Muir, 2015, 2018; Bucher et al., 2017). Even in
-resupinate leaves where the abaxial surface faces up toward the sky,
-stomata develop on the lower adaxial surface (Lyshede, 2002). Exceptions
-to this general pattern include some arid woody plants which typically
-have vertically oriented, isobilateral leaves (Wood, 1934; Jordan et
-al., 2014; Boer et al., 2016; Drake et al., 2019) and
-floating/amphibious leaves of aquatic plants (Kaul, 1976; Doll et al.,
-2021). The dearth of amphistomatous leaves should be quite surprising
-and has been described as one of the most important unsolved problems in
-the study of leaf structure-function relations despite some recent
-progress (Grubb, 1977, 2020).
-
-Amphistomatous leaves should be common because, all else being equal, a
-leaf with a given number of stomata per area could increase its
-photosynthetic rate simply by apportioning approximately half its
-stomata to each surface (Parkhurst, 1978; Gutschick, 1984a, b). The key
-difference between a hypo- and amphistomatous leaf, holding all other
-factors constant, is that an amphistomatous leaf has two parallel
-diffusion paths through the internal airspace to any given chloroplast.
-Those airspaces pose a resistance for CO\(_2\) diffusion, so CO\(_2\)
-concentration drops as it approaches chloroplasts. Shorter pathways mean
-a smaller drop in CO\(_2\) concentration. Thus, chloroplasts in
-amphistomatous leaves experience higher CO\(_2\) concentrations than in
-hypostomatous leaves, thereby increasing photosynthesis. The airspace
-resistance (or its inverse, the airspace conductance,
-\(g_\mathrm{ias}\)) is rarely measured directly and there is
-disagreement between empirical (Parkhurst and Mott, 1990; Morison et
-al., 2005; Evans et al., 2009; Tomás et al., 2013; Earles et al., 2018;
-Šantrůček et al., 2019; Nobel, 2020; Harwood et al., 2021; Márquez et
-al., 2023) and theoretical models (Tholen and Zhu, 2011; Ho et al.,
-2016; Théroux-Rancourt et al., 2021). The \(g_\mathrm{ias}\) in thin,
-porous leaves may be so large as to be inconsequential given much lower
-conductances for other components of the diffusion pathway, whereas the
-\(g_\mathrm{ias}\) of thick leaves with little airspace may greatly
-hinder CO\(_2\) diffusion to chloroplasts. Amphistomy should confer the
-largest photosynthetic benefit in leaves with intrinsically low
-\(g_\mathrm{ias}\).
-
-Amphistomy should also enhance photosynthesis when leaf boundary layer
-resistance is high, because apportioning total flux between two boundary
-layers rather than one results in a smaller CO\(_2\) concentration drop
-between the atmosphere and stomata. A similar effect has been validated
-with a computer model and measurements for transpiration: amphistomatous
-leaves lose somewhat more water for the same vapor pressure deficit and
-total \(g_\mathrm{sw}\) (Foster and Smith, 1986), but the additional
-carbon gain should be enough to offset this cost under most realistic
-conditions (Muir, 2019).
-
-The most promising adaptive hypothesis is that amphistomy is important
-for maximizing photosynthetic rate under high light. Mott et al. (1982)
-proposed that ``plants with a high photosynthetic capacity, living in
-full-sun environments, and experiencing rapidly fluctuating or
-continuously available soil water'' would benefit most, in terms of
-increased carbon gain, from having amphistomatous leaves. As described
-above, herbs from sunny habitats are often amphistomatous. Most
-variation in stomatal density ratio (\(\mathrm{SR}\), the ratio of
-stomatal density between the upper and lower surfaces) among species is
-assumed to be genetic, but there is also putatively adaptive plasticity
-in response to light. Leaves of \emph{Ambrosia cordifolia}, a desert
-perennial herb, are hypostomatous under low light (photosynthetic photon
-flux density, PPFD = 110 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
-but develop \textasciitilde20\% of their stomata on the upper surface
-under high light (1700 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
-(Mott and Michaelson, 1991). Similarly, \emph{Solanum lycopersicum}
-leaves are hypostomatous when grown in the shade but develop
-\textasciitilde20\% of their stomata on the upper surface grown under
-high light-intensity (Gay and Hurd, 1975). Adult leaves of
-\emph{Eucalyptus globulus} are amphistomatous, but the proportion of
-adaxial stomata increases from \textasciitilde10-20\% under low light to
-\textasciitilde30-40\% under high light (James and Bell, 2001). In
-summary, both genetic and plastic responses to evince a widespread
-association between light and \(\mathrm{SR}\).
-
-The association between high light and amphistomy suggests that `sun'
-leaves have the most to gain in terms of increased photosynthesis from
-having stomata on both surfaces, as Mott et al. (1982) hypothesized.
-Parkhurst (1978) proposed quantifying this benefit as `amphistomy
-advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
-(see Materials and Methods). This hypothesis has never been tested
-directly by comparing the photosynthetic rate of an amphistomatous leaf
-to that of an otherwise identical hypostomatous leaf with the same total
-stomatal conductance under the same conditions. We propose a
-straightforward method to do this by experimentally creating a
-pseudohypostomatous leaf with gas exchange blocked through the upper
-surface (see Materials and Methods). One reason that sun leaves might
-have greater \(\mathrm{AA}\) is that they are usually thicker or denser
-(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
-(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
-leaves have a stomatal density ratio closer to 0.5 (same density on each
-leaf surface), this will confer a greater advantage than an
-amphistomatous leaf with most stomata on one surface. In other words,
-amphistomy doesn't make much difference if one leaf surface has very
-open stomata on it. We therefore predict that sun leaves will have
-greater \(\mathrm{AA}\) possibly because they have thicker leaves and/or
-\(\mathrm{SR}\) closer to 0.5. We actually report
-\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
-that takes into account differences in both stomatal density and size
-between surfaces.
-
-The native flora of the Hawaiian archipelago is a excellent system to
-test the relationship between light habitat and \(\mathrm{AA}\). Many
-lineages have adapted to different light habitats after colonization and
-leaf anatomical traits such as \(\mathrm{SR}\) and thickness vary within
-and among closely related species. It is hypothesized that the common
-ancestor in many Hawaiian clades was a ``weedy'' species with high
-dispersal ability adapted to open habitats (Carlquist, 1966).
-Colonization was followed by adaptive radiation into higher elevation,
-montane, closed, forested habitats. Consequently, adaptation to sun and
-shade is a common axis of phenotypic variation among Hawaiian plants
-such as lobeliads (Givnish et al., 2004; Montgomery and Givnish, 2008;
-Givnish et al., 2009; Givnish and Montgomery, 2014; Scoffoni et al.,
-2015), \emph{Bidens} (Carlquist, 1966; Knope et al., 2020),
-\emph{Scaevola} (Robichaux and Pearcy, 1984; McKown et al., 2016),
-\emph{Euphorbia} (Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al.,
-2009).
-
-Here we focus on variation within an indigenous plant species \emph{Sida
-fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
-ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
-Hawaiian islands. Coastal populations are morphologically different from
-montane populations (\autoref{fig:study-system}). Coastal regions of
-Hawaiʻi are characterized by high sun exposure, warmer temperatures,
-high winds, salinity, and variation in water availability. Coastal
-populations of ʻilima tend to be short and prostrate which likely helps
-them to withstand the windy environment (\autoref{fig:study-system}a).
-The leaves of these populations are covered on both surfaces in dense,
-soft hairs that give the leaves a silvery green appearance
-(\autoref{fig:study-system}b), which helps mitigate water loss by
-reflecting solar radiation, thereby lowering leaf temperature
-(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
-provide very different challenges. Many other tall species grow on the
-slopes of these wet mountainous regions, which makes light competition a
-factor that plants may need to adapt to. Possibly due to this, montane
-populations are erect and shrub- or tree-like, capable of growing meters
-tall with strong, woody stems. These individuals have smooth, green
-foliage with serrated edges. Montane populations exhibit traits that
-help them to compete for light availability (Raunch et al.~1997). This
-montane morphology is not found in \emph{S. fallax} populations on other
-Pacific Islands (Pejhanmehr, 2022). The variable morphology between
-coastal and montane populations has even led some scientists to argue
-that it makes up a species complex, rather than a single species (Raunch
-et al.~1997).
-
-The objectives of our study are to test whether 1) `sun' leaves of
-coastal ʻilima plants will have greater \(\mathrm{AA}\) than `shade'
-leaves of montane plants; and if so, is this because 2a) coastal plants
-have thicker leaves than montane plants and/or 2b) coastal plants have a
-\(g_{\mathrm{smax,ratio}}\) closer to 0.5?
-
-\begin{figure}[H]
-  \includegraphics{../figures/study-system.pdf}
-  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
-  \label{fig:study-system}
-\end{figure}
-
-\hypertarget{methods}{%
-\section{MATERIALS AND METHODS ---}\label{methods}}
-
-\hypertarget{plant-sampling}{%
-\subsection{Plant sampling ---}\label{plant-sampling}}
-
-We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on
-Hawaiʻi Island by consulting (Yorkston and Daehler, 2006) and citizen
-scientist records on iNaturalist (Anon, 2022)
-(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
-that appeared to be cultivated. We visited sites between August and
-November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
-in one three-day trip. We haphazardly sampled eight plants distributed
-evenly between the highest and lowest elevation plants along a transect
-at each site. For safety and conservation reasons, transects were along
-a trail or road. We did not sample small individuals if there was risk
-removing leaves would cause mortality. From each plant, we collected two
-fully expanded leaves for traits measurements. We sampled stomatal
-traits on all leaves; leaf thickness on one leaf from three randomly
-selected plants per site; and, due to limited time, a single leaf from a
-single plant at the middle of each transect for gas exchange
-measurements.
-
-\hypertarget{tbl-locations}{}
-\begin{table}
-\caption{\label{tbl-locations}ʻIlima study sites for this study. }\tabularnewline
-
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
-\toprule
-Site & Island & Habitat & Latitude & Longitude & Elevation (mas)\\
-\midrule
-\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.710} & \cellcolor{gray!6}{-157.982} & \cellcolor{gray!6}{4}\\
-Kaloko beach & Oʻahu & coastal & 21.293 & -157.661 & 4\\
-\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{19.676} & \cellcolor{gray!6}{-156.024} & \cellcolor{gray!6}{6}\\
-Kaʻena Point & Oʻahu & coastal & 21.574 & -158.278 & 4\\
-\cellcolor{gray!6}{Makapuʻu beach} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.313} & \cellcolor{gray!6}{-157.661} & \cellcolor{gray!6}{3}\\
-\addlinespace
-Puakō petroglyph park & Hawaiʻi & coastal & 19.957 & -155.858 & 8\\
-\cellcolor{gray!6}{Hawaiʻi loa ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.294} & \cellcolor{gray!6}{-157.727} & \cellcolor{gray!6}{352}\\
-Hāloa ʻĀina & Hawaiʻi & montane & 19.552 & -155.793 & 1567\\
-\cellcolor{gray!6}{Kaʻohe game management area} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{19.817} & \cellcolor{gray!6}{-155.616} & \cellcolor{gray!6}{1946}\\
-Koaiʻa tree sanctuary & Hawaiʻi & montane & 20.048 & -155.737 & 970\\
-\addlinespace
-\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.305} & \cellcolor{gray!6}{-157.779} & \cellcolor{gray!6}{248}\\
-Waʻahila ridge & Oʻahu & montane & 21.314 & -157.793 & 357\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{leaf-traits}{%
-\subsection{Leaf traits ---}\label{leaf-traits}}
-
-\hypertarget{stomata}{%
-\subsubsection{Stomata ---}\label{stomata}}
-
-We estimated the stomatal density and size on ab- and adaxial leaf
-surfaces from all leaves. For pubescent leaves (usually coastal), we
-dried and pressed leaves for \(\approx 1\) week (Hill et al., 2014),
-carefully scraped trichomes off with a razor blade, and rehydrated the
-leaf. For glabrous leaves, we used fresh leaves. We applied clear nail
-polish to both leaf surfaces of fresh or rehydrated leaves in the middle
-of the lamina away from major veins. After nail polish dried, we mounted
-impressions on a microscope slide using transparent tape (Mott and
-Michaelson, 1991). We digitized a portion of each leaf surface
-impression using a brightfield microscope (Leica DM2000, Wetzlar,
-Germany). We counted all stomata and divided by the visible leaf area
-(0.890 mm\(^2\)) to estimate density and measured guard cell length from
-five randomly chosen stomata per field using ImageJ (Schneider et al.,
-2012).
-
-\hypertarget{leaf-thickness}{%
-\subsubsection{Leaf thickness ---}\label{leaf-thickness}}
-
-We cut thin sections using two razor blades taped together and cutting
-the leaf in a petri dish of water, wet-mounted sections onto a slide,
-and took digital micrographs using a brightfield microscope, as
-described above.
-
-\hypertarget{gas-exchange-measurements}{%
-\subsubsection{Gas exchange measurements
----}\label{gas-exchange-measurements}}
-
-At each site, we selected one representative leaf from one plant near
-the middle of the transect for gas exchange measurements using a
-portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln,
-NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
-conductance to water vapor (\(g_\text{sw}\)) at saturating light
-(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
-ambient CO\(_2\) (415 ppm), and
-\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
-`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
-but with modification. For each leaf, we measured the photosynthetic
-rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
-range of \(g_\text{sw}\) values, holding other environmental variables
-constant. We refer to this as an \(A \textendash g_\text{sw}\) curve. We
-compared the \(A \textendash g_\text{sw}\) curve of the untreated leaf
-to the photosynthetic rate of ``pseudohypostomatous'' leaf
-(\(A_\mathrm{hypo}\)), which is the same leaf but with gas exchange
-through the upper surface blocked by a neutral density plastic
-(propafilm). We refer to the untreated and partially blocked leaves as
-``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
-calculated as the log-response ratio of \(A\) compared at the same total
-\(g_\text{sw}\):
-
-\[\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})\]
-
-The log-response ratio is easy to interpret because values above 0
-indicate a photosynthetic advantage of amphistomy, whereas values less
-than 0 indicate a disadvantage. The response ratio is also better than
-the absolute difference because it indicates a proportional change in
-\(A\), which facilitates comparisons across leaves and environments with
-different baseline photosynthetic rates. The irradiance of the light
-source in the pseudohypo leaf is higher because the propafilm reduces
-transmission. Because the propafilm reduces transmission, we used a
-higher incident \(\mathrm{PPFD}\) for pseudohypo leaves (higher by a
-factor 1/0.91, the inverse of the measured transmissivity of the
-propafilm). We also set the stomatal conductance ratio, for purposes of
-calculating boundary layer conductance, to 0 for pseudohypo leaves
-following manufacturer directions.
-
-To control for order effects, we alternated between starting with amphi
-(untreated) or pseudohypo (treated) leaf measurements. In the field, we
-acclimated the focal leaf to high light
-(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
-high relative humidity (\(\mathrm{RH} = 70\%\)),
-\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
-until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
-as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
-decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
-closure without photosynthetic downregulation. We logged data until
-\(g_\text{sw}\) reached its nadir. We then repeated the process with the
-other treatment (amphi or pseudohypo).
-
-\hypertarget{data-analyis}{%
-\subsection{Data analysis ---}\label{data-analyis}}
-
-\hypertarget{objective-1-do-coastal-leaves-have-greater-amphistomy-advantage-than-montane-leaves}{%
-\subsubsection{Objective 1: Do coastal leaves have greater amphistomy
-advantage than montane leaves?
----}\label{objective-1-do-coastal-leaves-have-greater-amphistomy-advantage-than-montane-leaves}}
-
-It is not feasible to record \(A_\mathrm{amphi}\) and
-\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
-we fit \(A \textendash g_\text{sw}\) curves using a flexible generalized
-additive model (GAM) to interpolate modeled \(A\) for amphi and
-pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12 leaves, the
-minimum \(g_\text{sw}\) of the amphi curve was smaller than the maximum
-\(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves overlapped for
-a range of \(g_\text{sw}\) values). In those cases, we estimated
-\(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle of the
-range of overlap. In 2 of 12 leaves, the \(A \textendash g_\text{sw}\)
-curves did not quite overlap because the minimum \(g_\text{sw}\) of the
-amphi curve was slightly greater than the maximum \(g_\text{sw}\) of the
-pseudohypo curve. In those cases, we estimated \(\mathrm{AA}\) by
-extrapolating slightly,
-\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
-beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
-curves. We fit GAMs using the \emph{R} package \textbf{brms} version
-2.20.1 (Bürkner, 2017) for Bayesian statistical inference with MCMC
-sampling in \emph{Stan} (Stan Development Team, 2023). We used CmdStan
-version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
-2023) to interface with \emph{R} version 4.3.1 (R Core Team, 2023). We
-sampled from posterior distribution from 4 chains with 1000 iterations
-each after 1000 warmup iterations. We estimated parameters and
-confidence intervals as the median and 95\% quantile intervals of the
-posterior, respectively. The key prediction is that
-\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).
-
-\hypertarget{objective-2a-are-coastal-leaves-thicker-than-montane-leaves}{%
-\subsubsection{Objective 2a: Are coastal leaves thicker than montane
-leaves?
----}\label{objective-2a-are-coastal-leaves-thicker-than-montane-leaves}}
-
-We tested whether leaf thickness (log-transformed) varied between
-coastal and montane populations and among individuals within populations
-using a Bayesian mixed-effects model with site-type as a fixed effect
-and individual plant and site as random effects. We used the \emph{R}
-package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
-\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.32.2
-and \textbf{cmdstanr} version 0.5.3 (Gabry and Češnovar, 2023). We
-sampled the posterior distribution from 4 chains with 1000 iterations
-each after 1000 warmup iterations. We estimated the relationship between
-population average leaf thickness and \(\mathrm{AA}\) measured from a
-single individual per population. We used this approach most of the
-variation in leaf thickness was among sites (see Results) and the plant
-selected for gas exchange measurements was not always among the plants
-randomly selected for leaf thickness, precluding individual level
-correlation. We propagated uncertainty about in \(\mathrm{AA}\) and leaf
-thickness by integrating over the entire posterior distribution sample
-for each variable. The key prediction is that the effect of leaf
-thickness on \(\mathrm{AA}\) is positive.
-
-\hypertarget{objective-2b-is-g_mathrmsmaxratio-closer-to-0.5-in-coastal-leaves-than-montane-leaves}{%
-\subsubsection{\texorpdfstring{Objective 2b: Is
-\(g_{\mathrm{smax,ratio}}\) closer to 0.5 in coastal leaves than montane
-leaves?
----}{Objective 2b: Is g\_\{\textbackslash mathrm\{smax,ratio\}\} closer to 0.5 in coastal leaves than montane leaves?
----}}\label{objective-2b-is-g_mathrmsmaxratio-closer-to-0.5-in-coastal-leaves-than-montane-leaves}}
-
-We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
-montane populations and among individuals within populations using a
-Bayesian mutliresponse, mixed-effects model. The modeled response
-variables are stomatal count and guard cell length on each surface.
-Counts were modeled as negative binomially distributed variable from a
-latent stomatal density and a parameter \(\phi\) to estimate
-overdispersion in counts relative to a Poisson model. For all traits,
-the explanatory variables were site-type as a fixed effect and leaf
-within individual plant, individual plant, and site as random effects.
-We used the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner,
-2017) to fit the model in \emph{Stan} (Stan Development Team, 2023) with
-CmdStan version 2.32.2 and \textbf{cmdstanr} version 0.5.3 (Gabry and
-Češnovar, 2023). We interpolated missing adaxial guard cell lengths from
-6 out of 185 samples with zero adaxial stomata using the ``mi'' function
-in \textbf{brms} package. We sampled the posterior distribution from 4
-chains with 1000 iterations each after 1000 warmup iterations. From each
-posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:
-
-\[g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},\]
-where \(g_\mathrm{smax,lower}\) and \(g_\mathrm{smax,upper}\) are
-maximum stomatal conductance to water vapor at
-\(T_\mathrm{leaf} = 25 ^ \circ\) C on the lower and upper surface,
-respectively. The maximum stomatal conductance is calculated from
-stomatal density and length, assuming that stomata are fully open,
-following Sack and Buckley (2016).
-
-We estimated the relationship between leaf \(g_\mathrm{smax,ratio}\) and
-\(\mathrm{AA}\) measured from a single leaf per population. We
-propagated uncertainty about in \(\mathrm{AA}\) and
-\(g_\mathrm{smax,ratio}\) by integrating over the entire posterior
-distribution sample for each variable. The key prediction is that the
-effect of \(g_\mathrm{smax,ratio}\) on \(\mathrm{AA}\) is positive until
-\(g_\mathrm{smax,ratio} < 0.5\).
-
-\hypertarget{results}{%
-\section{RESULTS ---}\label{results}}
-
-\hypertarget{amphistomy-advantage-is-greater-in-coastal-leaves}{%
-\subsection{Amphistomy advantage is greater in coastal leaves
----}\label{amphistomy-advantage-is-greater-in-coastal-leaves}}
-
-Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
-more than that of montane plants. \(\mathrm{AA}\) was significantly
-greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
-coastal leaves, but only 1 of 6 montane leaves
-(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
-curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
-leaves is \(0.12~[\numrange{0.078}{0.15}]\) and
-\(0.027~[\numrange{-0.0033}{0.057}]\), respectively; the difference in
-average \(\mathrm{AA}\) between habitat types is
-\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.04}{0.14}]\).
-The difference between coastal and montane leaves is lower,
-\(0.054~[\numrange{-0.012}{0.11}]\), but still greater than 0 if we
-remove two sites where we extrapolated slightly beyond fitted
-\(A \textendash g_\mathrm{sw}\) curves. Maximum photosynthetic rate was
-slightly, but not significantly higher in coastal leaves (Welch Two
-Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
-conductance was similar (Welch Two Sample \emph{t}-test,
-\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
-(\autoref{fig:habitat-Ags}).
-
-\begin{figure}[H]
-  \includegraphics{../figures/habitat-aa.pdf}
-  \caption{Coastal leaves benefit more amphistomy than montane leaves.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf.
-Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat.
-The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves.
-$g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:habitat-aa}
-\end{figure}
-
-\hypertarget{leaf-thickness-is-associated-with-amphistomy-advantage-between-but-not-within-habitats}{%
-\subsection{Leaf thickness is associated with amphistomy advantage
-between but not within habitats
----}\label{leaf-thickness-is-associated-with-amphistomy-advantage-between-but-not-within-habitats}}
-
-Coastal ʻilima leaves are \(91~[\numrange{26}{164}]\) \(\mu\text{m}\)
-thicker than their montane counterparts. Although coastal leaves are
-thicker and have greater \(\mathrm{AA}\), the relationship between leaf
-thickness and \(\mathrm{AA}\) within habitats is actually slightly
-negative (\autoref{fig:traits-aa}A;
-\(\text{slope}~= -0.11~[\numrange{-0.28}{0.033}]\)), the opposite
-direction our hypothesis predicts.
-
-\hypertarget{g_mathrmsmaxratio-is-not-associated-with-amphistomy-advantage}{%
-\subsection{\texorpdfstring{\(g_\mathrm{smax,ratio}\) is not associated
-with amphistomy advantage
----}{g\_\textbackslash mathrm\{smax,ratio\} is not associated with amphistomy advantage ---}}\label{g_mathrmsmaxratio-is-not-associated-with-amphistomy-advantage}}
-
-Coastal and montane leaves have similar average
-\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
-\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
-water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
-\(0.057~[\numrange{-0.16}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
-montane leaves, but the 95\% confidence intervals overlap 0 difference.
-The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
-sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
-and others with \(g_\mathrm{smax,ratio} > 0.2\)
-(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
-montane sites where those on the Big Island of Hawaiʻi all have low
-\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
-\(g_\mathrm{smax,ratio}\). There is no relationship between
-\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
-(\autoref{fig:traits-aa}B;
-\(\text{slope}~= 0.11~[\numrange{-0.032}{0.25}]\)) in our sample.
-
-\begin{figure}[H]
-  \includegraphics{../figures/traits-aa.pdf}
-  \caption{Relationships between leaf amphistomy advantage, (A) $g_\mathrm{smax,ratio}$ and (B) leaf thickness among ʻilima (*Sida fallax*) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi., and amphistomy advantage.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value.
-Each leaf is from a different montane (blue) or coastal (orange) site.
-Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
-$g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:traits-aa}
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{DISCUSSION ---}\label{discussion}}
-
-Amphistomy is a seemingly simple way that leaves can increase carbon
-gain without significant additional water loss, yet it is rare in nature
-and we do not know why. The strong association between amphiostomy and
-sunny, open habitats suggests that amphistomy may benefit sun leaves
-more than shade leaves, but progress has been limited by the lack of
-evidence that amphiostomy \emph{per se} affects improves photosynthesis
-in a given leaf. By experimentally blocking gas exchange through the
-upper surface in a controlled environment, we directly compared an
-amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
-This allows us to quantity the `amphistomy advantage' (\(\mathrm{AA}\))
-holding all else constant. Taking advantage of the of the steep climatic
-gradients in the Hawaiian archipegalo, we applied this new method to
-show for the first time that sun leaves benefit more from amphistomy
-than shade leaves in ʻilima (\emph{Sida fallax}). While this result has
-not yet been validated in other speciesthere are limits to how much we
-can generalize to all land plants, our results indicate that part of the
-reason amphistomatous leaves are found most commonly in high light
-habitats is that the adaptive benefit is greater in this environment.
-
-If \(\mathrm{AA}\) is typically greater in sun leaves than shade leaves,
-it could partially explain the distribution of amphi- and hypostoamtous
-leaves, but the precise mechanism(s) require further study. One
-hypothesis is that the internal airspace conductance,
-\(g_\mathrm{ias}\), from stomata to mesophyll cell walls is lower in
-thicker sun leaves (Parkhurst, 1978). All else being equal, a leaf with
-lower \(g_\mathrm{ias}\) will benefit more from amphistomy. Our results
-partially support this hypothesis. Coastal ʻilima leaves with high
-\(\mathrm{AA}\) (\autoref{fig:habitat-aa}) are thicker than montane
-leaves, but the relationship between \(\mathrm{AA}\) and leaf thickness
-within habitats is actually slightly negative
-(\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and
-montane leaves differ in many respects besides thickness, we do not have
-enough data to conclude that leaf thickness explains the variation in
-\(\mathrm{AA}\) between habitats. Alternatively, other biochemical or
-anatomical differences between coastal and montane leaves may explain
-why \(\mathrm{AA}\) is greater in coastal leaves. The negative
-relationship between leaf thickness and \(\mathrm{AA}\) could be
-explained if thicker leaves compensated by having a more porous
-mesophyll and/or less tortuous airspaces (Théroux-Rancourt et al.,
-2021).
-
-A second natural hypothesis is that amphistomatous leaves with few
-adaxial (upper) stomata benefit less than those with similar densities
-on both surfaces. We predicted that leaves with
-\(g_\mathrm{smax,ratio}\) closer to 0.5 would have higher
-\(\mathrm{AA}\) based on biophysical models (Gutschick, 1984a). The
-logic is that a small number of stomata on the upper surface are
-insufficient to supply the entire upper mesophyll due to limited lateral
-diffusion (Morison et al., 2005). Our results do not support this
-hypothesis. Montane leaves from Big Island sites had low
-\(g_\mathrm{smax,ratio}\) and low \(\mathrm{AA}\) whereas low montane
-leaves on Oʻahu had high \(g_\mathrm{smax,ratio}\), but similarly low
-\(\mathrm{AA}\) (\autoref{fig:traits-AA}B). Among coastal sites, the
-site with the lowest \(g_\mathrm{smax,ratio}\) had the lowest
-\(\mathrm{AA}\), but there was little variation in
-\(g_\mathrm{smax,ratio}\) among coastal leaves in our sample. We
-therefore cannot rule out that a larger sample of coastal leaves with
-greater variance in \(g_\mathrm{smax,ratio}\) might support this
-hypothesis.
-
-Two major implications from our study are that 1) photosynthesis in
-hypostomatous leaves is likely limited by CO\(_2\) concentration
-drawdown within leaf airspaces; and 2) amphistomy \emph{per se}
-contributes to, but is not wholly responsible for, higher photosynthetic
-rates among amphistomatous leaves. The amphistomy advantage we observe
-in ʻilima leaves implies decreased CO\(_2\) supply in
-pseudohypostomatous leaves because of concentration drawdowns in the
-leaf airspace. Limited diffusion through the airspace has long been
-hypothesized to depress photosynthesis in hypostomatous leaves
-(Parkhurst, 1994), with empirical support from helox studies (Parkhurst
-and Mott, 1990). However, these studies relied on interspecific
-comparisons of amphi- and hypostomatous leaves that differ
-systematically in many traits that affect gas exchange and
-photosynthesis (Xiong and Flexas, 2020). Our experimental approach
-overcomes this limitation and implies that the drop in CO\(_2\)
-concentration from substomatal cavities to the upper surface depresses
-photosynthesis.
-
-Among land plants grown in a common garden, amphistomatous leaves have
-on average nearly \(2\times\) higher area-based photosynthetic rates
-(Xiong and Flexas, 2020), naively implying an
-\(\mathrm{AA} \approx \text{log}~2 = 0.69\). This is much higher than
-our estimate of 0.12 among coastal ʻilima leaves. The most likely
-explanation is that amphistomy is not the only cause of high
-photosynthetic rate. Indeed, species adapted to open, high light
-habitats with amphistomatous leaves also have higher concentrations of
-Rubisco, overall stomatal conductance, and photosynthetic capacity
-(Smith et al., 1997; Xiong and Flexas, 2020). For a leaf with high
-photosynthetic capacity that is well illuminated and hydrated, the major
-limitation becomes CO\(_2\). Under these conditions, amphistomy may
-substantially increase photosynthesis, as we observe in coastal ʻilima
-leaves. Selection on increased photosynthesis under similar conditions
-may explain why crops leaves tend to increase stomatal density ratio
-during domestication (Milla et al., 2013).
-
-Two limitations of this study are the small sample sample sizes and
-experimental design that precludes distinguishing genetic from
-environmental differences in leaf traits. Understanding the mechanistic
-basis of higher \(\mathrm{A}A\) in sun leaves would require much larger
-sample sizes. Sun leaves tend to be thicker, more densely packed with
-mesophyll cells, and have greater photosynthetic capacity and higher
-stomatal conductance, among other traits (Lambers et al., 2008). Each of
-these factors and others potentially modulate \(\mathrm{AA}\).
-Quantifying the contribution of all these factors requires larger
-samples and additional measurements that are beyond the scope of this
-study, but exciting avenues for future research on leaf
-structure-function relations. Although many morphological traits that
-distinguish coastal and montane ʻilima populations persist in a common
-environment (Yorkston and Daehler, 2006), we cannot distinguish between
-genetic effects and plastic responses to habitat as causes of difference
-in \(\mathrm{AA}\) because we measured naturally occurring plants
-\emph{in situ}. While disentangling genetic and plastic contributions is
-not necessarily important for understanding the distribution of
-amphistomatous leaves, it would be useful understand genetic and
-environmental contributions to trait variation. A reciprocal transplant
-would be able to determine the genetic and environmental contributions,
-as well their interaction, to trait variance in nature.
-
-\hypertarget{conclusions}{%
-\section{CONCLUSIONS ---}\label{conclusions}}
-
-This study reports the first direct experimental evidence that having
-stomata open on both leaf surfaces, amphistomy, increases photosynthesis
-for a given total stomatal conductance, particularly in leaves from the
-type of open, sunny habitats where this trait is most common. By
-developing a straightforward gas exchange method to block gas exchange
-through the upper surface, we directly compared the photosynthetic rate
-of a leaf with gas exchange through both surfaces or just one, holding
-all other factors constant. In doing so, we found that coastal leaves of
-the indigenous Hawaiian ʻilima (\emph{Sida fallax}) enjoyed a greater
-photosynthetic benefit from amphistomy than nearby montane leaves living
-in more closed forest. We cannot yet ascribe the difference in
-amphistomy advantage between coastal and montane leaves to particular
-physiological or anatomical variation, but this is a promising area for
-future research.
-
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments ---}\label{acknowledgments}}
-
-The authors thank Kasey Barton for feedback on an earlier version of
-this manuscript, the University of Hawaiʻi honors council for guidance
-to GT, Tawn Speetjens for access to one montane site, startup funds from
-the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
-2307341 to TNB.
-
-\hypertarget{author-ontributions}{%
-\section{Author Contributions ---}\label{author-ontributions}}
-
-GT and CDM contributed equally to all stages of this project; TNB
-contributed to development of the method and helped edit the manuscript.
-
-\hypertarget{data-availability-statement}{%
-\section{Data Availability Statement
----}\label{data-availability-statement}}
-
-Custom scripts are available on a GitHub repository
-(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
-with a DOI and stable URL upon publication. Raw data will be deposited
-on Dryad with a DOI and stable URL upon publication.
-
-\hypertarget{supporting-information}{%
-\section{Supporting Information ---}\label{supporting-information}}
-
-Additional supporting information may be found online in the Supporting
-Information section at the end of the article
-
-\begin{itemize}
-\tightlist
-\item
-  Appendix S1: Supplemental figures and table
-\end{itemize}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\begin{landscape}
+\begin{longtable}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}rrrrrrr}
+\caption{Average traits values among ʻilima leaves at each site.
+\(\mathrm{SD}_\text{abaxial}\) is the stomatal density per mm\(^2\) on
+the abaxial (lower) surface; \(\mathrm{SD}_\text{adaxial}\) is the
+stomatal density per mm\(^2\) on the adaxial (upper) surface;
+\(\mathrm{GCL}_\text{abaxial}\) is the guard cell length in \(\mu\)m on
+the abaxial (lower) surface; \(\mathrm{GCL}_\text{adaxial}\) is the
+guard cell length in \(\mu\)m on the adaxial (upper) surface; Leaf
+thickness is the length from upper cuticle to lower cuticle in \(\mu\)m;
+\(A\) is the photosynthetic rate in
+\(\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}\); \(g_\mathrm{sw}\)
+is the stomatal conductance to water vapor in
+\(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\).}\tabularnewline
+
+\toprule
+Site & Island & Habitat & $\mathrm{SD}_\text{abaxial}$ & $\mathrm{SD}_\text{adaxial}$ & $\mathrm{GCL}_\text{abaxial}$ & $\mathrm{GCL}_\text{adaxial}$ & Leaf thickness & $A$ & $g_\text{sw}$\\
+\midrule
+\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{412.99} & \cellcolor{gray!6}{92.31} & \cellcolor{gray!6}{14.73} & \cellcolor{gray!6}{20.32} & \cellcolor{gray!6}{189.60} & \cellcolor{gray!6}{17.40} & \cellcolor{gray!6}{0.123}\\
+Puakō petroglyph park & Hawaiʻi & coastal & 310.12 & 52.55 & 13.35 & 20.94 & 194.20 & 17.85 & 0.134\\
+\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{420.43} & \cellcolor{gray!6}{80.80} & \cellcolor{gray!6}{16.23} & \cellcolor{gray!6}{22.77} & \cellcolor{gray!6}{304.24} & \cellcolor{gray!6}{20.88} & \cellcolor{gray!6}{0.155}\\
+Kaloko beach & Oʻahu & coastal & 400.27 & 109.03 & 15.69 & 23.26 & 399.54 & 31.28 & 0.314\\
+\cellcolor{gray!6}{Kaʻena Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{370.14} & \cellcolor{gray!6}{120.34} & \cellcolor{gray!6}{17.61} & \cellcolor{gray!6}{22.89} & \cellcolor{gray!6}{295.82} & \cellcolor{gray!6}{34.77} & \cellcolor{gray!6}{0.336}\\
+\addlinespace
+Makapuʻu beach & Oʻahu & coastal & 408.28 & 23.18 & 15.99 & 22.03 & 249.71 & 32.87 & 0.349\\
+\cellcolor{gray!6}{Hāloa ʻĀina} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{307.94} & \cellcolor{gray!6}{7.61} & \cellcolor{gray!6}{16.49} & \cellcolor{gray!6}{22.59} & \cellcolor{gray!6}{149.41} & \cellcolor{gray!6}{13.07} & \cellcolor{gray!6}{0.116}\\
+Kaʻohe game management area & Hawaiʻi & montane & 270.24 & 13.07 & 14.87 & 21.51 & 183.72 & 12.94 & 0.130\\
+\cellcolor{gray!6}{Koaiʻa tree sanctuary} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{318.58} & \cellcolor{gray!6}{9.55} & \cellcolor{gray!6}{13.89} & \cellcolor{gray!6}{22.86} & \cellcolor{gray!6}{138.89} & \cellcolor{gray!6}{27.88} & \cellcolor{gray!6}{0.358}\\
+Hawaiʻi loa ridge & Oʻahu & montane & 329.25 & 115.56 & 17.83 & 21.76 & 205.97 & 21.95 & 0.215\\
+\addlinespace
+\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{298.77} & \cellcolor{gray!6}{138.32} & \cellcolor{gray!6}{16.53} & \cellcolor{gray!6}{20.72} & \cellcolor{gray!6}{162.02} & \cellcolor{gray!6}{24.60} & \cellcolor{gray!6}{0.436}\\
+Waʻahila ridge & Oʻahu & montane & 346.75 & 150.26 & 17.76 & 20.61 & 193.33 & 13.96 & 0.194\\
+\bottomrule
+\end{longtable}
+\end{landscape}
+
+\newpage
+
 \begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
+  \includegraphics{../figures/licor.pdf}
+  \caption{See next page.}
   \label{fig:licor}
@@
 
+\newpage
+
+\setcounter{figure}{\numexpr\value{figure}-1\relax}
+
 \begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
+  \caption{(Contunited from previous page) Individual $A \textendash g_\mathrm{sw}$ curves used to estimate $\mathrm{AA}$.
+For each leaf, one per site, we measured $A$ ($y$-axis) over a range of $g_\mathrm{sw}$ ($x$-axis) on the same leaf with two treatments: `amphi' (circles, solid line) leaves were untreated; `pseudohypo' (squares, dashed line) leaves had no conductance through the upper (adaxial) surface.
+In all coastal (orange) and montane (blue) leaves, we fit generalized additive models and 95\% confidence ribbons to estimate $\mathrm{AA}$ at a $g_\mathrm{sw}$ where the curves overlap (vertical black line).
+In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap.
+$\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+  \label{fig:licor}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+  \includegraphics{../figures/habitat-Ags.pdf}
+  \caption{The photsynthetic rate (left facet) and stomatal conductance to water vapor (right facet) of montane (blue) and coastal (orange) ʻilima leaves.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of trait value for that habitat.
+Smaller points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants, one per site.
+$A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:habitat-Ags}
@@
 
+\newpage
+
 \begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
+  \includegraphics{../figures/habitat-gmaxratio.pdf}
+  \caption{The $g_\mathrm{smax,ratio}$ ($y$-axis) of montane (blue) and coastal (orange) ʻilima leaves.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of $g_\mathrm{smax,ratio}$ for that site.
+Sites are arranged by habitat and ascending $g_\mathrm{smax,ratio}$ within habitat.
+Smaller, transparent points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants.
+$g_\mathrm{smax,ratio}$, the ratio of anatomical maximum stomatal conductance to water vapor on the the adaxial surface over the total.}
   \label{fig:habitat-gmaxratio}
@@
 
-\begin{table}
-  \captionsetup{labelformat=empty}
-  \caption{}
-
-\end{table}
-
-\hypertarget{literature-cited}{%
-\section*{Literature Cited ---}\label{literature-cited}}
-\addcontentsline{toc}{section}{Literature Cited ---}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-noauthor_inaturalist_2022}{}}%
-Anon. 2022. \href{https://www.inaturalist.org}{{iNaturalist}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-biggins_model_1987}{}}%
-Ball, J. T., I. E. Woodrow, and J. A. Berry. 1987.
-\href{https://doi.org/10.1007/978-94-017-0519-6_48}{A {Model}
-{Predicting} {Stomatal} {Conductance} and its {Contribution} to the
-{Control} of {Photosynthesis} under {Different} {Environmental}
-{Conditions}}. \emph{In} J. Biggins {[}ed.{]}, Progress in
-{Photosynthesis} {Research}, 221--224. Springer Netherlands, Dordrecht.
-
-\leavevmode\vadjust pre{\hypertarget{ref-berry_stomata:_2010}{}}%
-Berry, J. A., D. J. Beerling, and P. J. Franks. 2010.
-\href{https://doi.org/10.1016/j.pbi.2010.04.013}{Stomata: Key players in
-the earth system, past and present}. \emph{Current Opinion in Plant
-Biology} 13: 232--239.
-
-\leavevmode\vadjust pre{\hypertarget{ref-de_boer_apparent_2016}{}}%
-Boer, H. J. de, P. L. Drake, E. Wendt, C. A. Price, E.-D. Schulze, N. C.
-Turner, D. Nicolle, and E. J. Veneklaas. 2016.
-\href{https://doi.org/10.1104/pp.16.01313}{Apparent overinvestment in
-leaf venation relaxes leaf morphological constraints on photosynthesis
-in arid habitats}. \emph{Plant Physiology} 172: 2286--2299.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bucher_stomatal_2017}{}}%
-Bucher, S. F., K. Auerswald, C. Grün-Wenzel, S. I. Higgins, J. Garcia
-Jorge, and C. Römermann. 2017.
-\href{https://doi.org/10.1016/j.flora.2017.02.011}{Stomatal traits
-relate to habitat preferences of herbaceous species in a temperate
-climate}. \emph{Flora} 229: 107--115.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burkner_brms_2017}{}}%
-Bürkner, P.-C. 2017.
-\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
-\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
-\emph{stan}}. \emph{Journal of Statistical Software} 80.
-
-\leavevmode\vadjust pre{\hypertarget{ref-carlquist_biota_1966}{}}%
-Carlquist, S. 1966. \href{https://doi.org/10.2307/2406147}{The {Biota}
-of {Long}-{Distance} {Dispersal}. {II}. {Loss} of {Dispersibility} in
-{Pacific} {Compositae}}. \emph{Evolution} 20: 30.
-
-\leavevmode\vadjust pre{\hypertarget{ref-collatz_physiological_1991}{}}%
-Collatz, G. J., J. T. Ball, C. Grivet, and J. A. Berry. 1991.
-\href{https://doi.org/10.1016/0168-1923(91)90002-8}{Physiological and
-environmental regulation of stomatal conductance, photosynthesis and
-transpiration: A model that includes a laminar boundary layer}.
-\emph{Agricultural and Forest Meteorology} 54: 107--136.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cowan_stomatal_1977}{}}%
-Cowan, I. R., and G. D. Farquhar. 1977. Stomatal function in relation to
-leaf metabolism and environment. \emph{Symposia of the Society for
-Experimental Biology} 31: 471--505.
-
-\leavevmode\vadjust pre{\hypertarget{ref-deans_optimization_2020}{}}%
-Deans, R. M., T. J. Brodribb, F. A. Busch, and G. D. Farquhar. 2020.
-\href{https://doi.org/10.1038/s41477-020-00760-6}{Optimization can
-provide the fundamental link between leaf photosynthesis, gas exchange
-and water relations}. \emph{Nature Plants} 6: 1116--1125.
-
-\leavevmode\vadjust pre{\hypertarget{ref-doll_callitriche_2021}{}}%
-Doll, Y., H. Koga, and H. Tsukaya. 2021.
-\href{https://doi.org/10.1080/15592324.2021.1978201}{\emph{Callitriche}
-as a potential model system for evolutionary studies on the dorsiventral
-distribution of stomata}. \emph{Plant Signaling \& Behavior} 16:
-1978201.
-
-\leavevmode\vadjust pre{\hypertarget{ref-drake_two_2019}{}}%
-Drake, P. L., H. J. de Boer, S. J. Schymanski, and E. J. Veneklaas.
-2019. \href{https://doi.org/10.1111/nph.15652}{Two sides to every leaf:
-Water and {CO}\(_{\textrm{2}}\) transport in hypostomatous and
-amphistomatous leaves}. \emph{New Phytologist} 222: 1179--1187.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dunbar-co_leaf_2009}{}}%
-Dunbar-Co, S., M. J. Sporck, and L. Sack. 2009. Leaf trait
-diversification and design in seven rare taxa of the {Hawaiian}
-\emph{p}lantago radiation. \emph{International Journal of Plant
-Sciences} 170: 61--75.
-
-\leavevmode\vadjust pre{\hypertarget{ref-earles_beyond_2018}{}}%
-Earles, J. M., G. Theroux-Rancourt, A. B. Roddy, M. E. Gilbert, A. J.
-McElrone, and C. R. Brodersen. 2018.
-\href{https://doi.org/10.1104/pp.18.00550}{Beyond porosity: {3D} leaf
-intercellular airspace traits that impact mesophyll conductance}.
-\emph{Plant Physiology} 178: 148--162.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ehleringer_pubescence_1978}{}}%
-Ehleringer, J. R., and O. Björkman. 1978.
-\href{https://doi.org/10.1007/BF00349805}{Pubescence and leaf spectral
-characteristics in a desert shrub, {Encelia} farinosa}. \emph{Oecologia}
-36: 151--162.
-
-\leavevmode\vadjust pre{\hypertarget{ref-evans_resistances_2009}{}}%
-Evans, J. R., R. Kaldenhoff, B. Genty, and I. Terashima. 2009.
-\href{https://doi.org/10.1093/jxb/erp117}{Resistances along the
-{CO}\(_{\textrm{2}}\) diffusion pathway inside leaves}. \emph{Journal of
-Experimental Botany} 60: 2235--2248.
-
-\leavevmode\vadjust pre{\hypertarget{ref-foster_influence_1986}{}}%
-Foster, J. R., and W. K. Smith. 1986. Influence of stomatal distribution
-on transpiration in low-wind environments. \emph{Plant, Cell \&
-Environment} 9: 751--759.
-
-\leavevmode\vadjust pre{\hypertarget{ref-franklin_organizing_2020}{}}%
-Franklin, O., S. P. Harrison, R. Dewar, C. E. Farrior, Å. Brännström, U.
-Dieckmann, S. Pietsch, et al. 2020.
-\href{https://doi.org/10.1038/s41477-020-0655-x}{Organizing principles
-for vegetation dynamics}. \emph{Nature Plants} 6: 444--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-franks_comparing_2018}{}}%
-Franks, P. J., G. B. Bonan, J. A. Berry, D. L. Lombardozzi, N. M.
-Holbrook, N. Herold, and K. W. Oleson. 2018.
-\href{https://doi.org/10.1111/gcb.14445}{Comparing optimal and empirical
-stomatal conductance models for application in {Earth} system models}.
-\emph{Global Change Biology} 24: 5708--5723.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
-Gabry, J., and R. Češnovar. 2023.
-\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
-{R} {Interface} to '{CmdStan}'}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gay_influence_1975}{}}%
-Gay, A. P., and R. G. Hurd. 1975.
-\href{https://doi.org/10.1111/j.1469-8137.1975.tb01368.x}{The influence
-of light on stomatal density in the tomato}. \emph{New Phytologist} 75:
-37--46.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibson_structure-function_1996}{}}%
-Gibson, A. C. 1996.
-\href{http://public.eblib.com/choice/PublicFullRecord.aspx?p=6495247}{Structure-{Function}
-{Relations} of {Warm} {Desert} {Plants}}. Springer Berlin / Heidelberg,
-Berlin, Heidelberg.
-
-\leavevmode\vadjust pre{\hypertarget{ref-givnish_economy_1986}{}}%
-Givnish, T. J. ed. 1986. On the economy of plant form and function.
-Cambridge University Press, Cambridge.
-
-\leavevmode\vadjust pre{\hypertarget{ref-givnish_origin_2009}{}}%
-Givnish, T. J., K. C. Millam, A. R. Mast, T. B. Paterson, T. J. Theim,
-A. L. Hipp, J. M. Henss, et al. 2009.
-\href{https://doi.org/10.1098/rspb.2008.1204}{Origin, adaptive radiation
-and diversification of the {Hawaiian} lobeliads ({Asterales}:
-{Campanulaceae})}. \emph{Proceedings of the Royal Society B: Biological
-Sciences} 276: 407--416.
-
-\leavevmode\vadjust pre{\hypertarget{ref-givnish_common-garden_2014}{}}%
-Givnish, T. J., and R. A. Montgomery. 2014.
-\href{https://doi.org/10.1098/rspb.2013.2944}{Common-garden studies on
-adaptive radiation of photosynthetic physiology among {Hawaiian}
-lobeliads}. \emph{Proceedings of the Royal Society B: Biological
-Sciences} 281: 20132944--20132944.
-
-\leavevmode\vadjust pre{\hypertarget{ref-givnish_adaptive_2004}{}}%
-Givnish, T. J., R. A. Montgomery, and G. Goldstein. 2004.
-\href{https://doi.org/10.3732/ajb.91.2.228}{Adaptive radiation of
-photosynthetic physiology in the {Hawaiian} lobeliads: Light regimes,
-static light responses, and whole-plant compensation points}.
-\emph{American Journal of Botany} 91: 228--246.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grubb_leaf_2020}{}}%
-Grubb, P. J. 2020. Leaf structure and function. \emph{In} A. Dobson, D.
-Tilman, and R. D. Holt {[}eds.{]}, Unsolved {Problems} in {Ecology},
-124--144. Princeton University Press, Princeton.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grubb_leaf_1977}{}}%
-Grubb, P. J. 1977. Leaf structure and function. \emph{In} R. Duncan, and
-M. Weston-Smith {[}eds.{]}, The encyclopedia of ignorance, 317--330.
-Pergamon, Oxford.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gutschick_photosynthesis_1984}{}}%
-Gutschick, V. P. 1984a. Photosynthesis model for {C}\(_{\textrm{3}}\)
-leaves incorporating {CO}\(_{\textrm{2}}\) transport, propagation of
-radiation, and biochemistry 1. Kinetics and their parameterization.
-\emph{Photosynthetica} 18: 549--568.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gutschick_photosynthesis_1984-1}{}}%
-Gutschick, V. P. 1984b. Photosynthesis model for {C}\(_{\textrm{3}}\)
-leaves incorporating {CO}\(_{\textrm{2}}\) transport, propagation of
-radiation, and biochemistry 2. Ecological and agricultural utility.
-\emph{Photosynthetica} 18: 569--595.
-
-\leavevmode\vadjust pre{\hypertarget{ref-harrison_ecoevolutionary_2021}{}}%
-Harrison, S. P., W. Cramer, O. Franklin, I. C. Prentice, H. Wang, Å.
-Brännström, H. Boer, et al. 2021.
-\href{https://doi.org/10.1111/nph.17558}{Eco‐evolutionary optimality as
-a means to improve vegetation and land‐surface models}. \emph{New
-Phytologist} 231: 2125--2141.
-
-\leavevmode\vadjust pre{\hypertarget{ref-harwood_understanding_2021}{}}%
-Harwood, R., G. Théroux‐Rancourt, and M. M. Barbour. 2021.
-\href{https://doi.org/10.1111/pce.14079}{Understanding airspace in
-leaves: {3D} anatomy and directional tortuosity}. \emph{Plant, Cell \&
-Environment} 44: 2455--2465.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hetherington_role_2003}{}}%
-Hetherington, A. M., and F. I. Woodward. 2003.
-\href{https://doi.org/10.1038/nature01843}{The role of stomata in
-sensing and driving environmental change}. \emph{Nature} 424: 901--908.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hill_temperature_2014}{}}%
-Hill, K. E., G. R. Guerin, R. S. Hill, and J. R. Watling. 2014.
-\href{https://doi.org/10.1071/BT14204}{Temperature influences stomatal
-density and maximum potential water loss through stomata of {Dodonaea}
-viscosa subsp. Angustissima along a latitude gradient in southern
-{Australia}}. \emph{Australian Journal of Botany} 62: 657.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ho_three-dimensional_2016}{}}%
-Ho, Q. T., H. N. C. Berghuijs, R. Watté, P. Verboven, E. Herremans, X.
-Yin, M. A. Retta, et al. 2016.
-\href{https://doi.org/10.1111/pce.12590}{Three-dimensional microscale
-modelling of {CO}\(_{\textrm{2}}\) transport and light propagation in
-tomato leaves enlightens photosynthesis: 3-{D} modelling of
-photosynthesis in leaves}. \emph{Plant, Cell \& Environment} 39: 50--61.
-
-\leavevmode\vadjust pre{\hypertarget{ref-james_leaf_2001}{}}%
-James, S. A., and D. T. Bell. 2001.
-\href{https://doi.org/10.1071/BT99044}{Leaf morphological and anatomical
-characteristics of heteroblastic {Eucalyptus} globulus ssp. Globulus
-({Myrtaceae})}. \emph{Australian Journal of Botany} 49: 259.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jarvis_interpretation_1976}{}}%
-Jarvis, P. G. 1976. \href{https://doi.org/10.1098/rstb.1976.0035}{The
-interpretation of the variations in leaf water potential and stomatal
-conductance found in canopies in the field}. \emph{Philosophical
-Transactions of the Royal Society of London. B, Biological Sciences}
-273: 593--610.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jordan_using_2014}{}}%
-Jordan, G. J., R. J. Carpenter, and T. J. Brodribb. 2014.
-\href{https://doi.org/10.1016/j.palaeo.2013.12.035}{Using fossil leaves
-as evidence for open vegetation}. \emph{Palaeogeography,
-Palaeoclimatology, Palaeoecology} 395: 168--175.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kaul_anatomical_1976}{}}%
-Kaul, R. B. 1976.
-\href{https://doi.org/10.1016/0304-3770(76)90022-X}{Anatomical
-observations on floating leaves}. \emph{Aquatic Botany} 2: 215--234.
-
-\leavevmode\vadjust pre{\hypertarget{ref-knope_dispersal_2020}{}}%
-Knope, M. L., V. A. Funk, M. A. Johnson, W. L. Wagner, E. M. Datlof, G.
-Johnson, D. J. Crawford, et al. 2020.
-\href{https://doi.org/10.1111/jse.12704}{Dispersal and adaptive
-radiation of \emph{bidens} ({Compositae}) across the remote
-archipelagoes of {Polynesia}}. \emph{Journal of Systematics and
-Evolution} 58: 805--822.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lambers_plant_2008}{}}%
-Lambers, H., F. S. Chapin, and T. L. Pons. 2008.
-\href{https://doi.org/10.1007/978-0-387-78341-3}{Plant {Physiological}
-{Ecology}}. Springer New York, New York, NY.
-
-\leavevmode\vadjust pre{\hypertarget{ref-leuning_critical_1995}{}}%
-Leuning, R. 1995.
-\href{https://doi.org/10.1111/j.1365-3040.1995.tb00370.x}{A critical
-appraisal of a combined stomatal-photosynthesis model for {C3} plants}.
-\emph{Plant, Cell and Environment} 18: 339--355.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lin_optimal_2015}{}}%
-Lin, Y.-S., B. E. Medlyn, R. A. Duursma, I. C. Prentice, H. Wang, S.
-Baig, D. Eamus, et al. 2015.
-\href{https://doi.org/10.1038/nclimate2550}{Optimal stomatal behaviour
-around the world}. \emph{Nature Climate Change} 5: 459--464.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lyshede_comparative_2002}{}}%
-Lyshede, O. B. 2002. Comparative and functional leaf anatomy of selected
-{Alstroemeriaceae} of mainly {Chilean} origin. \emph{Botanical Journal
-of the Linnean Society} 140: 261--272.
-
-\leavevmode\vadjust pre{\hypertarget{ref-marquez_assessing_2023}{}}%
-Márquez, D. A., H. Stuart‐Williams, L. A. Cernusak, and G. D. Farquhar.
-2023. \href{https://doi.org/10.1111/nph.18784}{Assessing the
-{\textless{}}span style="font-variant:small-caps;"{\textgreater{}} {CO}
-\(_{\textrm{2}}\) {\textless{}}/span{\textgreater{}} concentration at
-the surface of photosynthetic mesophyll cells}. \emph{New Phytologist}
-238: 1446--1460.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mckown_trait_2016}{}}%
-McKown, A. D., M. E. Akamine, and L. Sack. 2016.
-\href{https://doi.org/10.1007/s00442-016-3640-3}{Trait convergence and
-diversification arising from a complex evolutionary history in
-{Hawaiian} species of {Scaevola}}. \emph{Oecologia} 181: 1083--1100.
-
-\leavevmode\vadjust pre{\hypertarget{ref-medlyn_reconciling_2011}{}}%
-Medlyn, B. E., R. A. Duursma, D. Eamus, D. S. Ellsworth, I. C. Prentice,
-C. V. M. Barton, K. Y. Crous, et al. 2011.
-\href{https://doi.org/10.1111/j.1365-2486.2010.02375.x}{Reconciling the
-optimal and empirical approaches to modelling stomatal conductance}.
-\emph{Global Change Biology} 17: 2134--2144.
-
-\leavevmode\vadjust pre{\hypertarget{ref-milla_shifts_2013}{}}%
-Milla, R., N. de Diego-Vico, and N. Martín-Robles. 2013.
-\href{https://doi.org/10.1093/jxb/ert147}{Shifts in stomatal traits
-following the domestication of plant species}. \emph{Journal of
-Experimental Botany} 64: 3137--3146.
-
-\leavevmode\vadjust pre{\hypertarget{ref-montgomery_adaptive_2008}{}}%
-Montgomery, R. A., and T. J. Givnish. 2008.
-\href{https://doi.org/10.1007/s00442-007-0936-3}{Adaptive radiation of
-photosynthetic physiology in the {Hawaiian} lobeliads: Dynamic
-photosynthetic responses}. \emph{Oecologia} 155: 455--467.
-
-\leavevmode\vadjust pre{\hypertarget{ref-morison_lateral_2005}{}}%
-Morison, J. I. L., E. Gallouët, T. Lawson, G. Cornic, R. Herbin, and N.
-R. Baker. 2005. \href{https://doi.org/10.1104/pp.105.062950}{Lateral
-diffusion of {CO}\(_{\textrm{2}}\) in leaves is not sufficient to
-support photosynthesis}. \emph{Plant Physiology} 139: 254--266.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mott_adaptive_1982}{}}%
-Mott, K. A., A. C. Gibson, and J. W. O'Leary. 1982.
-\href{https://doi.org/10.1111/1365-3040.ep11611750}{The adaptive
-significance of amphistomatic leaves}. \emph{Plant, Cell \& Environment}
-5: 455--460.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mott_amphistomy_1991}{}}%
-Mott, K. A., and O. Michaelson. 1991. Amphistomy as an adaptation to
-high light intensity in \emph{{Ambrosia} cordifolia} ({Compositae}).
-\emph{American Journal of Botany} 78: 76--79.
-
-\leavevmode\vadjust pre{\hypertarget{ref-muir_is_2019}{}}%
-Muir, C. D. 2019. \href{https://doi.org/10.1093/icb/icz085}{Is
-{Amphistomy} an {Adaptation} to {High} {Light}? {Optimality} {Models} of
-{Stomatal} {Traits} along {Light} {Gradients}}. \emph{Integrative and
-Comparative Biology} 59: 571--584.
-
-\leavevmode\vadjust pre{\hypertarget{ref-muir_light_2018}{}}%
-Muir, C. D. 2018. Light and growth form interact to shape stomatal ratio
-among {British} angiosperms. \emph{New Phytologist} 218: 242--252.
-
-\leavevmode\vadjust pre{\hypertarget{ref-muir_making_2015}{}}%
-Muir, C. D. 2015. \href{https://doi.org/10.1098/rspb.2015.1498}{Making
-pore choices: Repeated regime shifts in stomatal ratio}.
-\emph{Proceedings of the Royal Society B: Biological Sciences} 282:
-20151498.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nobel_physicochemical_2020}{}}%
-Nobel, P. S. 2020. Physicochemical and environmental plant physiology.
-Fifth edition. Academic Press, an imprint of Elsevier, London, United
-Kingdom ; San Diego, CA.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_diffusion_1994}{}}%
-Parkhurst, D. F. 1994.
-\href{http://www.jstor.org/stable/2557929}{Diffusion of
-{CO}\(_{\textrm{2}}\) and other gases inside leaves}. \emph{New
-Phytologist} 126: 449--479.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_adaptive_1978}{}}%
-Parkhurst, D. F. 1978. \href{https://doi.org/10.2307/2259142}{The
-adaptive significance of stomatal occurrence on one or both surfaces of
-leaves}. \emph{The Journal of Ecology} 66: 367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_intercellular_1990}{}}%
-Parkhurst, D. F., and K. A. Mott. 1990.
-\href{https://doi.org/10.1104/pp.94.3.1024}{Intercellular diffusion
-limits to {CO}\(_{\textrm{2}}\) uptake in leaves: Studies in air and
-helox}. \emph{Plant Physiology} 94: 1024--1032.
-
-\leavevmode\vadjust pre{\hypertarget{ref-peat_comparative_1994}{}}%
-Peat, H. J., and A. H. Fitter. 1994. A comparative study of the
-distribution and density of stomata in the {British} flora.
-\emph{Biological Journal of the Linnean Society} 52: 377--393.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_phylogenetic_2022}{}}%
-Pejhanmehr, M. 2022. Phylogenetic affinities among
-{\textless{}}i{\textgreater{}}{Sida}{\textless{}}{\textbackslash{}}i{\textgreater{}}
-species and allied genera ({Malvaceae}: {Malveae}), and examination of
-{\textless{}}i{\textgreater{}}{Sida}
-fallax{\textless{}}{\textbackslash{}}i{\textgreater{}} within the
-{Hawaiian} {Islands} and throughout the {Pacific}. PhD thesis.
-University of Hawai'i at Mānoa.
-
-\leavevmode\vadjust pre{\hypertarget{ref-poorter_metaanalysis_2019}{}}%
-Poorter, H., Ü. Niinemets, N. Ntagkas, A. Siebenkäs, M. Mäenpää, S.
-Matsubara, and T. L. Pons. 2019.
-\href{https://doi.org/10.1111/nph.15754}{A meta‐analysis of plant
-responses to light intensity for 70 traits ranging from molecules to
-whole plant performance}. \emph{New Phytologist}: nph.15754.
-
-\leavevmode\vadjust pre{\hypertarget{ref-r_core_team_r:_2023}{}}%
-R Core Team. 2023. \href{http://www.R-project.org/}{R: {A} {Language}
-and {Environment} for {Statistical} {Computing}}. R Foundation for
-Statistical Computing, Vienna, Austria.
-
-\leavevmode\vadjust pre{\hypertarget{ref-robichaux_evolution_1984}{}}%
-Robichaux, R. H., and R. W. Pearcy. 1984.
-\href{https://doi.org/10.1002/j.1537-2197.1984.tb12492.x}{Evolution of
-{C}\(_{\textrm{3}}\) and {C}\(_{\textrm{4}}\) plants along an
-environmental moisture gradient: Patterns of photosynthetic
-differentiation in {Hawaiian}
-{\textless{}}i{\textgreater{}}{Scaevola}{\textless{}}{\textbackslash{}}i{\textgreater{}}
-and
-{\textless{}}i{\textgreater{}}{Euphorbia}{\textless{}}{\textbackslash{}}i{\textgreater{}}
-species}. \emph{American Journal of Botany} 71: 121--129.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sack_developmental_2016}{}}%
-Sack, L., and T. N. Buckley. 2016.
-\href{https://doi.org/10.1104/pp.16.00476}{The developmental basis of
-stomatal density and flux}. \emph{Plant Physiology} 171: 2358--2363.
-
-\leavevmode\vadjust pre{\hypertarget{ref-salisbury_i_1928}{}}%
-Salisbury, E. J. 1928. \href{https://doi.org/10.1098/rstb.1928.0001}{I.
-{On} the causes and ecological significance of stomatal frequency, with
-special reference to the woodland flora}. \emph{Philosophical
-Transactions of the Royal Society of London. Series B, Containing Papers
-of a Biological Character} 216: 1--65.
-
-\leavevmode\vadjust pre{\hypertarget{ref-santrucek_partitioning_2019}{}}%
-Šantrůček, J., L. Schreiber, J. Macková, M. Vráblová, J. Květoň, P.
-Macek, and J. Neuwirthová. 2019. Partitioning of mesophyll conductance
-for {CO}\(_{\textrm{2}}\) into intercellular and cellular components
-using carbon isotope composition of cuticles from opposite leaf sides.
-\emph{Photosynthesis Research}: 1--19.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schneider_nih_2012}{}}%
-Schneider, C. A., W. S. Rasband, and K. W. Eliceiri. 2012.
-\href{https://doi.org/10.1038/nmeth.2089}{{NIH} {Image} to {ImageJ}: 25
-years of image analysis}. \emph{Nature Methods} 9: 671--675.
-
-\leavevmode\vadjust pre{\hypertarget{ref-scoffoni_light-induced_2015}{}}%
-Scoffoni, C., J. Kunkle, J. Pasquet-Kok, C. Vuong, A. J. Patel, R. A.
-Montgomery, T. J. Givnish, and L. Sack. 2015.
-\href{https://doi.org/10.1111/nph.13346}{Light-induced plasticity in
-leaf hydraulics, venation, anatomy, and gas exchange in ecologically
-diverse {Hawaiian} lobeliads}. \emph{New Phytologist} 207: 43--58.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sellers_modeling_1997}{}}%
-Sellers, P. J., R. E. Dickinson, D. A. Randall, A. K. Betts, F. G. Hall,
-J. A. Berry, G. J. Collatz, et al. 1997.
-\href{https://doi.org/10.1126/science.275.5299.502}{Modeling the
-exchanges of energy, water, and carbon between continents and the
-atmosphere}. \emph{Science} 275: 502--509.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith_leaf_1997}{}}%
-Smith, W. K., T. C. Vogelmann, E. H. DeLucia, D. T. Bell, and K. A.
-Shepherd. 1997. Leaf form and photosynthesis. \emph{BioScience} 11:
-785--793.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sporck_hawaiian_2011}{}}%
-Sporck, M. J. 2011. The {Hawaiian} {C}\(_{\textrm{4}}\) \emph{euphorbia}
-adaptive radiation: An ecophysiological approach to understanding leaf
-trait variation. Ph.\{D\}. University of Hawaii.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stan_development_team_stan_2023}{}}%
-Stan Development Team. 2023. \href{https://mc-stan.org}{Stan {Modeling}
-{Language} {Users} {Guide} and {Reference} {Manual}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-theroux-rancourt_maximum_2021}{}}%
-Théroux-Rancourt, G., A. B. Roddy, J. M. Earles, M. E. Gilbert, M. A.
-Zwieniecki, C. K. Boyce, D. Tholen, et al. 2021.
-\href{https://doi.org/10.1098/rspb.2020.3145}{Maximum
-{CO}\(_{\textrm{2}}\) diffusion inside leaves is limited by the scaling
-of cell size and genome size}. \emph{Proceedings of the Royal Society B:
-Biological Sciences} 288: 20203145.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tholen_mechanistic_2011}{}}%
-Tholen, D., and X.-G. Zhu. 2011.
-\href{https://doi.org/10.1104/pp.111.172346}{The mechanistic basis of
-internal conductance: {A} theoretical analysis of mesophyll cell
-photosynthesis and {CO}\(_{\textrm{2}}\) diffusion}. \emph{Plant
-Physiology} 156: 90--105.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tomas_importance_2013}{}}%
-Tomás, M., J. Flexas, L. Copolovici, J. Galmés, L. Hallik, H. Medrano,
-M. Ribas-Carbó, et al. 2013.
-\href{https://doi.org/10.1093/jxb/ert086}{Importance of leaf anatomy in
-determining mesophyll diffusion conductance to {CO2} across species:
-Quantitative limitations and scaling up by models}. \emph{Journal of
-Experimental Botany} 64: 2269--2281.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang_towards_2017}{}}%
-Wang, H., I. C. Prentice, T. F. Keenan, T. W. Davis, I. J. Wright, W. K.
-Cornwell, B. J. Evans, and C. Peng. 2017.
-\href{https://doi.org/10.1038/s41477-017-0006-8}{Towards a universal
-model for carbon dioxide uptake by plants}. \emph{Nature Plants} 3:
-734--741.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang_theoretical_2020}{}}%
-Wang, Y., J. S. Sperry, W. R. L. Anderegg, M. D. Venturas, and A. T.
-Trugman. 2020. \href{https://doi.org/10.1111/nph.16572}{A theoretical
-and empirical assessment of stomatal optimization modeling}. \emph{New
-Phytologist} 227: 311--325.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood_physiology_1934}{}}%
-Wood, J. G. 1934. The physiology of xerophytism in {Australian} plants:
-The stomatal frequencies, transpiration and osmotic pressures of
-sclerophyll and tomentose-succulent leaved plants. \emph{Journal of
-Ecology} 22: 69--87.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xiong_one_2020}{}}%
-Xiong, D., and J. Flexas. 2020.
-\href{https://doi.org/10.1111/nph.16801}{From one side to two sides: The
-effects of stomatal distribution on photosynthesis}. \emph{New
-Phytologist} 228: 1754--1766.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yorkston_interfertility_2006}{}}%
-Yorkston, M., and C. C. Daehler. 2006.
-\href{https://doi.org/10.1086/499615}{Interfertility between {Hawaiian}
-ecotypes of \emph{{Sida} fallax} ({Malvaceae}) and evidence of a hybrid
-disadvantage}. \emph{International Journal of Plant Sciences} 167:
-221--230.
-
-\end{CSLReferences}
-
 \end{document}

### 2023-09-06 r/02_process-trait-data.R

@@ -1,0 +1,144 @@
+# Join data sheets for analysis
+source("r/header.R")
+
+# Climate rasters
+# Extract climate data
+dir = "raw-data/climate"
+tair_ann = raster::raster(glue("{dir}/Tair_month_raster/tair_ann"))
+sl_mst_ann = raster::raster(glue("{dir}/SoilMoisture_month_raster/sl_mst_ann"))
+
+# Read in population information ----
+write_csv(site, "raw-data/site.csv")
+write_csv(plant, "raw-data/plant.csv")
+write_csv(licor_leaf, "raw-data/licor_leaf.csv")
+write_csv(stomata_density, "raw-data/stomata_density.csv")
+write_csv(stomata_size, "raw-data/stomata_size.csv")
+write_csv(leaf_thickness, "raw-data/leaf_thickness.csv")
+
+site = read_csv("raw-data/site.csv", col_types = "ccccccDdddd")
+
+plant = read_csv("raw-data/plant.csv", col_types = "ccdd") |>
+  # Individual with no data because it was too low quality
+  filter(!(site_code == "pkpp" & individual == "i6"))
+
+licor_leaf = read_csv("raw-data/licor_leaf.csv", col_types = "ccc") |>
+  mutate(licor_leaf = TRUE)
+
+stomata_density = read_csv("raw-data/stomata_density.csv", col_types = "ccccdc") |>
+  mutate(stomatal_density_mm2 = number_of_stomata / image_area_mm2)
+
+stomata_size = read_csv("raw-data/stomata_size.csv", col_types = "ccccidd") |>
+  mutate(length_um = length_pixels / pixels_per_um)
+
+leaf_thickness = read_csv("raw-data/leaf_thickness.csv", col_types = "cccddd")
+
+## Check site ----
+assert_character(site$site, any.missing = FALSE)
+assert_character(site$label, any.missing = FALSE)
+assert_character(site$site_code, pattern = "^[a-z]{4}$", any.missing = FALSE)
+assert_character(site$alias)
+assert_true(all(site$site_type %in% c("coastal", "montane")))
+assert_true(all(site$island %in% c("hawaii", "oahu")))
+assert_date(site$date_sampled, lower = "2022-08-07", upper = "2022-11-20")
+assert_numeric(site$latitude_degree, lower = 18, upper = 23,
+               any.missing = FALSE)
+assert_numeric(site$longitude_degree, lower = -160, upper = -154,
+               any.missing = FALSE)
[…]

### 2023-09-06 r/03_process-licor.R

@@ -1,0 +1,110 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+
+# Changing K has some effect on gsw. With Solanum data I compared results
+# for amphi leaves when K = 0.5 or K is estimated iteratively from gas exchange
+# data. The difference was trivial, so I am assuming K = 0.5 for amphi leaves
+
+# Use measured Energy Balance for g_sw calculation? If not, recalc_licor() uses
+# T_leaf measured with the thermocouple
+use_EB = FALSE 
+
+# 2022-08-07: Makapuʻu Beach Park ----
+mkpb = photosynthesis::read_licor("raw-data/licor/2022-08-07-0716_logdata") |>
+  as_tibble(.name_repair = "unique") |>
+  filter(!is.na(gsw), !is.na(A)) |>
+  mutate(aperture = rep(c("two-sided", "one-sided"), c(18, 10))) |>
+  process_licor(S = c(6, 6), use_EB = use_EB, site_code = "mkpb")
+
+# 2022-08-14: Hawaiʻi loa ridge ----
+hwlr = read_licor("raw-data/licor/2022-08-14-0802_logdata") |>
+  as_tibble(.name_repair = "unique") |>
+  filter(!is.na(gsw), !is.na(A), gsw > 0) |>
+  mutate(aperture = rep(c("one-sided", "two-sided"), c(5, 5))) |>
+  process_licor(S = c(6, 6), use_EB = use_EB, site_code = "hwlr")
+
+# 2022-08-19: Kaʻohe game management area ----
+kgma = read_licor("raw-data/licor/2022-08-19-0745_logdata") |>
+  as_tibble(.name_repair = "unique") |>
+  filter(!is.na(gsw), !is.na(A), gsw > 0) |>
+  mutate(aperture = rep(c("two-sided", "one-sided"), c(10, 6))) |>
+  process_licor(S = c(6, 6), use_EB = use_EB, site_code = "kgma")
+
+# 2022-08-19: Puakō petroglyph park ----
+pkpp = read_licor("raw-data/licor/2022-08-19-1038_logdata") |>
+  as_tibble(.name_repair = "unique") |>
+  filter(!is.na(gsw), !is.na(A), gsw > 0) |>
+  mutate(aperture = rep(c("one-sided", "two-sided"), c(14, 13))) |>
[…]

### 2023-09-06 r/04_fit-licor.R

@@ -1,0 +1,29 @@
+source("r/header.R")
+
+licor = read_rds("processed-data/licor.rds")
+
+# fit_licor = brm(bf(
+#   A ~ site_code_aperture + s(gsw) + s(gsw, by = site_code_aperture)),
+#   data = licor,
+#   chains = 4L,
+#   cores = 4L,
+#   seed = 744874146,
+#   backend = "cmdstanr",
+#   control = list(max_treedepth = 11),
+#   silent = 0
+# )
+
+fit_licor = brm(bf(
+  A ~ log(gsw) * site_code_aperture, sigma ~ 1 + (1|site_code_aperture)),
+  data = licor,
+  iter = 4e3,
+  thin = 2e0,
+  chains = 4L,
+  cores = 4L,
+  seed = 744874146,
+  backend = "cmdstanr",
+  control = list(max_treedepth = 12),
+  silent = 0
+)
+
+write_rds(fit_licor, "objects/fit_licor.rds")

### 2023-09-06 r/05_fit-traits.R

@@ -1,0 +1,57 @@
+source("r/header.R")
+
+trait = read_rds("processed-data/trait.rds") |>
+  unite(ind, site_code, individual, remove = FALSE) |>
+  unite(leaf, site_code, individual, leaf_number, remove = FALSE)
+
+# Fit all traits except leaf thickness ----
+# Need separate data for leaf thickness because no leaf replication
+# 
+# There were too many missing values for leaf thickness because it was measured
+# on a small subset of leaves. Therefore, the model did not converge well when
+# it was included among all traits.
+bform =
+  bf(lower_number_of_stomata ~ site_type + (1|p|leaf) + (1|q|ind) +
+       (1|r|site_code)) + negbinomial() +
+  bf(upper_number_of_stomata ~ site_type + (1|p|leaf) + (1|q|ind) +
+       (1|r|site_code)) + negbinomial() +
+  bf(lower_length_um ~ site_type + (1|p|leaf) + (1|q|ind) + (1|r|site_code)) +
+  bf(upper_length_um | mi() ~ site_type + (1|p|leaf) + (1|q|ind) + (1|r|site_code)) +
+  set_rescor(FALSE)
+
+# make_stancode(bform, data = trait)
+
+fit_stomata = brm(
+  bform,
+  data = trait,
+  iter = 16e3,
+  thin = 8e0,
+  chains = 4L,
+  cores = 4L,
+  seed = 912691191,
+  backend = "cmdstanr",
+  control = list(max_treedepth = 12)
+)
+
+write_rds(fit_stomata, "objects/fit_stomata.rds")
+
+# Fit leaf thickness ----
+# Center-scale leaf_thickness_um
+trait1 = filter(trait, !is.na(leaf_thickness_um)) |>
+  group_by(site_type, site_code, ind) |>
+  # There is only 1 average leaf_thickness_um per ind
+  summarize(leaf_thickness_um = first(leaf_thickness_um))
+
+fit_thickness = brm(
+  log(leaf_thickness_um) ~ site_type + (1|ind) + (1|site_code),
+  data = trait1,
+  iter = 2e3,
+  thin = 1e0,
+  chains = 4L,
+  cores = 4L,
+  seed = 995134466,
+  backend = "cmdstanr",
+  control = list(max_treedepth = 10)
+)
+
+write_rds(fit_thickness, "objects/fit_thickness.rds")

### 2023-09-06 r/06_estimate-aa.R

@@ -1,0 +1,50 @@
+source("r/header.R")
+
+licor = read_rds("processed-data/licor.rds")
+fit_licor =  read_rds("objects/fit_licor.rds")
+
+# Estimate AA and CIs for each site ----
+licor2 = licor |>
+  summarize(min = min(gsw), max = max(gsw), .by = c("site_code", "aperture")) |>
+  pivot_wider(names_from = aperture, values_from = min:max) |>
+  rowwise() |>
+  mutate(gsw = ifelse(
+    `max_two-sided` > `max_one-sided`,
+    mean(c(`max_one-sided`, `min_two-sided`)),
+    mean(c(`min_one-sided`, `max_two-sided`))
+  )) |>
+  crossing(aperture = c("one-sided", "two-sided")) |>
+  unite(site_code_aperture, site_code, aperture, remove = FALSE) |>
+  select(-matches("^(max|min)_(one|two)-sided$")) |>
+  mutate(i = row_number())
+
+# fit_licor_draws = prepare_predictions(fit_licor)
+# ndraw = fit_licor_draws$ndraws
+ndraw = 4000
+
+pred_licor2 = predict(fit_licor, new = licor2, summary = FALSE, ndraws = ndraw)
+
+aa = licor2 %>%
+  full_join(
+    pred_licor2 |>
+      matrix(ncol = 1) |>
+      set_colnames("A_model") |>
+      as_tibble() |>
+      mutate(
+        i = rep(seq_len(nrow(.)), each = ndraw),
+        draw = rep(seq_len(ndraw), nrow(.))
+      ),
+    by = join_by(i)
+  ) |>
+  pivot_wider(
+    id_cols = c("site_code", "draw"),
+    names_from = "aperture",
+    values_from = "A_model"
+  ) |>
+  mutate(AA = log(`two-sided` / `one-sided`)) |>
+  left_join(licor2 |>
+              filter(aperture == "two-sided") |>
+              select(site_code, gsw),
+            by = join_by(site_code))
+
+write_rds(aa, "objects/aa.rds")

### 2023-09-06 r/07_plot-licor.R

@@ -1,0 +1,128 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds") |>
+  mutate(label = c(
+"Makapu\u02bbu beach",                           
+"Hawai\u02bbi loa ridge",          
+"Ka\u02bbohe game\nmanagement area",
+"Puak\u014d petroglyph park",             
+"Koai\u02bba tree sanctuary",                   
+"Hāloa \u02bb\u0100ina",                   
+"Kaloko-Honok\u014dhau\nnational historical park",
+"Kaloko beach",
+"Wa\u02bbahila ridge",            
+"Kahuku Point",                          
+"Nu\u02bbuanu Pali Lookout",
+"Ka\u02bbena Point",                    
+"Mau\u02bbumae Ridge"
+)) |>
+  filter(site_code != "nnpl")
+
+# fit_licor = read_rds("objects/fit_licor.rds")
+
+aa = read_rds("objects/aa.rds") |>
+  filter(draw == 1L) |>
+  select(site_code, gsw) |>
+  left_join(site, by = join_by(site_code)) |>
+  add_site1() %>%
+  mutate(
+    A = predict(
+      fit_licor,
+      newdata = . |>
+        mutate(aperture = "two-sided") |>
+        unite("site_code_aperture", site_code, aperture)
+    ) |>
+      magrittr::extract(, "Estimate"),
+    treatment = "amphi"
+  )
+
+licor = read_rds("processed-data/licor.rds") |>
+  left_join(select(site, site, label, site_code, site_type), by = join_by(site_code)) |>
+  add_site1() |>
+  mutate(
+    treatment = fct_recode(aperture, pseudohypo = "one-sided", 
+                           amphi = "two-sided")
+  )
+
+new_licor = licor |>
+  summarize(
+    min_gsw = min(gsw),
+    max_gsw = max(gsw),
+    .by = c("site_code", "aperture", "site_type", "site_code_aperture")
+  ) |>
+  mutate(range_gsw = max_gsw - min_gsw) |>
+  crossing(.i = seq(0, 1, 0.1)) |>
+  mutate(gsw = min_gsw + .i * range_gsw) 
+
+pred_licor = bind_cols(
+  new_licor,
+  predict(fit_licor, newdata = new_licor) |>
+    as_tibble() |>
+    select(
+      A = Estimate,
+      A_lower = `Q2.5`,
+      A_upper = `Q97.5`
+    )
+) |>
+  add_site1() |>
+  mutate(
[…]

### 2023-09-06 r/08_plot-habitat-aa.R

@@ -1,0 +1,49 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+aa = read_rds("objects/aa.rds")
+
+# Prepare data for plotting habitat and site vs. AA
+df1 = aa |>
+  group_by(site_code) |>
+  point_interval(AA) |>
+  left_join(site, by = join_by(site_code)) |>
+  mutate(
+    habitat = factor(site_type, levels = c("montane", "coastal")),
+    site_code1 = fct_reorder(site_code, AA)
+  )
+
+# Prepare and write estimated difference in AA between habitats
+habitat_aa = aa |>
+  select(site_code, draw, AA) |>
+  left_join(select(site, site_code, site_type), by = join_by(site_code)) |>
+  summarize(AA = mean(AA), .by = c("site_type", "draw"))
+
+write_rds(habitat_aa, "objects/habitat_aa.rds") # all data
+
+habitat_aa1 = aa |>
+  select(site_code, draw, AA) |>
+  filter(!(site_code %in% c("pkpp", "knhp"))) |>
+  left_join(select(site, site_code, site_type), by = join_by(site_code)) |>
+  summarize(AA = mean(AA), .by = c("site_type", "draw"))
+
+write_rds(habitat_aa1, "objects/habitat_aa1.rds") # removing two sites with extrapolation
+
+# Prepare data for plotting habitat vs. average AA
+df2 = habitat_aa |>
+  summarize(AA = median(AA), .by = "site_type") |>
+  mutate(habitat = factor(site_type, levels = c("montane", "coastal")))
+
+ggplot(df1, aes(habitat, AA, ymin = .lower, ymax = .upper, color = habitat, 
+                group = site_code1)) +
+  geom_hline(yintercept = 0, color = "grey") +
+  geom_spoke(data = df2, aes(habitat, AA, color = habitat), 
+             inherit.aes = FALSE, angle = 0, radius = 0.45) +
+  geom_spoke(data = df2, aes(habitat, AA, color = habitat), 
+             inherit.aes = FALSE, angle = 0, radius = -0.45) +
+  geom_pointinterval(position = position_dodge()) +
+  ylab("amphistomy advantage") +
+  ylim(-0.1, 0.3) +
+  scale_color_manual(values = c("steelblue", "tomato"), guide = NULL)
+
+ggsave("figures/habitat-aa.pdf", width = 3, height = 4)

### 2023-09-06 r/09_plot-thickness-aa.R

@@ -1,0 +1,96 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+aa = read_rds("objects/aa.rds")
+fit_thickness = read_rds("objects/fit_thickness.rds")
+
+# Summarize population-level posteriors of AA and leaf thickness
+df_thickness_aa = fit_thickness |>
+  as_draws_df() |>
+  select(
+    "b_Intercept",
+    "b_site_typemontane",
+    matches("^r_site_code\\[[a-z]{4},Intercept\\]$"),
+    draw = .draw
+  ) |>
+  pivot_longer(matches("^r_site_code\\[[a-z]{4},Intercept\\]$")) |>
+  mutate(site_code = str_replace(name, "^r_site_code\\[([a-z]{4}),Intercept\\]$", "\\1")) |>
+  left_join(select(site, site_code, site_type), by = join_by(site_code)) |>
+  mutate(
+    log_leaf_thickness_um = b_Intercept +
+      (site_type == "montane") * b_site_typemontane + value,
+    site_code = factor(site_code, levels = unique(aa$site_code))
+  ) |>
+  select(site_code, site_type, draw, log_leaf_thickness_um) |>
+  full_join(aa, by = join_by(site_code, draw)) |>
+  filter(!is.na(AA))
+
+# Plot leaf thickness vs. AA
+
+## dataframe for plotting points
+df_points = df_thickness_aa |>
+  group_by(site_code, site_type) |>
+  point_interval(log_leaf_thickness_um, AA) |>
+  mutate(
+    leaf_thickness_um = exp(log_leaf_thickness_um),
+    leaf_thickness_um.lower = exp(log_leaf_thickness_um.lower),
+    leaf_thickness_um.upper = exp(log_leaf_thickness_um.upper),
+    habitat = factor(site_type, levels = c("montane", "coastal"))
+  ) 
+
+## dataframe for estimates and CIs of coefficients
+df_coef = df_thickness_aa |>
+  split(~ draw) |>
+  map_dfr(\(.d) {
+    fit = lm(AA ~ site_type + log_leaf_thickness_um, data = .d)
+    b = coef(fit)
+    tibble(intercept = b[1], b_montane = b[2], slope = b[3])
+  })
+
+## dataframe for plotting lines
+df_line = df_coef |>
+  cross_join(
+    df_points |>
+      ungroup() |>
+      summarize(
[…]

### 2023-09-06 r/10_calculate-gmaxratio.R

@@ -1,0 +1,281 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+aa = read_rds("objects/aa.rds")
+fit_stomata = read_rds("objects/fit_stomata.rds")
+
+# stomatal number and size per site type ----
+stomata_sitetype = fit_stomata |>
+  as_draws_df() |>
+  select(
+    .draw,
+    matches("^b_(low|upp)ernumberofstomata"),
+    matches("^b_(low|upp)erlengthum")
+  ) |>
+  mutate(
+    coastal_lower_stomata_number = b_lowernumberofstomata_Intercept,
+    coastal_upper_stomata_number = b_uppernumberofstomata_Intercept,
+    montane_lower_stomata_number = b_lowernumberofstomata_Intercept +
+        b_lowernumberofstomata_site_typemontane,
+    montane_upper_stomata_number = b_uppernumberofstomata_Intercept +
+        b_uppernumberofstomata_site_typemontane,
+    
+    coastal_lower_length_um = b_lowerlengthum_Intercept,
+    coastal_upper_length_um = b_upperlengthum_Intercept,
+    montane_lower_length_um = b_lowerlengthum_Intercept +
+      b_lowerlengthum_site_typemontane,
+    montane_upper_length_um = b_upperlengthum_Intercept +
+      b_upperlengthum_site_typemontane
+  ) |>
+  select(
+    .draw,
+    matches(
+      "^(coastal|montane)_(upper|lower)_(stomata_number|length_um)$"
+    )
+  ) |>
+  pivot_longer(
+    -.draw,
+    names_to = c("site_type", "surface", "trait", "unit"),
+    names_sep = "_",
+    values_to = "b_sitetype"
+  )
+
+# stomatal number and size per site ----
+stomata_site = fit_stomata |>
+  as_draws_df() |>
+  select(
+    .draw,
+    starts_with("r_site_code")
+  ) |>
+  pivot_longer(-.draw) |>
+  mutate(
+    surface = str_extract(name, "(upp|low)er"),
+    trait = str_extract(name, "numberofstomata|lengthum"),
+    s = glue("^r_site_code__{surface}{trait}\\[([a-z]{{4}}),Intercept\\]$"),
+    site_code = str_replace(name, s, "\\1") |>
+      str_extract("^[a-z]{4}$")
+  ) |>
[…]

### 2023-09-06 r/11_plot-habitat-gmaxratio.R

@@ -1,0 +1,58 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+gmaxratio_site = read_rds("objects/gmaxratio_site.rds")
+gmaxratio_ind = read_rds("objects/gmaxratio_ind.rds")
+
+# Prepare data for plotting habitat and site vs. gsmax-ratio
+df_site = gmaxratio_site |>
+  group_by(site_code) |>
+  point_interval(gmax_ratio) |>
+  left_join(site, by = join_by(site_code)) |>
+  mutate(
+    habitat = factor(site_type, levels = c("montane", "coastal")),
+    site_code1 = fct_reorder(site_code, gmax_ratio)
+  )
+
+df_ind = gmaxratio_ind |>
+  group_by(site_code, ind) |>
+  point_interval(gmax_ratio) |>
+  left_join(site, by = join_by(site_code)) |>
+  mutate(
+    habitat = factor(site_type, levels = c("montane", "coastal")),
+    site_code1 = factor(site_code, levels(df_site$site_code1))
+  )
+
+ggplot(
+  df_site,
+  aes(
+    habitat,
+    gmax_ratio,
+    ymin = .lower,
+    ymax = .upper,
+    color = habitat,
+    group = site_code1
+  )
+) +
+  geom_pointinterval(
+    data = df_ind,
+    position = position_jitterdodge(
+      jitter.width = 0.1,
+      dodge.width = 1,
+      seed = 20230823
+    ),
+    linewidth = NA,
+    alpha = 0.5
+  ) +
+  geom_pointinterval(
+    position = position_dodgenudge(x = 0.1),
+    size = 10,
+    shape = 21,
+    fill = "white",
+    linewidth = 1
+  ) +
+  ylab(expression(italic(g)[paste("smax, ratio")])) +
+  ylim(0, 0.5) +
+  scale_color_manual(values = c("steelblue", "tomato"), guide = NULL)
+
+ggsave("figures/habitat-gmaxratio.pdf", width = 4, height = 4)

### 2023-09-06 r/12_plot-gmaxratio-aa.R

@@ -1,0 +1,77 @@
+source("r/header.R")
+
+aa = read_rds("objects/aa.rds") |>
+  rename(.draw = draw)
+gmaxratio_ind = read_rds("objects/gmaxratio_ind.rds")
+gmaxratio_leaf = read_rds("objects/gmaxratio_leaf.rds")
+trait = read_rds("processed-data/trait.rds") |>
+  select(site_code, ind = individual, leaf = leaf_number, licor_leaf) |>
+  crossing(.draw = unique(aa$.draw))
+
+df_gmaxratio_aa = gmaxratio_ind |>
+  full_join(trait, by = join_by(.draw, site_code, ind)) |>
+  full_join(aa, by = join_by(.draw, site_code)) |>
+  filter(licor_leaf)
+
+# Plot gmaxratio vs. AA
+
+## dataframe for plotting points
+df_points = df_gmaxratio_aa |>
+  group_by(site_code, site_type) |>
+  point_interval(gmax_ratio, AA) |>
+  mutate(habitat = factor(site_type, levels = c("montane", "coastal"))) 
+
+## dataframe for estimates and CIs of coefficients
+df_coef = df_gmaxratio_aa |>
+  split(~ .draw) |>
+  map_dfr(\(.d) {
+    fit = lm(AA ~ site_type + gmax_ratio, data = .d)
+    b = coef(fit)
+    tibble(intercept = b[1], b_montane = b[2], slope = b[3])
+  })
+
+## dataframe for plotting lines
+df_line = df_coef |>
+  cross_join(
+    df_points |>
+      ungroup() |>
+      summarize(
+        min_gmax_ratio = min(gmax_ratio),
+        max_gmax_ratio = max(gmax_ratio),
+        .by = "habitat"
+      ) |>
+      crossing(i = seq(0, 1, length.out = 1e2)) |>
+      mutate(
+        r = max_gmax_ratio - min_gmax_ratio,
+        gmax_ratio = min_gmax_ratio + r * i
+      ) |>
+      select(habitat, gmax_ratio)
+  ) |>
+  mutate(
+    AA = intercept + b_montane * (habitat == "montane") +
+           slope * gmax_ratio,
+  ) |>
+  group_by(habitat, gmax_ratio) |>
+  point_interval(AA)
+
+gp = ggplot(df_points, aes(gmax_ratio, AA, color = habitat)) +
+    geom_lineribbon(
+      data = df_line, 
[…]

### 2023-09-06 r/13_plot-habitat-Ags.R

@@ -1,0 +1,37 @@
+source("r/header.R")
+
+site = read_rds("processed-data/site.rds")
+licor = read_rds("processed-data/licor.rds")
+
+site_Ags = licor |>
+  filter(aperture == "two-sided") |>
+  group_by(site_code) |>
+  summarise(across(gsw:A, max), Tleaf = Tleaf[which.max(A)]) |>
+  left_join(site, join_by(site_code)) |>
+  select(site_code, site_type, island, A, gsw, Tleaf) |>
+  pivot_longer(cols = A:gsw) |>
+  mutate(
+    name2 = case_when(
+      name == "A" ~ "paste(italic(A), ' [', mu, 'mol ', m^-2, ' ', s^-1, ']')",
+      name == "gsw" ~ "paste(italic(g)[sw], ' [mol ', m^-2, ' ', s^-1, ']')"
+    ),
+    habitat = factor(site_type, levels = c("montane", "coastal"))
+  )
+
+ggplot(site_Ags, aes(habitat, value, color = habitat)) +
+  facet_wrap(~ name2, scales = "free_y", labeller = label_parsed) +
+  geom_point(position = position_nudge(x = 0.1)) +
+  stat_summary(geom = "pointinterval", shape = 21, fill = "white",
+               fun.data = mean_se, size = 5) +
+  ylab("trait value") +
+  scale_color_manual(values = c("steelblue", "tomato"), guide = NULL)
+
+fit_habitat_Ags = site_Ags |>
+  split(~ name) |>
+  map(\(.x) {
+    t.test(value ~ site_type, data = .x)
+  })
+
+ggsave("figures/habitat-Ags.pdf", width = 5, height = 4)
+write_rds(site_Ags, "objects/site_Ags.rds")
+write_rds(fit_habitat_Ags, "objects/fit_habitat_Ags.rds")

### 2023-09-06 r/14_plot-traits-aa.R

@@ -1,0 +1,19 @@
+source("r/header.R")
+
+plot_gmaxratio_aa = read_rds("objects/plot_gmaxratio_aa.rds") +
+  theme(legend.position = "none")
+
+plot_thickness_aa = read_rds("objects/plot_thickness_aa.rds") +
+  theme(legend.position = "top")
+
+habitat_legend = get_legend(plot_thickness_aa)
+
+plot_thickness_aa = plot_thickness_aa +
+  theme(legend.position = "none")
+
+gp1 = plot_grid(plot_thickness_aa, plot_gmaxratio_aa, nrow = 1, labels = "AUTO")
+
+gp = grid.arrange(grobs = list(habitat_legend, gp1), ncol = 1, nrow = 2, 
+                  heights = c(0.1, 0.9))
+
+ggsave("figures/traits-aa.pdf", gp, width = 6.5, height = 4)

### 2023-09-06 r/functions.R

@@ -1,0 +1,170 @@
+# Calculations of gsmax follow Sack and Buckley 2016
+biophysical_constant = function(D_wv, v) D_wv / v
+
+morphological_constant = function(c, h, j) {
+  (pi * c ^ 2) / (j ^ 0.5 * (4 * h * j + pi * c))
+}
+
+# Recalculate LI-6800 data with different S and K values
+recalc_licor = function(licor_data, S, K, dynamic = FALSE, use_EB = FALSE) {
+  
+  assert_number(S, lower = 0, upper = 6)
+  assert_numeric(K, lower = 0, any.missing = FALSE, min.len = 1L,
+                 max.len = nrow(licor_data))
+  assert_flag(dynamic)
+  
+  vars = c(
+    "obs", "Flow", "H2O_s", "H2O_r", "CorrFact", "Qamb_in", "Qamb_out", "Q",
+    "f_farred", "Q_red", "Q_modavg", "Q_blue",
+    "Tair", "Tleaf", "Tleaf2", "Fan_speed", "Pa", "TleafCnd", "ΔPcham",
+    "Ca", "CO2_r", "CO2_s"
+  )
+  
+  if (dynamic) {
+    vars = c(vars, "Crd", "Csd",	"dCsd/dt", "αVc")
+  }
+  
+  ret = licor_data |>
+    select(all_of(vars)) |>
+    mutate(
+      
+      # System constants
+      Oxygen = 21,
+      
+      # Chamber constants for calculating gbw
+      blc_a = 0.578,
+      blc_b = 0.5229739,
+      blc_c = 0.003740252,
+      blc_d = -0.06197961,
+      blc_e = -0.005608586,
+      blc_minS = 1,
+      blc_maxS = 6,
+      blc_Po = 96.9,
+      
+      # Leaf constants
+      S = S,
+      K = K,
+      
+      # Leaf temperature constants
+      # Determines whether to use Tleaf, Tleaf2, or TleafEB
+      deltaTw = 0, fT1 = !use_EB, fT2 = 0, fTeb = use_EB,
+      
+      # Leaf absorbtance constants
+      abs_ambient	= 0.76,
+      abs_redLED = 0.84,
+      abs_greenLED = 0.7,
+      abs_blueLED = 0.87,	
+      abs_whiteLED = 0.75,
+      abs_redFlr = 0.84,
[…]

### 2023-09-06 r/header.R

@@ -1,0 +1,31 @@
+rm(list = ls())
+
+library(brms)
+library(checkmate)
+library(cowplot)
+library(dplyr)
+library(forcats)
+library(ggdist)
+library(ggplot2)
+library(ggpp)
+library(glue)
+library(gridExtra)
+library(magrittr)
+library(photosynthesis)
+library(purrr)
+library(readr)
+library(stringr)
+library(tidybayes)
+library(tidyr)
+
+theme_set(theme_cowplot())
+
+source("r/functions.R")
+
+# Scale:
+# 1024 pixels / 1089.16 um = 0.9401741 pixels / um
+# images are 1089.16 * 816.87 um = 889702.1 um ^ 2 = 0.8897021 mm ^ 2
+# 1024 * 768 pixels = 786432 ^ 2
+pixels_per_um = 1024 / 1089.16
+pixels2_per_mm2 = (1024 * 768) / (1089.16 * 816.87) # == pixels_per_um^2
+image_area_mm2 = 0.8897021

### 2023-09-06 r/install-packages.R

@@ -1,0 +1,19 @@
+install.packages("brms")
+install.packages("checkmate")
+install.packages("cowplot")
+install.packages("dplyr")
+install.packages("forcats")
+install.packages("ggdist")
+install.packages("ggplot2")
+install.packages("ggpp")
+install.packages("glue")
+install.packages("gridExtra")
+install.packages("magrittr")
+install.packages("photosynthesis")
+install.packages("purrr")
+install.packages("readr")
+install.packages("stringr")
+install.packages("tidybayes")
+install.packages("tidyr")
+
+install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

### 2023-09-06 ms/ms.qmd
fixing issues with make

@@
 
-\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term ʻpseudohypostomyʻ.
+\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy".
 We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).
 
-\textbf{Key results}  Coastal ʻilima leaves benefit 4.58 times more from amphistomy compared to their montane counterparts.
+\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts.
 Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.

### 2023-09-06 r/07_plot-licor.R

@@ -18,7 +18,7 @@ site = read_rds("processed-data/site.rds") |>
 )) |>
   filter(site_code != "nnpl")
 
-# fit_licor = read_rds("objects/fit_licor.rds")
+fit_licor = read_rds("objects/fit_licor.rds")
 
 aa = read_rds("objects/aa.rds") |>
   filter(draw == 1L) |>

### 2023-09-06 r/02_process-trait-data.R
added climate to raw-data and fixed makefile bugs

@@ -7,14 +7,7 @@ dir = "raw-data/climate"
 tair_ann = raster::raster(glue("{dir}/Tair_month_raster/tair_ann"))
 sl_mst_ann = raster::raster(glue("{dir}/SoilMoisture_month_raster/sl_mst_ann"))
 
-# Read in population information ----
-write_csv(site, "raw-data/site.csv")
-write_csv(plant, "raw-data/plant.csv")
-write_csv(licor_leaf, "raw-data/licor_leaf.csv")
-write_csv(stomata_density, "raw-data/stomata_density.csv")
-write_csv(stomata_size, "raw-data/stomata_size.csv")
-write_csv(leaf_thickness, "raw-data/leaf_thickness.csv")
-
+# Read in raw data ----
 site = read_csv("raw-data/site.csv", col_types = "ccccccDdddd")
 
 plant = read_csv("raw-data/plant.csv", col_types = "ccdd") |>

### 2023-09-07 ms/ms.qmd
final checks before submission

@@
 
-```{=html}
-
-<!-- TO DO,
-
-1. at end of INTRODUCTION, address Tom's comments
-
-"suggest you rigorously define sun leaves and shade leaves here, and dispense with the quotes."
-
-Brief Communications—Brief Communications are short (c 3000–4000 word) research articles reporting exciting, significant new findings.
-They include no more than 4 figures and tables, combined.
-Manuscripts, and their abstracts, should be organized as described for Research articles.
-Authors must adhere to the same membership and page charge requirements as for regular Research Articles.
-
-a. Title and Running Heads
-The manuscript title for research papers should be specific and informative, conveying the key findings of the research in an active voice.
-Provide a title written with sentence-style capitalization.
-Below the title, list authors: each author’s first name, middle initial, surname.
-On the next line, give affiliation and unabbreviated address.
-If authors have different affiliations and addresses, add a superscript number after each author’s name to indicate the footnoted address.
-Include another footnote superscript number to indicate the author for correspondence.
-
-Include the line: Manuscript received _______; revision accepted _______.
-
-After this text, add a short title no longer than 65 characters (including spaces) to be used as a running head in the article.
-
-b. Abstract
-AJB requires structured abstracts for Research Articles and Brief Communications. The abstract is 250 words or fewer, written in the following structured format:
-• Premise of the study (why the work was done, what major questions of plant biology are addressed, and why it is important to the broad AJB readership)
-• Methods
-• Key results
-• Conclusions (what major points should the reader take from this article)
-
-The abstract may be all a reader sees initially, so it should be written to capture the interest of the general botanical community as well as specialists.
-Avoid references; if essential, cite parenthetically with journal name, volume number, pages, and year.
--->
-```
 ```{r, echo = FALSE, message = FALSE, warning = FALSE}
@@
                  .width = 0.95, .point = median, .interval = qi)
+
+AA_coastal1 = pretty_est(habitat_aa1$coastal,
+                        habitat_aa1$coastal.lower,
+                        habitat_aa1$coastal.upper,
+                        digits = 2L)
 
@@
 
-Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts (\autoref{fig:compare-circuit-diagrams}), yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why.
+Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why.
 Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010].
@@
 Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of adaxial stomata increases from \~10-20% under low light to \~30-40% under high light [@james_leaf_2001].
-In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.
+In summary, both genetic and plastic responses evince a widespread association between light and $\mathrm{SR}$.
 
@@
 A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface.
-In other words, amphistomy doesn't make much difference if one leaf surface has very open stomata on it.
+In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it.
 We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5.
-We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that takes into account differences in both stomatal density and size between surfaces.
+We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.
 
@@
 Many lineages have adapted to different light habitats after colonization and leaf anatomical traits such as $\mathrm{SR}$ and thickness vary within and among closely related species.
-It is hypothesized that the common ancestor in many Hawaiian clades was a "weedy" species with high dispersal ability adapted to open habitats [@carlquist_biota_1966].
+It is hypothesized that the common ancestor in many Hawaiian clades was a weedy species with high dispersal ability adapted to open habitats [@carlquist_biota_1966].
 Colonization was followed by adaptive radiation into higher elevation, montane, closed, forested habitats.
@@
 These individuals have smooth, green foliage with serrated edges.
-Montane populations exhibit traits that help them to compete for light availability (Raunch et al.
-1997).
+Montane populations exhibit traits that may help them to compete for light availability.
 This montane morphology is not found in *S.
 fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022].
-The variable morphology between coastal and montane populations has even led some scientists to argue that it makes up a species complex, rather than a single species (Raunch et al.
-1997).
 
-The objectives of our study are to test whether 1) 'sun' leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than 'shade' leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?
+Because of their contrasting habitat and morphology, we treat leaves from coastal and montane plants as representatives of sun and shade leaves, respectively, for testing hypotheses about amphistomy advantage.
+Specifically, the objectives of our study are to test whether 1) sun leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than shade leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?
 
-```{=html}
-<!-- TEXT FROM PROPOSAL THAT MAY HAVE SOMETHING USEFUL:
-
-*Knowing the adaptive significance of amphistomy will advance our understanding of leaf structure-function relationships, aid reconstructing paleoenvironments, and guide crop optimization*.
-Stomata are studied in disparate fields such as cell biology, ecosystem ecology, and paleontology [@hetherington_role_2003; @bergmann_stomatal_2007; @berry_stomata:_2010; @mcelwain_paleoecology_2017].
-The intense focus on stomata makes it all the more surprising that we know relatively little about the adaptive significance of where stomata are on the leaf!
-Amphistomy is deeply integrated with leaf traits associated with high photosynthetic capacity and leaf water transport capacity [@smith_leaf_1997; @drake_two_2019; @xiong_one_2020].
-Knowing the upper limits of photosynthesis will help breed more productive crops [@long_can_2006] and predict future terrestrial carbon sinks at higher atmospheric CO$_2$ [@keenan_terrestrial_2018].
-In paleontology, stomata visible on fossil leaves help reconstruct paleo-CO$_2$ concentrations based on density/index [@konrad_modelling_2008; @franks_co2-forced_2009; @mcelwain_paleoecology_2017; but see @jordan_links_2020], but stomatal density ratio ($\mathrm{SR}$) may add information about whether plants lived in open or closed vegetation [@salisbury_i_1928; @jordan_using_2014; @bucher_stomatal_2017].
-Finally, $\mathrm{SR}$ tends to increase (higher proportion of stomata on upper surface) during domestication [@milla_shifts_2013], but we do not know why.
-Our work will help infer how changes in $\mathrm{SR}$ and other leaf anatomical traits affect photosynthesis and water use in crops.
-This is particularly promising because we may learn ways to increase leaf photosynthesis without increasing water loss by altering $\mathrm{SR}$ and/or the regulation of stomatal aperture on each leaf surface.
-
-To redress these gaps in our knowledge, in Aim 1 we will model the benefits and costs of amphistomy under low- and high-light conditions and in Aims 2-3 we will test model predictions using the wild tomatoes as model clade (*Solanum* spp.; Solanaceae).
-Specifically, we will experimentally manipulate whether gas exchange proceeds through one or both leaf surfaces, holding all else constant.
-This will enable us to measure how much amphistomy increases photosynthesis compared to an otherwise identical hypostomatous leaf (Aim 2).
-In Aim 3, we will test whether amphistomy is costly in shade leaves because it causes dehydration of the palisade mesophyll under high evaporative demand.
-The result of this project will determine whether high light, photosynthetic capacity, and adequate hydration are sufficient to explain the distribution of amphistomatous leaves in nature, as Mott and colleagues hypothesized 40 years ago.
-
-The dearth of amphistomatous leaves implies a fitness cost associated with stomata on the upper leaf surface in particular [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984-1; @muir_is_2019].
-Two nonmutually exclusive hypotheses for the cost of upper stomata have received the most attention, but neither has been tested experimentally.
-We will test one hypothesis, which is that amphistomy hinders photosynthesis by causing dehydration of the palisade mesophyll under high evaporative demand [@buckley_sites_2017; @drake_two_2019].
-This cost of amphistomy could be greater in shade leaves because they dehydrate severely when exposed to intense sunflecks, warm temperatures, and/or low humidity, whereas sun leaves have constitutive traits that mitigate the risk of dehydration.
-Sensitivity to dehydration would be more consequential if it affected the palisade mesophyll, which is adjacent to the illuminated upper surface and is thus more photosynthetically active [@smith_leaf_1997; @evans_profiles_2003; @earles_embracing_2019] than the spongy mesophyll, which is adjacent to the lower surface (\autoref{fig:compare-circuit-diagrams}).
-An alternative hypothesis with some empirical support is that stomata on the upper surface are more likely to be colonized by bacterial or fungal pathogens, presumably because a propagule is more likely to land on the upper surface [@gutschick_photosynthesis_1984-1; @mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021].
-The current proposal does not address this pathogen susceptibility hypothesis because PI Muir is seeking funding for this work elsewhere.
-Although not the focus of our aims, our modeling and data can address two less well theorized alternative hypotheses that upper stomata compete for space on the epidermis [@de_boer_optimal_2016; @haworth_allocation_2018; @muir_how_2023] or alter palisade mesophyll structure [@dow_disruption_2017; @lundgren_mesophyll_2019; @baillie_developmental_2020].
-
-*Despite the fact that we do not know the adaptive significance of amphistomy, the distribution of hypo- and amphistomatous leaves is nonrandom and most likely caused by natural selection.* Consistent relationships between stomatal traits, growth form, and light environment strongly suggest convergent evolution, a hallmark of adaptive evolution.
-Hypo- and amphistomatous leaves are two ends of a continuum in $\mathrm{SR}$.
-Hypostomatous leaves have a $\mathrm{SR}$ close to 0; fully amphistomatous leaves with equal densities on each leaf have a $\mathrm{SR}$ close to 0.5.
-Herbaceous plants from high-light habitats tend to be amphistomatous and have higher $\mathrm{SR}$ values than herbaceous plants from shaded habitats or woody species [@salisbury_i_1928; @parkhurst_adaptive_1978; @mott_adaptive_1982; @parkhurst_intercellular_1990; @peat_comparative_1994; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018].
-A second line of evidence for adaptation is that the distribution of $\mathrm{SR}$ among species is inconsistent with the expected distribution for a ratio trait evolving neutrally; instead, a nonneutral bimodal distribution evolves repeatedly in land plants [@muir_making_2015].
-Finally, amphistomatous leaves are strongly coordinated with other traits involved with high photosynthetic capacity [@beerling_evolutionary_1996; @xiong_one_2020] and domestication [@milla_shifts_2013], suggesting that amphistomy is favored by selection in some circumstances.
-What specific fitness costs and benefits underlie these adaptive patterns?
-
-*The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light*.
-To reiterate, herbaceous leaves in high light habitats tend to be amphistomatous.
-Most variation in $\mathrm{SR}$ among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light.
-Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop ~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991].
-Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop ~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975].
-Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of upper stomata increases from ~10-20% under low light to ~30-40% under high light [@james_leaf_2001].
-In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.
-
-*We don’t know whether amphistomy is an adaptation to high light intensity because no one has shown experimentally that amphistomy* per se *increases CO$_2$ assimilation more under high light than under low light*.
-Other factors correlated with light habitat could drive selection on $\mathrm{SR}$ and result in a spurious association between light habitat and $\mathrm{SR}$.
-Even if light intensity is the primary causal selective agent for $\mathrm{SR}$, its effect on selection for higher/lower $\mathrm{SR}$ could be the result of 1) *facultative* responses to light intensity and/or 2) *constitutive* differences between shade and sun leaves (\autoref{fig:causal-graph}).
-The *facultative* hypothesis is that amphistomy increases photosynthesis under high light intensity because of high CO$_2$ demand and/or CO$_2$ supply is constrained by low $g_\text{ias}$.
-Under this hypothesis, the advantage of amphistomy depends on the immediate light intensity, not the light environment in which plants develop.
-The nonmutually exclusive *constitutive* hypothesis is that developmental plasticity results in 'sun' and 'shade' leaves with manifold differences in anatomy and biochemistry such as thicker leaves, denser cell packing, or increased Rubisco concentration [@terashima_why_2001; @lambers_plant_2008].
-Sun leaves benefit more from amphistomy than shade leaves under all light intensities because they have lower $g_\text{ias}$ and/or greater CO$_2$ demand.
-The proposed research will quantify the relative contributions of facultative and constitutive effects on the photosynthetic advantage of amphistomy.
-We will describe predictions of these hypotheses in more detail below.
-
-MOVE THIS TO INTRO Parkhurst (1978) introduced the concept of “amphistomy advantage” to quantify the photosynthetic advantage of an amphistomatous leaf over an otherwise identical hypostomatous leaf.
-An important limitation of his approach is that he assumed photosynthesis was simply proportional to the total diffusive conductance from the atmosphere to the sites of carboxylation; that tacitly assumes CO$_2$ concentration at the sites of carboxylation is zero, which in turn infinite local did not model assumes CO$_2$ demand is locally infinite throughout the leaf.
-In reality, photosynthesis at any given point in the leaf depends on the local concentrations of CO$_2$, chlorophyll, and Rubisco, and the local irradiance.
-Thus, some mesophyll layers can draw down chloroplastic CO$_2$ concentration to a low level and potentially assimilate a lot of carbon when CO$_2$ is adequately supplied, whereas other layers have much weaker CO$_2$ demand.
-By not incorporating a CO$_2$ demand function, Parkhurst assumed that any increase in CO$_2$ supply was matched by a concomitant increase in CO$_2$ assimilation.
-We now know that increasing assimilation depends on both CO$_2$ supply and demand.
-If supply increases, but the demand function is flat because of low light intensity and/or low photosynthetic capacity, then $A$ won’t change much (demand-limited).
-Conversely, if the demand function is steep because there is excess light and/or enzymatic capacity, then increased CO$_2$ supply will yield a large increase in $A$ (supply-limited).
-
- -->
-```
 ```{=tex}
@@
 ```
+
 # MATERIALS AND METHODS --- {#methods}
@@
 
-We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting [@yorkston_interfertility_2006] and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}).
+We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}).
 We avoided sites that appeared to be cultivated.
 We visited sites between August and November 2022.
-For logistical reasons, the sites on Hawaiʻi were sampled in one three-day trip.
+For logistical reasons, the sites on Hawaiʻi were sampled during one three-day trip.
 We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site.
@@
 We did not sample small individuals if there was risk removing leaves would cause mortality.
-From each plant, we collected two fully expanded leaves for traits measurements.
+From each plant, we collected two fully expanded leaves for trait measurements.
 We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.
@@
 
-We cut thin sections using two razor blades taped together and cutting the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.
+We cut thin sections using two razor blades taped together.
+We sectioned the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.
 
@@
 
-At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA).
+At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA).
 We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$.
 We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification.
-For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values, holding other environmental variables constant.
+For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values.
 We refer to this as an $A \textendash g_\text{sw}$ curve.
-We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of "pseudohypostomatous" leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
+We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
+Hypostomy refers to leaves with stomata only present on the lower, typically abaxial, surface.
 We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively.
@@
 After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation.
+All other environmental conditions in the leaf chamber remained the same.
 We logged data until $g_\text{sw}$ reached its nadir.
@@
 It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$.
-To overcome this, we fit $A \textendash g_\text{sw}$ curves using a flexible generalized additive model (GAM) to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$.
+To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$.
 In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e.
@@
 In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves.
-We fit GAMs using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] for Bayesian statistical inference with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
+We fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
 We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
-We sampled from posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations.
+We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations.
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
@@
 
-We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with site-type as a fixed effect and individual plant and site as random effects.
+We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with habitat as a fixed effect and individual plant and site as random effects.
 We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023].
@@
 We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population.
-We used this approach most of the variation in leaf thickness was among sites (see Results) and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation.
+We used this approach most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation.
 We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable.
@@
 Counts were modeled as negative binomially distributed variable from a latent stomatal density and a parameter $\phi$ to estimate overdispersion in counts relative to a Poisson model.
-For all traits, the explanatory variables were site-type as a fixed effect and leaf within individual plant, individual plant, and site as random effects.
+For all traits, the explanatory variables were habitat as a fixed effect and leaf within individual plant, individual plant, and site as random effects.
 We used the *R* package **brms** version `r fit_stomata$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_stomata$version$cmdstan` and **cmdstanr** version `r fit_stomata$version$cmdstanr` [@gabry_cmdstanr_2023].
@@
 Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$.
-The difference between coastal and montane leaves is lower,  $`r AA_diff1`$, but still greater than 0 if we remove two sites where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves.
+Posterior predictions closely match observed values of $A$ (\autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably.
+It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain.
+When we remove two leaves where extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.
 Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}).
@@
   \includegraphics{../figures/traits-aa.pdf}
-  \caption{Relationships between leaf amphistomy advantage, (A) $g_\mathrm{smax,ratio}$ and (B) leaf thickness among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi., and amphistomy advantage.
+  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
 A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
@@
 Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
-$g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
+Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
   \label{fig:traits-aa}
@@
 By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
-This allows us to quantity the ‘amphistomy advantage’ ($\mathrm{AA}$) holding all else constant.
-Taking advantage of the of the steep climatic gradients in the Hawaiian archipegalo, we applied this new method to show for the first time that sun leaves benefit more from amphistomy than shade leaves in ʻilima (*Sida fallax*).
-While this result has not yet been validated in other speciesthere are limits to how much we can generalize to all land plants, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.
+This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant.
+Taking advantage of the of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs.
+$\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$).
+While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.
 
@@
 Our results do not support this hypothesis.
-Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-AA}B).
+Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-aa}B).
 Among coastal sites, the site with the lowest $g_\mathrm{smax,ratio}$ had the lowest $\mathrm{AA}$, but there was little variation in $g_\mathrm{smax,ratio}$ among coastal leaves in our sample.
@@
 
-Two limitations of this study are the small sample sample sizes and experimental design that precludes distinguishing genetic from environmental differences in leaf traits.
-Understanding the mechanistic basis of higher $\mathrm{A}A$ in sun leaves would require much larger sample sizes.
+Two limitations of this study are the small sample size and experimental design that precludes distinguishing genetic from environmental differences in leaf traits.
+Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes.
 Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008].
@@
 Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*.
-While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be useful understand genetic and environmental contributions to trait variation.
+While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation.
 A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.
@@
 Raw data will be deposited on Dryad with a DOI and stable URL upon publication.
+[THE GITHUB REPO AND DRYAD DATA ARE AVAILABLE TO REVIEWERS]
 
@@
 
-Additional supporting information may be found online in the Supporting Information section at the end of the article
+Additional supporting information may be found online in the Supporting Information section at the end of the article.
 
@@
   \caption{}
+  \label{fig:pp-licor}
+\end{figure}
+```
+```{=tex}
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
   \label{fig:habitat-Ags}
@@
 
+\newpage
+
 # Literature Cited --- {#literature-cited}

### 2023-09-07 ms/ms.tex
final checks before submission

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 stomatal conductance but with gas exchange blocked through the upper
-(adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated
+(adaxial) surface, which we term ``pseudohypostomy''. We thus estimated
 \(\mathrm{AA}\) and related physiological and anatomical traits in 12
@@
 
-\textbf{Key results} Coastal ʻilima leaves benefit 4.58 times more from
+\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
 amphistomy compared to their montane counterparts. Our evidence was
@@
 broad leaves, should increase carbon gain by reducing the average
-diffusion pathlength between stomata and chloroplasts
-(\autoref{fig:compare-circuit-diagrams}), yet paradoxically this
-seemingly simple adaptation is uncommon in nature and we don't know why.
-Understanding variation in stomatal traits like amphistomy is imperative
-because these tiny pores play an outsized ecological role in the global
-carbon and water cycles (Hetherington and Woodward, 2003; Berry et al.,
-2010). A widely applicable, accurate representation of how stomata
-mediate the relationship between CO\(_2\) gained through photosynthesis
-and water lost through evapotranspiration is essential to predict future
-climate using Earth Systems Models (Jarvis, 1976; Ball et al., 1987;
-Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997). Optimality
-models accurately predict the major cause of water loss, stomatal
-conductance (\(g_\mathrm{sw}\)), by assuming plants maximize carbon gain
-minus a cost of water (Cowan and Farquhar, 1977; Givnish, 1986; Medlyn
-et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks et al., 2018;
-Deans et al., 2020; Franklin et al., 2020; Wang et al., 2020; Harrison
-et al., 2021). Despite the success of optimality modeling in predicting
-\(g_\mathrm{sw}\), the same modeling approach has so far failed to
-explain the rarity of amphistomatous leaves (Muir, 2019). \textbf{This
-gap between theory and observations strongly implies that we remain
-ignorant about some key benefits and costs associated with stomata.}
+diffusion pathlength between stomata and chloroplasts, yet paradoxically
+this seemingly simple adaptation is uncommon in nature and we don't know
+why. Understanding variation in stomatal traits like amphistomy is
+imperative because these tiny pores play an outsized ecological role in
+the global carbon and water cycles (Hetherington and Woodward, 2003;
+Berry et al., 2010). A widely applicable, accurate representation of how
+stomata mediate the relationship between CO\(_2\) gained through
+photosynthesis and water lost through evapotranspiration is essential to
+predict future climate using Earth Systems Models (Jarvis, 1976; Ball et
+al., 1987; Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997).
+Optimality models accurately predict the major cause of water loss,
+stomatal conductance (\(g_\mathrm{sw}\)), by assuming plants maximize
+carbon gain minus a cost of water (Cowan and Farquhar, 1977; Givnish,
+1986; Medlyn et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks
+et al., 2018; Deans et al., 2020; Franklin et al., 2020; Wang et al.,
+2020; Harrison et al., 2021). Despite the success of optimality modeling
+in predicting \(g_\mathrm{sw}\), the same modeling approach has so far
+failed to explain the rarity of amphistomatous leaves (Muir, 2019).
+\textbf{This gap between theory and observations strongly implies that
+we remain ignorant about some key benefits and costs associated with
+stomata.}
 
@@
 \textasciitilde30-40\% under high light (James and Bell, 2001). In
-summary, both genetic and plastic responses to evince a widespread
+summary, both genetic and plastic responses evince a widespread
 association between light and \(\mathrm{SR}\).
@@
 amphistomatous leaf with most stomata on one surface. In other words,
-amphistomy doesn't make much difference if one leaf surface has very
-open stomata on it. We therefore predict that sun leaves will have
-greater \(\mathrm{AA}\) possibly because they have thicker leaves and/or
+amphistomy doesn't make much difference if one leaf surface has few open
+stomata on it. We therefore predict that sun leaves will have greater
+\(\mathrm{AA}\) possibly because they have thicker leaves and/or
 \(\mathrm{SR}\) closer to 0.5. We actually report
 \(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
-that takes into account differences in both stomatal density and size
+that it accounts for differences in both stomatal density and size
 between surfaces.
@@
 and among closely related species. It is hypothesized that the common
-ancestor in many Hawaiian clades was a ``weedy'' species with high
-dispersal ability adapted to open habitats (Carlquist, 1966).
-Colonization was followed by adaptive radiation into higher elevation,
-montane, closed, forested habitats. Consequently, adaptation to sun and
-shade is a common axis of phenotypic variation among Hawaiian plants
-such as lobeliads (Givnish et al., 2004; Montgomery and Givnish, 2008;
-Givnish et al., 2009; Givnish and Montgomery, 2014; Scoffoni et al.,
-2015), \emph{Bidens} (Carlquist, 1966; Knope et al., 2020),
-\emph{Scaevola} (Robichaux and Pearcy, 1984; McKown et al., 2016),
-\emph{Euphorbia} (Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al.,
-2009).
+ancestor in many Hawaiian clades was a weedy species with high dispersal
+ability adapted to open habitats (Carlquist, 1966). Colonization was
+followed by adaptive radiation into higher elevation, montane, closed,
+forested habitats. Consequently, adaptation to sun and shade is a common
+axis of phenotypic variation among Hawaiian plants such as lobeliads
+(Givnish et al., 2004; Montgomery and Givnish, 2008; Givnish et al.,
+2009; Givnish and Montgomery, 2014; Scoffoni et al., 2015),
+\emph{Bidens} (Carlquist, 1966; Knope et al., 2020), \emph{Scaevola}
+(Robichaux and Pearcy, 1984; McKown et al., 2016), \emph{Euphorbia}
+(Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al., 2009).
 
@@
 tall with strong, woody stems. These individuals have smooth, green
-foliage with serrated edges. Montane populations exhibit traits that
-help them to compete for light availability (Raunch et al.~1997). This
-montane morphology is not found in \emph{S. fallax} populations on other
-Pacific Islands (Pejhanmehr, 2022). The variable morphology between
-coastal and montane populations has even led some scientists to argue
-that it makes up a species complex, rather than a single species (Raunch
-et al.~1997).
-
-The objectives of our study are to test whether 1) `sun' leaves of
-coastal ʻilima plants will have greater \(\mathrm{AA}\) than `shade'
-leaves of montane plants; and if so, is this because 2a) coastal plants
-have thicker leaves than montane plants and/or 2b) coastal plants have a
-\(g_{\mathrm{smax,ratio}}\) closer to 0.5?
+foliage with serrated edges. Montane populations exhibit traits that may
+help them to compete for light availability. This montane morphology is
+not found in \emph{S. fallax} populations on other Pacific Islands
+(Pejhanmehr, 2022).
+
+Because of their contrasting habitat and morphology, we treat leaves
+from coastal and montane plants as representatives of sun and shade
+leaves, respectively, for testing hypotheses about amphistomy advantage.
+Specifically, the objectives of our study are to test whether 1) sun
+leaves of coastal ʻilima plants will have greater \(\mathrm{AA}\) than
+shade leaves of montane plants; and if so, is this because 2a) coastal
+plants have thicker leaves than montane plants and/or 2b) coastal plants
+have a \(g_{\mathrm{smax,ratio}}\) closer to 0.5?
 
@@
 
-We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on
-Hawaiʻi Island by consulting (Yorkston and Daehler, 2006) and citizen
+We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on
+Hawaiʻi Island by consulting Yorkston and Daehler (2006) and citizen
 scientist records on iNaturalist (Anon, 2022)
@@
 November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
-in one three-day trip. We haphazardly sampled eight plants distributed
-evenly between the highest and lowest elevation plants along a transect
-at each site. For safety and conservation reasons, transects were along
-a trail or road. We did not sample small individuals if there was risk
-removing leaves would cause mortality. From each plant, we collected two
-fully expanded leaves for traits measurements. We sampled stomatal
-traits on all leaves; leaf thickness on one leaf from three randomly
-selected plants per site; and, due to limited time, a single leaf from a
-single plant at the middle of each transect for gas exchange
+during one three-day trip. We haphazardly sampled eight plants
+distributed evenly between the highest and lowest elevation plants along
+a transect at each site. For safety and conservation reasons, transects
+were along a trail or road. We did not sample small individuals if there
+was risk removing leaves would cause mortality. From each plant, we
+collected two fully expanded leaves for trait measurements. We sampled
+stomatal traits on all leaves; leaf thickness on one leaf from three
+randomly selected plants per site; and, due to limited time, a single
+leaf from a single plant at the middle of each transect for gas exchange
 measurements.
@@
 
-We cut thin sections using two razor blades taped together and cutting
+We cut thin sections using two razor blades taped together. We sectioned
 the leaf in a petri dish of water, wet-mounted sections onto a slide,
@@
 the middle of the transect for gas exchange measurements using a
-portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln,
+portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
 NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
@@
 rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
-range of \(g_\text{sw}\) values, holding other environmental variables
-constant. We refer to this as an \(A \textendash g_\text{sw}\) curve. We
-compared the \(A \textendash g_\text{sw}\) curve of the untreated leaf
-to the photosynthetic rate of ``pseudohypostomatous'' leaf
-(\(A_\mathrm{hypo}\)), which is the same leaf but with gas exchange
-through the upper surface blocked by a neutral density plastic
-(propafilm). We refer to the untreated and partially blocked leaves as
+range of \(g_\text{sw}\) values. We refer to this as an
+\(A \textendash g_\text{sw}\) curve. We compared the
+\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
+photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
+which is the same leaf but with gas exchange through the upper surface
+blocked by a neutral density plastic (propafilm). Hypostomy refers to
+leaves with stomata only present on the lower, typically abaxial,
+surface. We refer to the untreated and partially blocked leaves as
 ``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
@@
 decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
-closure without photosynthetic downregulation. We logged data until
+closure without photosynthetic downregulation. All other environmental
+conditions in the leaf chamber remained the same. We logged data until
 \(g_\text{sw}\) reached its nadir. We then repeated the process with the
@@
 \(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
-we fit \(A \textendash g_\text{sw}\) curves using a flexible generalized
-additive model (GAM) to interpolate modeled \(A\) for amphi and
-pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12 leaves, the
-minimum \(g_\text{sw}\) of the amphi curve was smaller than the maximum
-\(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves overlapped for
-a range of \(g_\text{sw}\) values). In those cases, we estimated
-\(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle of the
-range of overlap. In 2 of 12 leaves, the \(A \textendash g_\text{sw}\)
-curves did not quite overlap because the minimum \(g_\text{sw}\) of the
-amphi curve was slightly greater than the maximum \(g_\text{sw}\) of the
-pseudohypo curve. In those cases, we estimated \(\mathrm{AA}\) by
-extrapolating slightly,
+we fit \(A \textendash g_\text{sw}\) curves using a linear regression of
+\(\text{log}(g_\mathrm{sw})\) on \(A\) to interpolate modeled \(A\) for
+amphi and pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12
+leaves, the minimum \(g_\text{sw}\) of the amphi curve was smaller than
+the maximum \(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves
+overlapped for a range of \(g_\text{sw}\) values). In those cases, we
+estimated \(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle
+of the range of overlap. In 2 of 12 leaves, the
+\(A \textendash g_\text{sw}\) curves did not quite overlap because the
+minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
+maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
+estimated \(\mathrm{AA}\) by extrapolating slightly,
 \(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
 beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
-curves. We fit GAMs using the \emph{R} package \textbf{brms} version
-2.20.1 (Bürkner, 2017) for Bayesian statistical inference with MCMC
-sampling in \emph{Stan} (Stan Development Team, 2023). We used CmdStan
-version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
-2023) to interface with \emph{R} version 4.3.1 (R Core Team, 2023). We
-sampled from posterior distribution from 4 chains with 1000 iterations
-each after 1000 warmup iterations. We estimated parameters and
-confidence intervals as the median and 95\% quantile intervals of the
-posterior, respectively. The key prediction is that
+curves. We fit Bayesian regressions using the \emph{R} package
+\textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling in
+\emph{Stan} (Stan Development Team, 2023). We used CmdStan version
+2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
+interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
+the posterior distribution from 4 chains with 1000 iterations each after
+1000 warmup iterations. We estimated parameters and confidence intervals
+as the median and 95\% quantile intervals of the posterior,
+respectively. The key prediction is that
 \(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).
@@
 coastal and montane populations and among individuals within populations
-using a Bayesian mixed-effects model with site-type as a fixed effect
-and individual plant and site as random effects. We used the \emph{R}
+using a Bayesian mixed-effects model with habitat as a fixed effect and
+individual plant and site as random effects. We used the \emph{R}
 package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
-\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.32.2
-and \textbf{cmdstanr} version 0.5.3 (Gabry and Češnovar, 2023). We
+\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
+and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
 sampled the posterior distribution from 4 chains with 1000 iterations
@@
 single individual per population. We used this approach most of the
-variation in leaf thickness was among sites (see Results) and the plant
-selected for gas exchange measurements was not always among the plants
-randomly selected for leaf thickness, precluding individual level
-correlation. We propagated uncertainty about in \(\mathrm{AA}\) and leaf
-thickness by integrating over the entire posterior distribution sample
-for each variable. The key prediction is that the effect of leaf
-thickness on \(\mathrm{AA}\) is positive.
+variation in leaf thickness was among sites and the plant selected for
+gas exchange measurements was not always among the plants randomly
+selected for leaf thickness, precluding individual level correlation. We
+propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness by
+integrating over the entire posterior distribution sample for each
+variable. The key prediction is that the effect of leaf thickness on
+\(\mathrm{AA}\) is positive.
 
@@
 overdispersion in counts relative to a Poisson model. For all traits,
-the explanatory variables were site-type as a fixed effect and leaf
-within individual plant, individual plant, and site as random effects.
-We used the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner,
-2017) to fit the model in \emph{Stan} (Stan Development Team, 2023) with
-CmdStan version 2.32.2 and \textbf{cmdstanr} version 0.5.3 (Gabry and
-Češnovar, 2023). We interpolated missing adaxial guard cell lengths from
-6 out of 185 samples with zero adaxial stomata using the ``mi'' function
-in \textbf{brms} package. We sampled the posterior distribution from 4
+the explanatory variables were habitat as a fixed effect and leaf within
+individual plant, individual plant, and site as random effects. We used
+the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
+the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
+version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
+2023). We interpolated missing adaxial guard cell lengths from 6 out of
+185 samples with zero adaxial stomata using the ``mi'' function in
+\textbf{brms} package. We sampled the posterior distribution from 4
 chains with 1000 iterations each after 1000 warmup iterations. From each
@@
 curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
-leaves is \(0.12~[\numrange{0.078}{0.15}]\) and
-\(0.027~[\numrange{-0.0033}{0.057}]\), respectively; the difference in
+leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
+\(0.027~[\numrange{-0.0044}{0.057}]\), respectively; the difference in
 average \(\mathrm{AA}\) between habitat types is
-\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.04}{0.14}]\).
-The difference between coastal and montane leaves is lower,
-\(0.054~[\numrange{-0.012}{0.11}]\), but still greater than 0 if we
-remove two sites where we extrapolated slightly beyond fitted
-\(A \textendash g_\mathrm{sw}\) curves. Maximum photosynthetic rate was
+\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.089~[\numrange{0.04}{0.14}]\).
+Posterior predictions closely match observed values of \(A\)
+(\autoref{fig:pp-licor}), indicating an adequate model fit from which it
+we can interpolate between measurements reliably. It also suggests that
+slight extrapolation beyond the data should be reliable, but this is
+less certain. When we remove two leaves where extrapolated slightly
+beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
+\(\mathrm{AA}_\text{coastal}\) is still positive,
+\(0.081~[\numrange{0.022}{0.14}]\), but the difference between coastal
+and montane leaves is smaller, \(0.053~[\numrange{-0.011}{0.12}]\), and
+confidence intervals slightly overlap 0. Maximum photosynthetic rate was
 slightly, but not significantly higher in coastal leaves (Welch Two
@@
 thicker than their montane counterparts. Although coastal leaves are
-thicker and have greater \(\mathrm{AA}\), the relationship between leaf
-thickness and \(\mathrm{AA}\) within habitats is actually slightly
-negative (\autoref{fig:traits-aa}A;
-\(\text{slope}~= -0.11~[\numrange{-0.28}{0.033}]\)), the opposite
-direction our hypothesis predicts.
+thicker and have greater \(\mathrm{AA}\), there is little relationship
+between leaf thickness and \(\mathrm{AA}\) within habitats
+(\autoref{fig:traits-aa}A;
+\(\text{slope}~= -0.11~[\numrange{-0.27}{0.036}]\)).
 
@@
 water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
-\(0.057~[\numrange{-0.16}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
+\(0.059~[\numrange{-0.14}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
 montane leaves, but the 95\% confidence intervals overlap 0 difference.
@@
 (\autoref{fig:traits-aa}B;
-\(\text{slope}~= 0.11~[\numrange{-0.032}{0.25}]\)) in our sample.
+\(\text{slope}~= 0.14~[\numrange{-0.06}{0.35}]\)) in our sample.
 
@@
   \includegraphics{../figures/traits-aa.pdf}
-  \caption{Relationships between leaf amphistomy advantage, (A) $g_\mathrm{smax,ratio}$ and (B) leaf thickness among ʻilima (*Sida fallax*) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi., and amphistomy advantage.
+  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
 A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
@@
 amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
-This allows us to quantity the `amphistomy advantage' (\(\mathrm{AA}\))
+This allows us to quantity the amphistomy advantage (\(\mathrm{AA}\))
 holding all else constant. Taking advantage of the of the steep climatic
-gradients in the Hawaiian archipegalo, we applied this new method to
-show for the first time that sun leaves benefit more from amphistomy
-than shade leaves in ʻilima (\emph{Sida fallax}). While this result has
-not yet been validated in other speciesthere are limits to how much we
-can generalize to all land plants, our results indicate that part of the
-reason amphistomatous leaves are found most commonly in high light
+gradients in the Hawaiian archipelago, we applied this new method to
+show for the first time that sun leaves benefit 4 times more from
+amphistomy than shade leaves on ʻilima (\emph{Sida fallax}) plants
+(\(\mathrm{AA}_\text{coastal} = 0.12\)
+vs.~\(\mathrm{AA}_\text{montane} = 0.027\)). While this result has not
+yet been validated in other species, our results indicate that part of
+the reason amphistomatous leaves are found most commonly in high light
 habitats is that the adaptive benefit is greater in this environment.
@@
 leaves on Oʻahu had high \(g_\mathrm{smax,ratio}\), but similarly low
-\(\mathrm{AA}\) (\autoref{fig:traits-AA}B). Among coastal sites, the
+\(\mathrm{AA}\) (\autoref{fig:traits-aa}B). Among coastal sites, the
 site with the lowest \(g_\mathrm{smax,ratio}\) had the lowest
@@
 
-Two limitations of this study are the small sample sample sizes and
-experimental design that precludes distinguishing genetic from
-environmental differences in leaf traits. Understanding the mechanistic
-basis of higher \(\mathrm{A}A\) in sun leaves would require much larger
-sample sizes. Sun leaves tend to be thicker, more densely packed with
-mesophyll cells, and have greater photosynthetic capacity and higher
-stomatal conductance, among other traits (Lambers et al., 2008). Each of
-these factors and others potentially modulate \(\mathrm{AA}\).
-Quantifying the contribution of all these factors requires larger
-samples and additional measurements that are beyond the scope of this
-study, but exciting avenues for future research on leaf
-structure-function relations. Although many morphological traits that
-distinguish coastal and montane ʻilima populations persist in a common
-environment (Yorkston and Daehler, 2006), we cannot distinguish between
-genetic effects and plastic responses to habitat as causes of difference
-in \(\mathrm{AA}\) because we measured naturally occurring plants
-\emph{in situ}. While disentangling genetic and plastic contributions is
-not necessarily important for understanding the distribution of
-amphistomatous leaves, it would be useful understand genetic and
-environmental contributions to trait variation. A reciprocal transplant
-would be able to determine the genetic and environmental contributions,
-as well their interaction, to trait variance in nature.
+Two limitations of this study are the small sample size and experimental
+design that precludes distinguishing genetic from environmental
+differences in leaf traits. Understanding the mechanistic basis of
+higher \(\mathrm{AA}\) in sun leaves would require much larger sample
+sizes. Sun leaves tend to be thicker, more densely packed with mesophyll
+cells, and have greater photosynthetic capacity and higher stomatal
+conductance, among other traits (Lambers et al., 2008). Each of these
+factors and others potentially modulate \(\mathrm{AA}\). Quantifying the
+contribution of all these factors requires larger samples and additional
+measurements that are beyond the scope of this study, but exciting
+avenues for future research on leaf structure-function relations.
+Although many morphological traits that distinguish coastal and montane
+ʻilima populations persist in a common environment (Yorkston and
+Daehler, 2006), we cannot distinguish between genetic effects and
+plastic responses to habitat as causes of difference in \(\mathrm{AA}\)
+because we measured naturally occurring plants \emph{in situ}. While
+disentangling genetic and plastic contributions is not necessarily
+important for understanding the distribution of amphistomatous leaves,
+it would be insightful to know about genetic and environmental
+contributions to trait variation. A reciprocal transplant would be able
+to determine the genetic and environmental contributions, as well their
+interaction, to trait variance in nature.
 
@@
 with a DOI and stable URL upon publication. Raw data will be deposited
-on Dryad with a DOI and stable URL upon publication.
+on Dryad with a DOI and stable URL upon publication. {[}THE GITHUB REPO
+AND DRYAD DATA ARE AVAILABLE TO REVIEWERS{]}
 
@@
 Additional supporting information may be found online in the Supporting
-Information section at the end of the article
+Information section at the end of the article.
 
@@
   \caption{}
+  \label{fig:pp-licor}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
   \label{fig:habitat-Ags}
@@
 \end{table}
+
+\newpage
 

### 2023-09-07 ms/si.qmd
final checks before submission

@@
 In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap.
-$\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:licor}
+\end{figure}
+```
+
+\newpage
+
+```{=tex}
+\begin{figure}
+  \includegraphics{../figures/pp-licor.pdf}
+  \caption{Posterior predictions (thin grey lines) from fitted $A \textendash g_\mathrm{sw}$ curves closely match the observed distribution (thick black line), indicating that the statistical model adequately captures variation in the response variable over the measured range.
+Symbols: $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+  \label{fig:pp-licor}
 \end{figure}
@@
 Smaller points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants, one per site.
-$A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:habitat-Ags}
@@
 Smaller, transparent points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants.
-$g_\mathrm{smax,ratio}$, the ratio of anatomical maximum stomatal conductance to water vapor on the the adaxial surface over the total.}
+Symbols: $g_\mathrm{smax,ratio}$, the ratio of anatomical maximum stomatal conductance to water vapor on the the adaxial surface over the total.}
   \label{fig:habitat-gmaxratio}

### 2023-09-07 ms/si.tex
final checks before submission

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 \begin{figure}
-  \caption{(Contunited from previous page) Individual $A \textendash g_\mathrm{sw}$ curves used to estimate $\mathrm{AA}$.
+  \caption{(Continued from previous page.) Individual $A \textendash g_\mathrm{sw}$ curves used to estimate $\mathrm{AA}$.
 For each leaf, one per site, we measured $A$ ($y$-axis) over a range of $g_\mathrm{sw}$ ($x$-axis) on the same leaf with two treatments: `amphi' (circles, solid line) leaves were untreated; `pseudohypo' (squares, dashed line) leaves had no conductance through the upper (adaxial) surface.
@@
 In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap.
-$\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:licor}
+\end{figure}
+
+\newpage
+
+\begin{figure}
+  \includegraphics{../figures/pp-licor.pdf}
+  \caption{Posterior predictions (thin grey lines) from fitted $A \textendash g_\mathrm{sw}$ curves closely match the observed distribution (thick black line), indicating that the statistical model adequately captures variation in the response variable over the measured range.
+Symbols: $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+  \label{fig:pp-licor}
 \end{figure}

### 2023-09-07 r/15_archive-data.R

@@ -1,85 +1,0 @@
-source("r/header.R")
-
-trait = read_rds("processed-data/trait.rds")
-site_Ags = read_rds("objects/site_Ags.rds")
-
-# Individual level data
-ind_data = trait |> 
-  mutate(
-    genus = "Sida", species = "fallax", authority = "Walp.",
-    exposition = "natural environment", plant_maturity = "mature",
-    leaf_age = "mature"
-  ) |>
-  select(
-    genus,
-    species,
-    authority,
-    exposition,
-    plant_maturity,
-    leaf_age,
-    site,
-    site_code,
-    site_type,
-    island,
-    latitude_degree,
-    longitude_degree,
-    elevation_m_site,
-    elevation_m_plant,
-    date_sampled,
-    plant_id = individual,
-    leaf_id = leaf_number,
-    lower_number_of_stomata,
-    upper_number_of_stomata,
-    lower_stomatal_density_mm2,
-    upper_stomatal_density_mm2,
-    lower_gcl_um = lower_length_um,
-    upper_gcl_um = upper_length_um,
-    leaf_thickness_um,
-    licor_leaf
-  ) |>
-    left_join(
-      site_Ags |>
-        select(site_code, Tleaf, name, value) |>
-        pivot_wider() |>
-        mutate(licor_leaf = TRUE),
-      by = join_by(site_code, licor_leaf)
-    ) |>
-  select(-licor_leaf)
-
-# Summary table
-ind_data |>
-  filter(site_code == "khkp") |>
-  select(site, A, gsw)
-sum_data = ind_data |>
-  select(site, site_type, island, plant_id, leaf_id, lower_stomatal_density_mm2,
-         upper_stomatal_density_mm2, lower_gcl_um, upper_gcl_um,
-         leaf_thickness_um, A, gsw) |>
-  summarise(across(lower_stomatal_density_mm2:gsw, \(x) mean(x, na.rm = TRUE)),
-            .by = c("site", "site_type", "island", "plant_id")) |>
-  summarise(across(lower_stomatal_density_mm2:gsw, \(x) mean(x, na.rm = TRUE)),
-            .by = c("site", "site_type", "island")) |>
-  arrange(site_type, island, site) |>
-  mutate(Island = case_when(
-    island == "hawaii" ~ "Hawaiʻi",
[…]

### 2023-09-07 r/15_plot-pp-licor.R

@@ -1,0 +1,19 @@
+source("r/header.R")
+
+fit_licor = read_rds("objects/fit_licor.rds")
+
+pp_licor = pp_check(fit_licor, ndraws = 4e3)
+
+pp_licor +
+  xlab(expression(paste(italic(A), ' [', mu, 'mol ', m^-2, ' ', s^-1, ']'))) +
+  ylab("probability density") +
+  scale_color_manual(values = c("black", "grey"),
+                     name = NULL, labels = c("data", "posterior\npredictions")) +
+  theme(
+    axis.text.y = element_text(),
+    axis.ticks.y = element_line(),
+    axis.title.y = element_text(),
+    legend.position = "top"
+  )
+
+ggsave("figures/pp-licor.pdf", width = 4, height = 4)

### 2023-09-07 r/16_archive-data.R

@@ -1,0 +1,85 @@
+source("r/header.R")
+
+trait = read_rds("processed-data/trait.rds")
+site_Ags = read_rds("objects/site_Ags.rds")
+
+# Individual level data
+ind_data = trait |> 
+  mutate(
+    genus = "Sida", species = "fallax", authority = "Walp.",
+    exposition = "natural environment", plant_maturity = "mature",
+    leaf_age = "mature"
+  ) |>
+  select(
+    genus,
+    species,
+    authority,
+    exposition,
+    plant_maturity,
+    leaf_age,
+    site,
+    site_code,
+    site_type,
+    island,
+    latitude_degree,
+    longitude_degree,
+    elevation_m_site,
+    elevation_m_plant,
+    date_sampled,
+    plant_id = individual,
+    leaf_id = leaf_number,
+    lower_number_of_stomata,
+    upper_number_of_stomata,
+    lower_stomatal_density_mm2,
+    upper_stomatal_density_mm2,
+    lower_gcl_um = lower_length_um,
+    upper_gcl_um = upper_length_um,
+    leaf_thickness_um,
+    licor_leaf
+  ) |>
+    left_join(
+      site_Ags |>
+        select(site_code, Tleaf, name, value) |>
+        pivot_wider() |>
+        mutate(licor_leaf = TRUE),
+      by = join_by(site_code, licor_leaf)
+    ) |>
+  select(-licor_leaf)
+
+# Summary table
+ind_data |>
+  filter(site_code == "khkp") |>
+  select(site, A, gsw)
+sum_data = ind_data |>
+  select(site, site_type, island, plant_id, leaf_id, lower_stomatal_density_mm2,
+         upper_stomatal_density_mm2, lower_gcl_um, upper_gcl_um,
+         leaf_thickness_um, A, gsw) |>
+  summarise(across(lower_stomatal_density_mm2:gsw, \(x) mean(x, na.rm = TRUE)),
+            .by = c("site", "site_type", "island", "plant_id")) |>
+  summarise(across(lower_stomatal_density_mm2:gsw, \(x) mean(x, na.rm = TRUE)),
+            .by = c("site", "site_type", "island")) |>
+  arrange(site_type, island, site) |>
+  mutate(Island = case_when(
+    island == "hawaii" ~ "Hawaiʻi",
[…]

### 2023-09-07 r/install-packages.R

@@ -11,6 +11,7 @@ install.packages("gridExtra")
 install.packages("magrittr")
 install.packages("photosynthesis")
 install.packages("purrr")
+install.packages("raster")
 install.packages("readr")
 install.packages("stringr")
 install.packages("tidybayes")

### 2023-09-07 ms/ms.tex
ajb submission

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 Posterior predictions closely match observed values of \(A\)
-(\autoref{fig:pp-licor}), indicating an adequate model fit from which it
-we can interpolate between measurements reliably. It also suggests that
+(\autoref{fig:pp-licor}), indicating an adequate model fit from which we
+can interpolate between measurements reliably. It also suggests that
 slight extrapolation beyond the data should be reliable, but this is

### 2023-09-07 ms/si.tex
ajb submission

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, interior hidden, enhanced, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 

### 2023-10-26 ms/ms.qmd
revising ms

@@
 
-$^2$ School of Life Sciences, University of California, Davis, CA, USA 95616
+$^2$ Department of Plant Sciences, University of California, Davis, CA, USA 95616
 
@@
 \textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts.
-Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.
+Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most of the conductance being on the lower (abaxial) surface.
 
@@
 
-**Keywords:** amphstomy, leaf, light, Hawaiʻi, leaf, *Sida fallax*, stomata
+**Keywords:** amphstomy, leaf, light, Hawaiʻi, *Sida fallax*, stomata
 
@@
 Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010].
-A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through evapotranspiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997].
+A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through transpiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997].
 Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021].
@@
 A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019].
+However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2021] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.
 
@@
 #| label: tbl-locations
-#| tbl-cap: "ʻIlima study sites for this study."
+#| tbl-cap: "ʻIlima study site location information."
 #| 
@@
 
-The log-response ratio is easy to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage.
+The log-response ratio is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage.
 The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates.
 The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission.
-Because the propafilm reduces transmission, we used a higher incident $\mathrm{PPFD}$ for pseudohypo leaves (higher by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm).
+To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
 We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
@@
 After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation.
+Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity.
 All other environmental conditions in the leaf chamber remained the same.
@@
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$.
+The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$, meaning the 95\% confidence intervals of $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}$ should be positive and not encompass 0.
 
@@
 We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable.
-The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive.
+The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive, meaning the 95\% confidence interval of the slope should be positive and not encompass 0.
 
@@
 This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant.
-Taking advantage of the of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs.
+Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs.
 $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$).
@@
 Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves.
-The negative relationship between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].
+The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].
 
@@
 This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common.
-By developing a straightforward gas exchange method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant.
+By developing a straightforward experimental method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant.
 In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest.
+This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that be an important consideration in predicting realized photosynthetic rates in nature.
+Rather, our experiments show that coastal amphistomatous leaves with the same total leaf stomatal conductance photosynthesize more than identical hypostomatous leaves.
 We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.

### 2023-10-26 ms/ms.qmd
revising ms

@@
 Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$.
+The airspace conductance is one component of the overall mesophyll conductance, $g_\mathrm{m}$, which is often strongly influenced by the chloroplast surface area exposed to airspace and mesophyll cell wall thickness [@evans_resistances_2009; @gago_photosynthesis_2020; @flexas_cell_2021].
+Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but would still benefit from amphistomy.
 
@@
 We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods).
-One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], resulting in lower $g_\mathrm{ias}$ [@parkhurst_adaptive_1978].
+One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], which will often result in lower $g_\mathrm{ias}$ either by increasing the diffusion path length [@parkhurst_adaptive_1978] or making the airspace less porous.
 A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface.
@@
 For pubescent leaves (usually coastal), we dried and pressed leaves for $\approx 1$ week [@hill_temperature_2014], carefully scraped trichomes off with a razor blade, and rehydrated the leaf.
+Rehydration restores leaf area to its fresh value [@blonder_leaf-area_2012].
 For glabrous leaves, we used fresh leaves.
@@
 
-The log-response ratio is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage.
+The log-response ratio is commonly used social and biological sciences (e.g.
+@hedges_metaanalysis_1999).
+It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage.
 The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates.

### 2023-10-28 ms/ms.qmd
revising and added habitat-climate analysis

@@
 
+# From 16_fit-habitat-climate.R
+df_stat = read_rds("../objects/habitat-climate-stats.rds")
+
 # Calculation amount of gsw extrapolation
@@
 ```{=tex}
+
 \begin{figure}[H]
-  \includegraphics{../figures/study-system.pdf}
-  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
+  \includegraphics[height=3.25in]{../figures/study-system.pdf}
+  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
+  \caption{A.
+Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
+leaves.
+C.
+Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
+D.
+Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
+Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
+The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right).
+MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
   \label{fig:study-system}
 \end{figure}
+
 ```
@@
 
-## Plant sampling --- {#plant-sampling}
+## Plant sampling and climate --- {#plant-sampling}
 
-We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}).
+We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}c; \autoref{tbl-locations}).
 We avoided sites that appeared to be cultivated.
@@
 We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.
+We downloaded climatic data on mean annual temperature, solar radiation, and vegetation height from the Climate and Solar Radiation of Hawaiʻi databases [@giambelluca_evapotranspiration_2014] using the latitude and longitude at the middle of each transect.
+We also downloaded mean annual precipitation from 1978-2007 from the Rainfall Atlas of Hawaiʻi [@giambelluca_online_2013].
+The spatial resolution is approximately $234 \times 250$m.
+The temperature data are calibrated from networks of meteorological stations operating in the late 20th century and 21st century; the solar radiation data are calibrated from satellite measurements collected between 2002 and 2009 [@giambelluca_evapotranspiration_2014].
+We tested whether climatic variables differed among our coastal and montane populations using Welch's two-sample $t$-test.
 
@@
 The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates.
-The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission.
+The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
 To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
@@
 We used these data as our estimates of maximum $A$ and $g_\text{sw}$.
-After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation.
+After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
 Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity.
@@
 We then repeated the process with the other treatment (amphi or pseudohypo).
+We did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).
 
@@
 
+Coastal ʻilima are surrounded by shorter vegetation than their montane counterparts (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Veg'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Veg'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Veg'))[['p']], 3)`$).
+The montane site with the lowest vegetation height is a remnant dry forest (Koaiʻa tree sanctuary) in a matrix of cattle pasture, hence the satellite derived vegetation height may be lower than what existed prior to human disturbance.
+Habitats did not differ significantly in average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$).
+Coastal habitats are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$) but mean annual precipitation is similar (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).
+
 ## Amphistomy advantage is greater in coastal leaves ---
@@
   \includegraphics{../figures/habitat-aa.pdf}
-  \caption{Coastal leaves benefit more amphistomy than montane leaves.
+  \caption{Coastal leaves benefit more from amphistomy than montane leaves.
 A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
@@
 ```
+
 ## Leaf thickness is associated with amphistomy advantage between but not within habitats ---
@@
 $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$).
+Coastal and montane ʻilima leaves are likely good representatives of classic sun and shade leaf syndromes because 1) they vary in traits like reflective pubescence [@ehleringer_pubescence_1978] and leaf thickness [@terashima_why_2001] that typically characterize sun-shade adaptation; and 2) since ʻilima shrubs are typically $<1$m tall, they are shaded by trees in montane, but not coastal habitats (\autoref{fig:study-system}d).
 While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.
@@
 Our results partially support this hypothesis.
-Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}A), opposite our prediction.
+Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}a), opposite our prediction.
 Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats.
@@
 Our results do not support this hypothesis.
-Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-aa}B).
+Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-aa}b).
 Among coastal sites, the site with the lowest $g_\mathrm{smax,ratio}$ had the lowest $\mathrm{AA}$, but there was little variation in $g_\mathrm{smax,ratio}$ among coastal leaves in our sample.
@@
 
-Two limitations of this study are the small sample size and experimental design that precludes distinguishing genetic from environmental differences in leaf traits.
+Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment.
 Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes.
@@
 A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.
+However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature.
+Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.
 

### 2023-10-28 r/10_calculate-gmaxratio.R

@@ -279,3 +279,7 @@ write_rds(gmaxratio_site, "objects/gmaxratio_site.rds")
 write_rds(gmaxratio_ind, "objects/gmaxratio_ind.rds")
 write_rds(gmaxratio_leaf, "objects/gmaxratio_leaf.rds")
 
+write_rds(stomata_sitetype, "objects/stomata_sitetype.rds")
+write_rds(stomata_site, "objects/stomata_site.rds")
+write_rds(stomata_ind, "objects/stomata_ind.rds")
+write_rds(stomata_leaf, "objects/stomata_leaf.rds")

### 2023-10-28 r/16_fit-habitat-climate.R

@@ -1,0 +1,41 @@
+source("r/header.R")
+
+trait = read_rds("processed-data/trait.rds") |>
+  summarize(
+    site_type = first(site_type),
+    `MAT~group('[', paste(phantom()*degree, 'C'), ']')` = first(tair_ann),
+    `MAP~group('[', mm, ']')` = first(rf_ann),                
+    `Solar~radiation~group('[', W~m^-2, ']')` = first(cl_sw_ann),
+    `Vegetation~height~group('[', m, ']')` = first(veg_ht_ann),
+    .by = "site_code"
+  ) |>
+  pivot_longer(`MAT~group('[', paste(phantom()*degree, 'C'), ']')`:`Vegetation~height~group('[', m, ']')`) |>
+  mutate(habitat = factor(site_type, levels = c("montane", "coastal"))) |>
+  select(-site_type)
+
+df_stat = trait |>
+  group_by(name) |>
+  t_test(value ~ habitat, var.equal = FALSE) |>
+  add_significance() |>
+  full_join(
+    trait |>
+      summarize(y.position = max(value) * 1.1, .by = "name") |>
+      mutate(xmin = 1, xmax = 2),
+    by = join_by(name)
+  )
+
+gp = trait |>
+  ggplot(aes(habitat, value, color = habitat)) +
+  facet_wrap(. ~ name, scales = "free_y", labeller = label_parsed) +
+  geom_jitter(width = 0.1, pch = 21) +
+  stat_summary(position = position_nudge(x = 0.25), fun.data = mean_se) +
+  stat_pvalue_manual(df_stat) +
+  scale_color_manual(values = c("steelblue", "tomato")) +
+  scale_y_continuous(expand = expansion(mult = 0.1)) +
+  theme(legend.position = "none") 
+
+plot_grid(gp, labels = "D")
+
+ggsave("figures/habitat-climate.pdf", width = 5, height = 5)
+
+write_rds(df_stat, "objects/habitat-climate-stats.rds")

### 2023-10-28 r/header.R

@@ -8,12 +8,14 @@ library(forcats)
 library(ggdist)
 library(ggplot2)
 library(ggpp)
+library(ggpubr)
 library(glue)
 library(gridExtra)
 library(magrittr)
 library(photosynthesis)
 library(purrr)
 library(readr)
+library(rstatix)
 library(stringr)
 library(tidybayes)
 library(tidyr)

### 2023-10-28 r/install-packages.R

@@ -6,6 +6,7 @@ install.packages("forcats")
 install.packages("ggdist")
 install.packages("ggplot2")
 install.packages("ggpp")
+install.packages("ggpubr")
 install.packages("glue")
 install.packages("gridExtra")
 install.packages("magrittr")
@@ -13,6 +14,7 @@ install.packages("photosynthesis")
 install.packages("purrr")
 install.packages("raster")
 install.packages("readr")
+install.packages("rstatix")
 install.packages("stringr")
 install.packages("tidybayes")
 install.packages("tidyr")

### 2023-10-29 ms/ms.qmd
added ags-curve example figure

@@
 We sectioned the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.
+Leaf thickness is measured as the length from upper cuticle to lower cuticle.
 
@@
 We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$.
+The midday irradiance in coastal ʻilima typically meets or even exceeds a PPFD of $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ and previous experiments with sun leaves revealed that $2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$ is always at or near saturating irradiance.
+Even though lower irradiance may be saturating for montane leaves, we used this higher value for all leaves to standardize conditions.
+
 We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification.
-For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values.
+For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values (see below for details).
 We refer to this as an $A \textendash g_\text{sw}$ curve.
@@
 
-To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements.
-In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum.
+\autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves.
+We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve.
+To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements.
+In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum.
 We used these data as our estimates of maximum $A$ and $g_\text{sw}$.
@@
 We logged data until $g_\text{sw}$ reached its nadir.
-We then repeated the process with the other treatment (amphi or pseudohypo).
+We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo).
 We did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).
@@
 We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
-We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations.
+We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations per chain.
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
@@
 We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023].
-We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations.
+We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations per chain.
 We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population.
-We used this approach most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation.
-We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable.
+We used this approach because most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation.
+We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness estimates by integrating over the entire posterior distribution sample for each variable.
 The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive, meaning the 95\% confidence interval of the slope should be positive and not encompass 0.
@@
 We interpolated missing adaxial guard cell lengths from `r length(which(fit_stomata$data$upper_number_of_stomata == 0))` out of `r nrow(fit_stomata$data)` samples with zero adaxial stomata using the "mi" function in **brms** package.
-We sampled the posterior distribution from `r dim_fit_stomata[2]` chains with `r dim_fit_stomata[1]` iterations each after `r dim_fit_stomata[1]` warmup iterations.
+We sampled the posterior distribution from `r dim_fit_stomata[2]` chains with `r dim_fit_stomata[1]` iterations each after `r dim_fit_stomata[1]` warmup iterations per chain.
 From each posterior sample, we calculated $g_{\mathrm{smax,ratio}}$ as:
 
-$$g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},$$ where $g_\mathrm{smax,lower}$ and $g_\mathrm{smax,upper}$ are maximum stomatal conductance to water vapor at $T_\mathrm{leaf} = 25 ^ \circ$ C on the lower and upper surface, respectively.
-The maximum stomatal conductance is calculated from stomatal density and length, assuming that stomata are fully open, following @sack_developmental_2016.
+$$g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},$$
 
-<!-- They defined a biophysical and morphological constant as: -->
+where $g_\mathrm{smax,lower}$ and $g_\mathrm{smax,upper}$ are maximum stomatal conductance to water vapor at $T_\mathrm{leaf} = 25 ^ \circ$ C on the lower and upper surface, respectively.
+The maximum stomatal conductance was calculated from stomatal density and length, assuming that stomata are fully open, following @sack_developmental_2016:
+
+$$g_\mathrm{smax} = bmds^{0.5}.$$
+In this equation, $b$ is a biophysical constant, $m$ is a morphological constant, $d$ is the stomatal density, and $s$ is the stomatal complex area.
+We assume that $b$, which is determined by the molecular species, temperature, and air pressure, is the same for both surfaces; we assume that $m$, which is determined by guard cell allometry is also the same for both surfaces.
+Hence, the $b$ and $m$ constants cancel out of $g_\mathrm{smax,ratio}$ and only density and length ($l$), which is proportional to the square root of area, affect the ratio: $g_\mathrm{smax} \propto dl$.
 
@@
 We estimated the relationship between leaf $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ measured from a single leaf per population.
-We propagated uncertainty about in $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable.
+We propagated uncertainty about $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable.
 The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$.
@@
 Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}).
+Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).
 
@@
 
-<!-- add tom buckley as coauthor or acknowledge help developing method -->
-
-The authors thank Kasey Barton for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
+The authors thank Kasey Barton for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
 
@@
 \setcounter{equation}{0}
+```
+```{=tex}
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:ags-curve}
+\end{figure}
 ```

### 2023-10-29 ms/si.qmd
added ags-curve example figure

@@
 
+```
+
+\newpage
+
+```{=tex}
+\begin{figure}
+  \includegraphics{../figures/ags-curve.pdf}
+  \caption{Idealized method for collecting $A \textendash g_\text{sw}$ curves on either amphi or pseudohypo leaves.
+A) After clamping the leaf into the LI-6800 chamber, it acclimates to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) and humidity ($\mathrm{RH} = 70\%$).
+This induces stomata to open, increasing $g_\mathrm{sw}$ and $A$ until they reach a maximum.
+We abruptly lower the chamber humidity to $\approx 10\%$ to close stomata and log data (black points) until $g_\mathrm{sw}$ and $A$ reach their nadir.
+B) We fit $A \textendash g_\text{sw}$ curves to logged data points.
+The asterisk in both panels indicates the data point used for maximum $A$ and $g_\mathrm{sw}$.
+C) $\mathrm{AA}$ is low (left panel) when the photosynthetic rate of an amphi leaf is similar to a pseudohypo leaf at the same total $g_\mathrm{sw}$ ($x$-axes); large $\mathrm{AA}$ (right panel) is indicated when an amphi leaf has a higher photosynthetic rate than a pseudohypo leaf.
+Abbreviations: $A$ is the photosynthetic rate; $\mathrm{AA}$ is the amphistomy advantage; $g_\mathrm{sw}$ is the stomatal conductance to water vapor; $\mathrm{PPFD}$ is photosynthetic photon flux density; $\mathrm{RH}$ is relative humidity.}
+  \label{fig:ags-curve}
+\end{figure}
 ```

### 2023-10-29 ms/si.tex
added ags-curve example figure

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, interior hidden, enhanced, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \end{landscape}
+
+\newpage
+
+\begin{figure}
+  \includegraphics{../figures/ags-curve.pdf}
+  \caption{Idealized method for collecting $A \textendash g_\text{sw}$ curves on either amphi or pseudohypo leaves.
+A) After clamping the leaf into the LI-6800 chamber, it acclimates to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) and humidity ($\mathrm{RH} = 70\%$).
+This induces stomata to open, increasing $g_\mathrm{sw}$ and $A$ until they reach a maximum.
+We abruptly lower the chamber humidity to $\approx 10\%$ to close stomata and log data (black points) until $g_\mathrm{sw}$ and $A$ reach their nadir.
+B) We fit $A \textendash g_\text{sw}$ curves to logged data points.
+The asterisk in both panels indicates the data point used for maximum $A$ and $g_\mathrm{sw}$.
+C) $\mathrm{AA}$ is low (left panel) when the photosynthetic rate of an amphi leaf is similar to a pseudohypo leaf at the same total $g_\mathrm{sw}$ ($x$-axes); large $\mathrm{AA}$ (right panel) is indicated when an amphi leaf has a higher photosynthetic rate than a pseudohypo leaf.
+Abbreviations: $A$ is the photosynthetic rate; $\mathrm{AA}$ is the amphistomy advantage; $g_\mathrm{sw}$ is the stomatal conductance to water vapor; $\mathrm{PPFD}$ is photosynthetic photon flux density; $\mathrm{RH}$ is relative humidity.}
+  \label{fig:ags-curve}
+\end{figure}
 

### 2023-10-29 r/13_plot-habitat-Ags.R

@@ -27,6 +27,10 @@ ggplot(site_Ags, aes(habitat, value, color = habitat)) +
   scale_color_manual(values = c("steelblue", "tomato"), guide = NULL)
 
 fit_habitat_Ags = site_Ags |>
+  select(site_code, site_type, name, value) |>
+  pivot_wider() |>
+  mutate(wue = A / gsw) |>
+  pivot_longer(A:wue) |>
   split(~ name) |>
   map(\(.x) {
     t.test(value ~ site_type, data = .x)

### 2023-10-29 r/18_plot-ags-curve.R

@@ -1,0 +1,130 @@
+source("r/header.R")
+
+# Example time course and A-gsw curve ----
+x1 = seq(-5, 5, 0.01)
+x2 = seq(5, 15, 0.01)
+
+df = tibble(
+  x = c(x1, x2), 
+  y = c(plogis(x1), 2 * (plogis(rev(x2 - 10))) - 1)
+)
+
+## Time course
+gp1 = ggplot(df, aes(x, y)) +
+  annotate("rect", xmin = 5, xmax = 15, ymin = -Inf, ymax = Inf, color = NA, fill = "grey") +
+  geom_segment(
+    data = filter(df, x %in% 5:15), 
+    mapping = aes(x = x, xend = x, y = y, yend = -Inf),
+    linetype = "dashed"
+  ) +
+  geom_line(linewidth = 1.1) +
+  geom_point(data = filter(df, x %in% 5:15), size = 3) +
+  geom_point(data = filter(df, x == 5), size = 20, shape = "*") +
+  xlab("time") +
+  ylab(expression(paste(italic(g)[sw], " and ", italic(A)))) +
+  annotate("text", x = 0, y = 1.275, label = "Acclimate\nRH = 70%\nstomata open") +
+  annotate("text", x = 10, y = 1.275, label = "Log data\nRH = 10%\nstomata close") +
+  ylim(-1, 1.45) +
+  theme(
+    axis.text = element_blank(),
+    axis.ticks = element_blank()
+  )
+
+## Connecting arrow
+gp2 = ggplot() +
+  annotate("segment", x = 0, xend = 1, y = 0, yend = 0, arrow = arrow(length = unit(0.1, "inches"))) +
+  theme_void()
+
+## A-gsw curve
+gp3 = df |>
+  filter(x %in% 5:15) %>%
+  mutate(y1 = y + rnorm(nrow(.), 0, 0.01)) |>
+  ggplot(aes(y, y1)) +
+  geom_smooth(color = "grey", method = "lm", formula = 'y ~ x') +
+  geom_point(size = 2) +  
+  annotate("point", x = 1, y = 1, size = 15, shape = "*") +
+  xlab(expression(paste(italic(g)[sw], " [mol ", m ^ -2 ~ s ^ -1, "]"))) +
+  ylab(expression(paste(italic(A), " [", mu, "mol ", m ^ -2 ~ s ^ -1, "]"))) +
+  coord_equal() +
+  theme(
+    axis.text = element_blank(),
[…]

### 2023-10-29 ms/ms.qmd
revised ms sent to coauthors

@@
 The airspace conductance is one component of the overall mesophyll conductance, $g_\mathrm{m}$, which is often strongly influenced by the chloroplast surface area exposed to airspace and mesophyll cell wall thickness [@evans_resistances_2009; @gago_photosynthesis_2020; @flexas_cell_2021].
-Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but would still benefit from amphistomy.
+Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but will still benefit from amphistomy as long as $g_\mathrm{ias}$ is finite.
 
@@
 A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019].
-However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2021] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.
+However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2022] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.
 
@@
 We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification.
-For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values (see below for details).
-We refer to this as an $A \textendash g_\text{sw}$ curve.
+For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values.
+We refer to this as an $A \textendash g_\text{sw}$ curve, which is described in more detail below.
 We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
@@
 The log-response ratio is commonly used social and biological sciences (e.g.
-@hedges_metaanalysis_1999).
+@hedges_meta-analysis_1999).
 It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage.
-The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates.
+The log-response ratio is preferable to the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates.
 The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
@@
 To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$.
+Let $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ be the estimated $A$ of the amphi and pseudohypo leaves, respectively.
+We estimated these quantities at the same $g_\mathrm{sw}$ using fitted parameters ($\hat{\beta}$'s):
+
+$$\hat{A}_\text{amphi} = \hat{\beta}_{0,\text{amphi}} + \hat{\beta}_{1,\text{amphi}} \times \text{log}(g_\mathrm{sw})$$
+$$\hat{A}_\text{hypo} = \hat{\beta}_{0,\text{hypo}} + \hat{\beta}_{1,\text{hypo}} \times \text{log}(g_\mathrm{sw})$$
+
 In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e.
 the curves overlapped for a range of $g_\text{sw}$ values).
-In those cases, we estimated $\mathrm{AA}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap.
+In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves.
 In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve.
 In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves.
-We fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
+The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf.
+$\mathrm{AA}$ for each leaf was calculated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ using the log-response ratio shown above.
+
+To estimate $\hat{\beta}$'s from the $A \textendash g_\text{sw}$ curve for each leaf, we fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
 We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
@@
 We propagated uncertainty about $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable.
-The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$.
+The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$, meaning the 95\% confidence interval of the slope should be positive and not encompass 0 in the domain $g_\mathrm{smax,ratio} < 0.5$.
 
@@
 In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest.
-This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that be an important consideration in predicting realized photosynthetic rates in nature.
+This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that likely influences realized photosynthetic rates in natural populations.
 Rather, our experiments show that coastal amphistomatous leaves with the same total leaf stomatal conductance photosynthesize more than identical hypostomatous leaves.

### 2023-10-29 ms/ms.tex
revised ms sent to coauthors

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\(^2\) School of Life Sciences, University of California, Davis, CA, USA
-95616
+\(^2\) Department of Plant Sciences, University of California, Davis,
+CA, USA 95616
 
@@
 equivocal with respect to two hypotheses -- that coastal leaves benefit
-more because they are thicker and therefore have lower CO\(_2\)
-conductance through the internal airspace, and that leaves with more
-similar conductances at each surface benefit more than leaves with few
-stomata on the upper surface.
+more because 1) they are thicker and therefore have lower CO\(_2\)
+conductance through the internal airspace, and 2) that they benefit more
+because they have similar conductance on each surface, as opposed to
+most of the conductance being on the lower (abaxial) surface.
 
@@
 
-\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, leaf, \emph{Sida
-fallax}, stomata
+\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, \emph{Sida fallax},
+stomata
 
@@
 stomata mediate the relationship between CO\(_2\) gained through
-photosynthesis and water lost through evapotranspiration is essential to
+photosynthesis and water lost through transpiration is essential to
 predict future climate using Earth Systems Models (Jarvis, 1976; Ball et
@@
 largest photosynthetic benefit in leaves with intrinsically low
-\(g_\mathrm{ias}\).
+\(g_\mathrm{ias}\). The airspace conductance is one component of the
+overall mesophyll conductance, \(g_\mathrm{m}\), which is often strongly
+influenced by the chloroplast surface area exposed to airspace and
+mesophyll cell wall thickness (Evans et al., 2009; Gago et al., 2020;
+Flexas et al., 2021). Hence, thicker leaves may compensate for lower
+\(g_\mathrm{ias}\) through increased chloroplast surface area exposed to
+airspace (Terashima et al., 2006), but will still benefit from
+amphistomy as long as \(g_\mathrm{ias}\) is finite.
 
@@
 carbon gain should be enough to offset this cost under most realistic
-conditions (Muir, 2019).
+conditions (Muir, 2019). However, if minimal stomatal conductance is
+related to stomatal density (Drake et al., 2013; Márquez et al., 2022)
+and the upper boundary layer conductance is higher, then amphistomy
+could cause additional, unavoidable water loss.
 
@@
 have greater \(\mathrm{AA}\) is that they are usually thicker or denser
-(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
-(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
-leaves have a stomatal density ratio closer to 0.5 (same density on each
-leaf surface), this will confer a greater advantage than an
-amphistomatous leaf with most stomata on one surface. In other words,
-amphistomy doesn't make much difference if one leaf surface has few open
-stomata on it. We therefore predict that sun leaves will have greater
-\(\mathrm{AA}\) possibly because they have thicker leaves and/or
-\(\mathrm{SR}\) closer to 0.5. We actually report
+(Poorter et al., 2019), which will often result in lower
+\(g_\mathrm{ias}\) either by increasing the diffusion path length
+(Parkhurst, 1978) or making the airspace less porous. A nonmutually
+exclusive hypothesis is that if sun leaves have a stomatal density ratio
+closer to 0.5 (same density on each leaf surface), this will confer a
+greater advantage than an amphistomatous leaf with most stomata on one
+surface. In other words, amphistomy doesn't make much difference if one
+leaf surface has few open stomata on it. We therefore predict that sun
+leaves will have greater \(\mathrm{AA}\) possibly because they have
+thicker leaves and/or \(\mathrm{SR}\) closer to 0.5. We actually report
 \(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
@@
 \begin{figure}[H]
-  \includegraphics{../figures/study-system.pdf}
-  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
+  \includegraphics[height=3.25in]{../figures/study-system.pdf}
+  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
+  \caption{A.
+Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
+leaves.
+C.
+Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
+D.
+Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
+Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
+The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right).
+MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
   \label{fig:study-system}
@@
 \hypertarget{plant-sampling}{%
-\subsection{Plant sampling ---}\label{plant-sampling}}
+\subsection{Plant sampling and climate ---}\label{plant-sampling}}
 
@@
 scientist records on iNaturalist (Anon, 2022)
-(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
+(\autoref{fig:study-system}c; \autoref{tbl-locations}). We avoided sites
 that appeared to be cultivated. We visited sites between August and
@@
 leaf from a single plant at the middle of each transect for gas exchange
-measurements.
+measurements. We downloaded climatic data on mean annual temperature,
+solar radiation, and vegetation height from the Climate and Solar
+Radiation of Hawaiʻi databases (Giambelluca et al., 2014) using the
+latitude and longitude at the middle of each transect. We also
+downloaded mean annual precipitation from 1978-2007 from the Rainfall
+Atlas of Hawaiʻi (Giambelluca et al., 2013). The spatial resolution is
+approximately \(234 \times 250\)m. The temperature data are calibrated
+from networks of meteorological stations operating in the late 20th
+century and 21st century; the solar radiation data are calibrated from
+satellite measurements collected between 2002 and 2009 (Giambelluca et
+al., 2014). We tested whether climatic variables differed among our
+coastal and montane populations using Welch's two-sample \(t\)-test.
 
@@
 \begin{table}
-\caption{\label{tbl-locations}ʻIlima study sites for this study. }\tabularnewline
+\caption{\label{tbl-locations}ʻIlima study site location information. }\tabularnewline
 
@@
 carefully scraped trichomes off with a razor blade, and rehydrated the
-leaf. For glabrous leaves, we used fresh leaves. We applied clear nail
+leaf. Rehydration restores leaf area to its fresh value (Blonder et al.,
+2012). For glabrous leaves, we used fresh leaves. We applied clear nail
 polish to both leaf surfaces of fresh or rehydrated leaves in the middle
@@
 and took digital micrographs using a brightfield microscope, as
-described above.
+described above. Leaf thickness is measured as the length from upper
+cuticle to lower cuticle.
 
@@
 ambient CO\(_2\) (415 ppm), and
-\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
-`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
-but with modification. For each leaf, we measured the photosynthetic
-rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
-range of \(g_\text{sw}\) values. We refer to this as an
-\(A \textendash g_\text{sw}\) curve. We compared the
+\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). The midday
+irradiance in coastal ʻilima typically meets or even exceeds a PPFD of
+\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) and previous
+experiments with sun leaves revealed that
+\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) is always at or near
+saturating irradiance. Even though lower irradiance may be saturating
+for montane leaves, we used this higher value for all leaves to
+standardize conditions.
+
+We also estimated `amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu}
+Parkhurst (1978), but with modification. For each leaf, we measured the
+photosynthetic rate of an untreated amphistomatous leaf
+(\(A_{\mathrm{amphi}}\)) over a range of \(g_\text{sw}\) values. We
+refer to this as an \(A \textendash g_\text{sw}\) curve, which is
+described in more detail below. We compared the
 \(A \textendash g_\text{sw}\) curve of the untreated leaf to the
@@
 
-The log-response ratio is easy to interpret because values above 0
-indicate a photosynthetic advantage of amphistomy, whereas values less
-than 0 indicate a disadvantage. The response ratio is also better than
-the absolute difference because it indicates a proportional change in
-\(A\), which facilitates comparisons across leaves and environments with
-different baseline photosynthetic rates. The irradiance of the light
-source in the pseudohypo leaf is higher because the propafilm reduces
-transmission. Because the propafilm reduces transmission, we used a
-higher incident \(\mathrm{PPFD}\) for pseudohypo leaves (higher by a
-factor 1/0.91, the inverse of the measured transmissivity of the
-propafilm). We also set the stomatal conductance ratio, for purposes of
-calculating boundary layer conductance, to 0 for pseudohypo leaves
+The log-response ratio is commonly used social and biological sciences
+(e.g. Hedges et al. (1999)). It is straightforward to interpret because
+values above 0 indicate a photosynthetic advantage of amphistomy,
+whereas values less than 0 indicate a disadvantage. The log-response
+ratio is preferable to the absolute difference because it indicates a
+proportional change in \(A\), which facilitates comparisons across
+leaves and environments with different baseline photosynthetic rates.
+The irradiance of the light source in the pseudohypo leaf was higher
+because the propafilm reduces transmission. To compensate for reduced
+transmission, we increased incident \(\mathrm{PPFD}\) for pseudohypo
+leaves by a factor 1/0.91, the inverse of the measured transmissivity of
+the propafilm. We also set the stomatal conductance ratio, for purposes
+of calculating boundary layer conductance, to 0 for pseudohypo leaves
 following manufacturer directions.
 
-To control for order effects, we alternated between starting with amphi
-(untreated) or pseudohypo (treated) leaf measurements. In the field, we
-acclimated the focal leaf to high light
-(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
-high relative humidity (\(\mathrm{RH} = 70\%\)),
-\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
-until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
-as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
-decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
-closure without photosynthetic downregulation. All other environmental
-conditions in the leaf chamber remained the same. We logged data until
-\(g_\text{sw}\) reached its nadir. We then repeated the process with the
-other treatment (amphi or pseudohypo).
+\autoref{fig:ags-curve} illustrates our method for collecting
+\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
+an amphi (untreated) curve and a pseudohypo (treated) curve. To control
+for order effects, we alternated between starting with amphi or
+pseudohypo leaf measurements. In the field, we acclimated the focal leaf
+to saturating light and high relative humidity (\(\mathrm{RH} = 70\%\)),
+as described above, until \(A\) and \(g_\text{sw}\) reach their maximum.
+We used these data as our estimates of maximum \(A\) and
+\(g_\text{sw}\). After that, we decreased \(\mathrm{RH}\) to
+\(\approx 10\%\) to induce rapid stomatal closure without biochemical
+downregulation. Hence, \(A_\text{amphi}\) and \(A_\text{hypo}\) were
+both measured at low chamber humidity after the leaf had acclimated to
+high humidity. All other environmental conditions in the leaf chamber
+remained the same. We logged data until \(g_\text{sw}\) reached its
+nadir. We then repeated the process of acclimating the leaf to 70\%
+\(\mathrm{RH}\) and inducing stomatal closure with low \(\mathrm{RH}\)
+with the other treatment (amphi or pseudohypo). We did not detect an
+effect of treatment order on \(\mathrm{AA}\) (results not shown).
 
@@
 \(\text{log}(g_\mathrm{sw})\) on \(A\) to interpolate modeled \(A\) for
-amphi and pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12
-leaves, the minimum \(g_\text{sw}\) of the amphi curve was smaller than
-the maximum \(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves
-overlapped for a range of \(g_\text{sw}\) values). In those cases, we
-estimated \(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle
-of the range of overlap. In 2 of 12 leaves, the
+amphi and pseudohypo leaves at the same \(g_\text{sw}\). Let
+\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) be the estimated
+\(A\) of the amphi and pseudohypo leaves, respectively. We estimated
+these quantities at the same \(g_\mathrm{sw}\) using fitted parameters
+(\(\hat{\beta}\)'s):
+
+\[\hat{A}_\text{amphi} = \hat{\beta}_{0,\text{amphi}} + \hat{\beta}_{1,\text{amphi}} \times \text{log}(g_\mathrm{sw})\]
+\[\hat{A}_\text{hypo} = \hat{\beta}_{0,\text{hypo}} + \hat{\beta}_{1,\text{hypo}} \times \text{log}(g_\mathrm{sw})\]
+
+In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
+smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
+(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
+those cases, we estimated \(\hat{A}_\text{amphi}\) and
+\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
+the range of overlap between the curves. In 2 of 12 leaves, the
 \(A \textendash g_\text{sw}\) curves did not quite overlap because the
@@
 beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
-curves. We fit Bayesian regressions using the \emph{R} package
-\textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling in
-\emph{Stan} (Stan Development Team, 2023). We used CmdStan version
+curves. The vertical lines in \autoref{fig:licor} show the
+\(g_\text{sw}\) for each leaf. \(\mathrm{AA}\) for each leaf was
+calculated \(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) using
+the log-response ratio shown above.
+
+To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
+curve for each leaf, we fit Bayesian regressions using the \emph{R}
+package \textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling
+in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
@@
 the posterior distribution from 4 chains with 1000 iterations each after
-1000 warmup iterations. We estimated parameters and confidence intervals
-as the median and 95\% quantile intervals of the posterior,
+1000 warmup iterations per chain. We estimated parameters and confidence
+intervals as the median and 95\% quantile intervals of the posterior,
 respectively. The key prediction is that
-\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).
+\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
+95\% confidence intervals of
+\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
+positive and not encompass 0.
 
@@
 sampled the posterior distribution from 4 chains with 1000 iterations
-each after 1000 warmup iterations. We estimated the relationship between
-population average leaf thickness and \(\mathrm{AA}\) measured from a
-single individual per population. We used this approach most of the
-variation in leaf thickness was among sites and the plant selected for
-gas exchange measurements was not always among the plants randomly
-selected for leaf thickness, precluding individual level correlation. We
-propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness by
-integrating over the entire posterior distribution sample for each
-variable. The key prediction is that the effect of leaf thickness on
-\(\mathrm{AA}\) is positive.
+each after 1000 warmup iterations per chain. We estimated the
+relationship between population average leaf thickness and
+\(\mathrm{AA}\) measured from a single individual per population. We
+used this approach because most of the variation in leaf thickness was
+among sites and the plant selected for gas exchange measurements was not
+always among the plants randomly selected for leaf thickness, precluding
+individual level correlation. We propagated uncertainty about in
+\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
+entire posterior distribution sample for each variable. The key
+prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
+positive, meaning the 95\% confidence interval of the slope should be
+positive and not encompass 0.
 
@@
 \textbf{brms} package. We sampled the posterior distribution from 4
-chains with 1000 iterations each after 1000 warmup iterations. From each
-posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:
+chains with 1000 iterations each after 1000 warmup iterations per chain.
+From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
+as:
 
 \[g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},\]
+
 where \(g_\mathrm{smax,lower}\) and \(g_\mathrm{smax,upper}\) are
@@
 \(T_\mathrm{leaf} = 25 ^ \circ\) C on the lower and upper surface,
-respectively. The maximum stomatal conductance is calculated from
+respectively. The maximum stomatal conductance was calculated from
 stomatal density and length, assuming that stomata are fully open,
-following Sack and Buckley (2016).
+following Sack and Buckley (2016):
+
+\[g_\mathrm{smax} = bmds^{0.5}.\] In this equation, \(b\) is a
+biophysical constant, \(m\) is a morphological constant, \(d\) is the
+stomatal density, and \(s\) is the stomatal complex area. We assume that
+\(b\), which is determined by the molecular species, temperature, and
+air pressure, is the same for both surfaces; we assume that \(m\), which
+is determined by guard cell allometry is also the same for both
+surfaces. Hence, the \(b\) and \(m\) constants cancel out of
+\(g_\mathrm{smax,ratio}\) and only density and length (\(l\)), which is
+proportional to the square root of area, affect the ratio:
+\(g_\mathrm{smax} \propto dl\).
 
@@
 \(\mathrm{AA}\) measured from a single leaf per population. We
-propagated uncertainty about in \(\mathrm{AA}\) and
+propagated uncertainty about \(\mathrm{AA}\) and
 \(g_\mathrm{smax,ratio}\) by integrating over the entire posterior
@@
 effect of \(g_\mathrm{smax,ratio}\) on \(\mathrm{AA}\) is positive until
+\(g_\mathrm{smax,ratio} < 0.5\), meaning the 95\% confidence interval of
+the slope should be positive and not encompass 0 in the domain
 \(g_\mathrm{smax,ratio} < 0.5\).
@@
 \section{RESULTS ---}\label{results}}
+
+Coastal ʻilima are surrounded by shorter vegetation than their montane
+counterparts (\autoref{fig:study-system}d; Welch Two Sample
+\emph{t}-test, \(t_{5.41} = 4.29\), \(P = 0.007\)). The montane site
+with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
+sanctuary) in a matrix of cattle pasture, hence the satellite derived
+vegetation height may be lower than what existed prior to human
+disturbance. Habitats did not differ significantly in average solar
+radiation at the top of the canopy (\autoref{fig:study-system}d; Welch
+Two Sample \emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)). Coastal
+habitats are significantly warmer (\autoref{fig:study-system}d; Welch
+Two Sample \emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)) but mean
+annual precipitation is similar (\autoref{fig:study-system}d; Welch Two
+Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).
 
@@
 \(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
-(\autoref{fig:habitat-Ags}).
+(\autoref{fig:habitat-Ags}). Water-use efficiency
+(\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
+(Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).
 
@@
 This allows us to quantity the amphistomy advantage (\(\mathrm{AA}\))
-holding all else constant. Taking advantage of the of the steep climatic
+holding all else constant. Taking advantage of the steep climatic
 gradients in the Hawaiian archipelago, we applied this new method to
@@
 (\(\mathrm{AA}_\text{coastal} = 0.12\)
-vs.~\(\mathrm{AA}_\text{montane} = 0.027\)). While this result has not
-yet been validated in other species, our results indicate that part of
-the reason amphistomatous leaves are found most commonly in high light
-habitats is that the adaptive benefit is greater in this environment.
+vs.~\(\mathrm{AA}_\text{montane} = 0.027\)). Coastal and montane ʻilima
+leaves are likely good representatives of classic sun and shade leaf
+syndromes because 1) they vary in traits like reflective pubescence
+(Ehleringer and Björkman, 1978) and leaf thickness (Terashima et al.,
+2001) that typically characterize sun-shade adaptation; and 2) since
+ʻilima shrubs are typically \(<1\)m tall, they are shaded by trees in
+montane, but not coastal habitats (\autoref{fig:study-system}d). While
+this result has not yet been validated in other species, our results
+indicate that part of the reason amphistomatous leaves are found most
+commonly in high light habitats is that the adaptive benefit is greater
+in this environment.
 
@@
 within habitats is actually slightly negative
-(\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and
+(\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and
 montane leaves differ in many respects besides thickness, we do not have
@@
 why \(\mathrm{AA}\) is greater in coastal leaves. The negative
-relationship between leaf thickness and \(\mathrm{AA}\) could be
+relationship, (albeit nonsignificant in that 95\% confidence intervals
+encompassed 0), between leaf thickness and \(\mathrm{AA}\) could be
 explained if thicker leaves compensated by having a more porous
@@
 leaves on Oʻahu had high \(g_\mathrm{smax,ratio}\), but similarly low
-\(\mathrm{AA}\) (\autoref{fig:traits-aa}B). Among coastal sites, the
+\(\mathrm{AA}\) (\autoref{fig:traits-aa}b). Among coastal sites, the
 site with the lowest \(g_\mathrm{smax,ratio}\) had the lowest
@@
 
-Two limitations of this study are the small sample size and experimental
+Three limitations of this study are the small sample size, experimental
 design that precludes distinguishing genetic from environmental
-differences in leaf traits. Understanding the mechanistic basis of
-higher \(\mathrm{AA}\) in sun leaves would require much larger sample
-sizes. Sun leaves tend to be thicker, more densely packed with mesophyll
-cells, and have greater photosynthetic capacity and higher stomatal
-conductance, among other traits (Lambers et al., 2008). Each of these
-factors and others potentially modulate \(\mathrm{AA}\). Quantifying the
-contribution of all these factors requires larger samples and additional
-measurements that are beyond the scope of this study, but exciting
-avenues for future research on leaf structure-function relations.
-Although many morphological traits that distinguish coastal and montane
-ʻilima populations persist in a common environment (Yorkston and
-Daehler, 2006), we cannot distinguish between genetic effects and
-plastic responses to habitat as causes of difference in \(\mathrm{AA}\)
-because we measured naturally occurring plants \emph{in situ}. While
-disentangling genetic and plastic contributions is not necessarily
-important for understanding the distribution of amphistomatous leaves,
-it would be insightful to know about genetic and environmental
-contributions to trait variation. A reciprocal transplant would be able
-to determine the genetic and environmental contributions, as well their
-interaction, to trait variance in nature.
+differences in leaf traits, and potentially confounding effects of other
+environmental differences besides light environment. Understanding the
+mechanistic basis of higher \(\mathrm{AA}\) in sun leaves would require
+much larger sample sizes. Sun leaves tend to be thicker, more densely
+packed with mesophyll cells, and have greater photosynthetic capacity
+and higher stomatal conductance, among other traits (Lambers et al.,
+2008). Each of these factors and others potentially modulate
+\(\mathrm{AA}\). Quantifying the contribution of all these factors
+requires larger samples and additional measurements that are beyond the
+scope of this study, but exciting avenues for future research on leaf
+structure-function relations. Although many morphological traits that
+distinguish coastal and montane ʻilima populations persist in a common
+environment (Yorkston and Daehler, 2006), we cannot distinguish between
+genetic effects and plastic responses to habitat as causes of difference
+in \(\mathrm{AA}\) because we measured naturally occurring plants
+\emph{in situ}. While disentangling genetic and plastic contributions is
+not necessarily important for understanding the distribution of
+amphistomatous leaves, it would be insightful to know about genetic and
+environmental contributions to trait variation. A reciprocal transplant
+would be able to determine the genetic and environmental contributions,
+as well their interaction, to trait variance in nature. However,
+reciprocal transplants cannot control for other differences between
+coastal and montane habitats besides vegetation height, such as
+temperature. Experimental studies in controlled environments will be
+necessary to isolate the effect light quantity and quality on
+\(\mathrm{AA}\).
 
@@
 type of open, sunny habitats where this trait is most common. By
-developing a straightforward gas exchange method to block gas exchange
+developing a straightforward experimental method to block gas exchange
 through the upper surface, we directly compared the photosynthetic rate
@@
 photosynthetic benefit from amphistomy than nearby montane leaves living
-in more closed forest. We cannot yet ascribe the difference in
-amphistomy advantage between coastal and montane leaves to particular
-physiological or anatomical variation, but this is a promising area for
-future research.
+in more closed forest. This is not because amphistomatous leaves
+necessarily have greater leaf surface available for stomata, although
+that likely influences realized photosynthetic rates in natural
+populations. Rather, our experiments show that coastal amphistomatous
+leaves with the same total leaf stomatal conductance photosynthesize
+more than identical hypostomatous leaves. We cannot yet ascribe the
+difference in amphistomy advantage between coastal and montane leaves to
+particular physiological or anatomical variation, but this is a
+promising area for future research.
 
@@
 this manuscript, the University of Hawaiʻi honors council for guidance
-to GT, Tawn Speetjens for access to one montane site, startup funds from
-the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
-2307341 to TNB.
+to GT, Tawn Speetjens for access to one montane site, TM Perez for
+advice on leaf sectioning, startup funds from the University of Hawaiʻi,
+NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
 
@@
 \setcounter{equation}{0}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:ags-curve}
+\end{figure}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-blonder_leaf-area_2012}{}}%
+Blonder, B., V. Buzzard, I. Simova, L. Sloat, B. Boyle, R. Lipson, B.
+Aguilar-Beaucage, et al. 2012.
+\href{https://doi.org/10.3732/ajb.1200062}{The leaf-area shrinkage
+effect can bias paleoclimate and ecology research}. \emph{American
+Journal of Botany} 99: 1756--1763.
+
 \leavevmode\vadjust pre{\hypertarget{ref-de_boer_apparent_2016}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-drake_smaller_2013}{}}%
+Drake, P. L., R. H. Froend, and P. J. Franks. 2013.
+\href{https://doi.org/10.1093/jxb/ers347}{Smaller, faster stomata:
+Scaling of stomatal size, rate of response, and stomatal conductance}.
+\emph{Journal of Experimental Botany} 64: 495--505.
+
 \leavevmode\vadjust pre{\hypertarget{ref-dunbar-co_leaf_2009}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-flexas_cell_2021}{}}%
+Flexas, J., M. J. Clemente-Moreno, J. Bota, T. J. Brodribb, J. Gago, Y.
+Mizokami, M. Nadal, et al. 2021.
+\href{https://doi.org/10.1093/jxb/erab144}{Cell wall thickness and
+composition are involved in photosynthetic limitation} P. Manavella
+{[}ed.{]},. \emph{Journal of Experimental Botany} 72: 3971--3986.
+
 \leavevmode\vadjust pre{\hypertarget{ref-foster_influence_1986}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gago_photosynthesis_2020}{}}%
+Gago, J., D. M. Daloso, M. Carriquí, M. Nadal, M. Morales, W. L. Araújo,
+A. Nunes-Nesi, et al. 2020.
+\href{https://doi.org/10.1016/j.envexpbot.2020.104174}{The
+photosynthesis game is in the "inter-play": {Mechanisms} underlying
+{CO2} diffusion in leaves}. \emph{Environmental and Experimental Botany}
+178: 104174.
+
 \leavevmode\vadjust pre{\hypertarget{ref-gay_influence_1975}{}}%
@@
 37--46.
+
+\leavevmode\vadjust pre{\hypertarget{ref-giambelluca_online_2013}{}}%
+Giambelluca, T. W., Q. Chen, A. G. Frazier, J. P. Price, Y.-L. Chen,
+P.-S. Chu, J. K. Eischeid, and D. M. Delparte. 2013.
+\href{https://doi.org/10.1175/BAMS-D-11-00228.1}{Online {Rainfall}
+{Atlas} of {Hawai}`i}. \emph{Bulletin of the American Meteorological
+Society} 94: 313--316.
+
+\leavevmode\vadjust pre{\hypertarget{ref-giambelluca_evapotranspiration_2014}{}}%
+Giambelluca, T. W., X. Shuai, M. L. Barnes, R. J. Alliss, R. J. Longman,
+T. Miura, Q. Chen, et al. 2014. Evapotranspiration of {Hawai}`i.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hedges_meta-analysis_1999}{}}%
+Hedges, L. V., J. Gurevitch, and P. S. Curtis. 1999. The meta-analysis
+of response ratios in experimental ecology. \emph{Ecology} 80:
+1150--1156.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hetherington_role_2003}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-marquez_cuticular_2022}{}}%
+Márquez, D. A., H. Stuart‐Williams, G. D. Farquhar, and F. A. Busch.
+2022. \href{https://doi.org/10.1111/nph.17588}{Cuticular conductance of
+adaxial and abaxial leaf surfaces and its relation to minimum leaf
+surface conductance}. \emph{New Phytologist} 233: 156--168.
+
 \leavevmode\vadjust pre{\hypertarget{ref-mckown_trait_2016}{}}%
@@
 Muir, C. D. 2019. \href{https://doi.org/10.1093/icb/icz085}{Is
-{Amphistomy} an {Adaptation} to {High} {Light}? {Optimality} {Models} of
-{Stomatal} {Traits} along {Light} {Gradients}}. \emph{Integrative and
-Comparative Biology} 59: 571--584.
+amphistomy an adaptation to high light? {Optimality} models of stomatal
+traits along light gradients}. \emph{Integrative and Comparative
+Biology} 59: 571--584.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-terashima_irradiance_2006}{}}%
+Terashima, I., Y. T. Hanba, Y. Tazoe, P. Vyas, and S. Yano. 2006.
+Irradiance and phenotype: Comparative eco-development of sun and shade
+leaves in relation to photosynthetic {CO}₂ diffusion. \emph{Journal of
+Experimental Botany} 57: 343--354.
+
+\leavevmode\vadjust pre{\hypertarget{ref-terashima_why_2001}{}}%
+Terashima, I., S.-I. Miyazawa, and Y. T. Hanba. 2001.
+\href{https://doi.org/10.1007/PL00013972}{Why are sun leaves thicker
+than shade leaves? --- {Consideration} based on analyses of
+{CO}\(_{\textrm{2}}\) diffusion in the leaf}. \emph{Journal of Plant
+Research} 114: 93--105.
+
 \leavevmode\vadjust pre{\hypertarget{ref-theroux-rancourt_maximum_2021}{}}%

### 2023-10-29 ms/si.tex
revised ms sent to coauthors

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 

### 2023-10-31 ms/ms.qmd
addressed TNB's feedback on revision

@@
 
-\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy".
-We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).
+\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy".
+We used humidity to modulate stomatal conductance and thus compare photosynthetic rates at the same total stomatal conductance.
+We estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).
 
@@
 A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through transpiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997].
-Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021].
+Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_eco-evolutionary_2021].
 Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019].
@@
 We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods).
+We use humidity to modulate stomatal conductance so that amphi- and pseudohypostomatous leaves can be compared at the same total stomatal conductance.
 One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], which will often result in lower $g_\mathrm{ias}$ either by increasing the diffusion path length [@parkhurst_adaptive_1978] or making the airspace less porous.
@@
 
-The authors thank Kasey Barton for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
+The authors thank Kasey Barton and two anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
 

### 2023-10-31 ms/ms.tex
addressed TNB's feedback on revision

@@
 \usepackage{lineno}
-\linenumbers
+% \linenumbers
 
@@
 \usepackage{setspace}
-% \singlespacing
-\doublespacing
+\singlespacing
+% \doublespacing
 
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \textbf{Methods} We developed a new method to quantify ``amphistomy
-advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
-amphistomatous leaf to that of the same leaf with the same total
-stomatal conductance but with gas exchange blocked through the upper
-(adaxial) surface, which we term ``pseudohypostomy''. We thus estimated
-\(\mathrm{AA}\) and related physiological and anatomical traits in 12
-populations, six coastal (open, sunny) and six montane (closed, shaded),
-of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).
+advantage'', \(\mathrm{AA}\), as the log-ratio of photosynthesis in an
+amphistomatous leaf to that of the same leaf but with gas exchange
+blocked through the upper (adaxial) surface, which we term
+``pseudohypostomy''. We used humidity to modulate stomatal conductance
+and thus compare photosynthetic rates at the same total stomatal
+conductance. We estimated \(\mathrm{AA}\) and related physiological and
+anatomical traits in 12 populations, six coastal (open, sunny) and six
+montane (closed, shaded), of the indigenous Hawaiian species ʻilima
+(\emph{Sida fallax}).
 
@@
 pseudohypostomatous leaf with gas exchange blocked through the upper
-surface (see Materials and Methods). One reason that sun leaves might
-have greater \(\mathrm{AA}\) is that they are usually thicker or denser
-(Poorter et al., 2019), which will often result in lower
-\(g_\mathrm{ias}\) either by increasing the diffusion path length
+surface (see Materials and Methods). We use humidity to modulate
+stomatal conductance so that amphi- and pseudohypostomatous leaves can
+be compared at the same total stomatal conductance. One reason that sun
+leaves might have greater \(\mathrm{AA}\) is that they are usually
+thicker or denser (Poorter et al., 2019), which will often result in
+lower \(g_\mathrm{ias}\) either by increasing the diffusion path length
 (Parkhurst, 1978) or making the airspace less porous. A nonmutually
@@
 
-The authors thank Kasey Barton for feedback on an earlier version of
-this manuscript, the University of Hawaiʻi honors council for guidance
-to GT, Tawn Speetjens for access to one montane site, TM Perez for
-advice on leaf sectioning, startup funds from the University of Hawaiʻi,
-NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
+The authors thank Kasey Barton and two anonymous reviewers for feedback
+on an earlier version of this manuscript, the University of Hawaiʻi
+honors council for guidance to GT, Tawn Speetjens for access to one
+montane site, TM Perez for advice on leaf sectioning, startup funds from
+the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
+2307341 to TNB.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-harrison_ecoevolutionary_2021}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-harrison_eco-evolutionary_2021}{}}%
 Harrison, S. P., W. Cramer, O. Franklin, I. C. Prentice, H. Wang, Å.
 Brännström, H. Boer, et al. 2021.
-\href{https://doi.org/10.1111/nph.17558}{Eco‐evolutionary optimality as
-a means to improve vegetation and land‐surface models}. \emph{New
+\href{https://doi.org/10.1111/nph.17558}{Eco-evolutionary optimality as
+a means to improve vegetation and land-surface models}. \emph{New
 Phytologist} 231: 2125--2141.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harwood_understanding_2021}{}}%
-Harwood, R., G. Théroux‐Rancourt, and M. M. Barbour. 2021.
+Harwood, R., G. Théroux-Rancourt, and M. M. Barbour. 2021.
 \href{https://doi.org/10.1111/pce.14079}{Understanding airspace in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-marquez_assessing_2023}{}}%
-Márquez, D. A., H. Stuart‐Williams, L. A. Cernusak, and G. D. Farquhar.
+Márquez, D. A., H. Stuart-Williams, L. A. Cernusak, and G. D. Farquhar.
 2023. \href{https://doi.org/10.1111/nph.18784}{Assessing the
-{\textless{}}span style="font-variant:small-caps;"{\textgreater{}} {CO}
-\(_{\textrm{2}}\) {\textless{}}/span{\textgreater{}} concentration at
-the surface of photosynthetic mesophyll cells}. \emph{New Phytologist}
-238: 1446--1460.
+{CO}\(_{\textrm{2}}\) concentration at the surface of photosynthetic
+mesophyll cells}. \emph{New Phytologist} 238: 1446--1460.
 
 \leavevmode\vadjust pre{\hypertarget{ref-marquez_cuticular_2022}{}}%
-Márquez, D. A., H. Stuart‐Williams, G. D. Farquhar, and F. A. Busch.
+Márquez, D. A., H. Stuart-Williams, G. D. Farquhar, and F. A. Busch.
 2022. \href{https://doi.org/10.1111/nph.17588}{Cuticular conductance of
@@
 Irradiance and phenotype: Comparative eco-development of sun and shade
-leaves in relation to photosynthetic {CO}₂ diffusion. \emph{Journal of
-Experimental Botany} 57: 343--354.
+leaves in relation to photosynthetic {CO}\(_{\textrm{2}}\) diffusion.
+\emph{Journal of Experimental Botany} 57: 343--354.
 

### 2023-11-01 r/02_process-trait-data.R
cleaned up makefile and raw climate data

@@ -1,12 +1,6 @@
 # Join data sheets for analysis
 source("r/header.R")
 
-# Climate rasters
-# Extract climate data
-dir = "raw-data/climate"
-tair_ann = raster::raster(glue("{dir}/Tair_month_raster/tair_ann"))
-sl_mst_ann = raster::raster(glue("{dir}/SoilMoisture_month_raster/sl_mst_ann"))
-
 # Read in raw data ----
 site = read_csv("raw-data/site.csv", col_types = "ccccccDdddd")
 
@@ -87,19 +81,6 @@ assert_numeric(leaf_thickness$leaf_thickness_pixels, lower = 0, upper = 500,
 assert_numeric(leaf_thickness$leaf_thickness_um, lower = 0, upper = 500, 
                any.missing = FALSE)
 
-# Add climate data to site ----
-site1 = site |>
-  mutate(
-    tair_ann = raster::extract(
-      tair_ann, 
-      cbind(longitude_degree, latitude_degree)
-    ),
-    sl_mst_ann = raster::extract(
-      sl_mst_ann, 
-      cbind(longitude_degree, latitude_degree)
-    )
-  )
-
 # Summarize multiple measurements ----
 stomata_size1 = stomata_size |>
   group_by(site_code, individual, leaf_number, surface) |>
@@ -124,7 +105,7 @@ stomata_density |>
   full_join(stomata_size1, by = c("site_code", "individual", "leaf_number")) |>
   full_join(leaf_thickness1, by = c("site_code", "individual")) |>
   full_join(plant, by = c("site_code", "individual")) |>
-  left_join(site1, by = "site_code", suffix = c("_plant", "_site")) |>
+  left_join(site, by = "site_code", suffix = c("_plant", "_site")) |>
   left_join(licor_leaf, by = join_by(site_code, individual, leaf_number)) |>
   mutate(licor_leaf = ifelse(is.na(licor_leaf), FALSE, licor_leaf)) |>
   

### 2023-11-01 r/16_fit-habitat-climate.R
fixed habitat-climate script

@@ -1,6 +1,10 @@
 source("r/header.R")
 
 trait = read_rds("processed-data/trait.rds") |>
+  full_join(read_rds("processed-data/site.rds"),
+            by = join_by(site_code, site, label, alias, site_type, island,
+                         date_sampled, latitude_degree, longitude_degree,
+                         leaf_area_chamber_cm2)) |>
   summarize(
     site_type = first(site_type),
     `MAT~group('[', paste(phantom()*degree, 'C'), ']')` = first(tair_ann),

### 2023-11-01 ms/ms.qmd
slightly revised habitat-climate results based on updated analysis

@@
 Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
-The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right).
+The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right).
 MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
@@
 The montane site with the lowest vegetation height is a remnant dry forest (Koaiʻa tree sanctuary) in a matrix of cattle pasture, hence the satellite derived vegetation height may be lower than what existed prior to human disturbance.
-Habitats did not differ significantly in average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$).
-Coastal habitats are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$) but mean annual precipitation is similar (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).
+Coastal sites receive greater average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$); coastal sites are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$); and coastal sites receive less precipitation (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).
 
@@
 A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.
-However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature.
+However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton.
 Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.

### 2023-11-01 ms/ms.tex
slightly revised habitat-climate results based on updated analysis

@@
 \usepackage{lineno}
-% \linenumbers
+\linenumbers
 
@@
 \usepackage{setspace}
-\singlespacing
-% \doublespacing
+% \singlespacing
+\doublespacing
 
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 vegetation height may be lower than what existed prior to human
-disturbance. Habitats did not differ significantly in average solar
-radiation at the top of the canopy (\autoref{fig:study-system}d; Welch
-Two Sample \emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)). Coastal
-habitats are significantly warmer (\autoref{fig:study-system}d; Welch
-Two Sample \emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)) but mean
-annual precipitation is similar (\autoref{fig:study-system}d; Welch Two
+disturbance. Coastal sites receive greater average solar radiation at
+the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample
+\emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)); coastal sites are
+significantly warmer (\autoref{fig:study-system}d; Welch Two Sample
+\emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)); and coastal sites
+receive less precipitation (\autoref{fig:study-system}d; Welch Two
 Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).
@@
 coastal and montane habitats besides vegetation height, such as
-temperature. Experimental studies in controlled environments will be
-necessary to isolate the effect light quantity and quality on
-\(\mathrm{AA}\).
+temperature and precipitaiton. Experimental studies in controlled
+environments will be necessary to isolate the effect light quantity and
+quality on \(\mathrm{AA}\).
 

### 2023-11-03 ms/ms.qmd
revised ms

@@
 \textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts.
-The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic basis of this observation is an area for future research.
+The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis of this observation is an area for future research.
 
@@
 We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve.
-To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements.
+To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements, though we did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).
 In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum.
@@
 We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo).
-We did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).
 
@@
 The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf.
-$\mathrm{AA}$ for each leaf was calculated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ using the log-response ratio shown above.
+We estimated $\mathrm{AA}$ from $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ for each leaf using the log-response ratio shown above.
 
@@
 It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain.
-When we remove two leaves where extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.
+When we remove two leaves where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.
 Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}).
@@
 Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why.
-The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf.
+The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* improves photosynthesis in a given leaf.
 By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
@@
 Coastal and montane ʻilima leaves are likely good representatives of classic sun and shade leaf syndromes because 1) they vary in traits like reflective pubescence [@ehleringer_pubescence_1978] and leaf thickness [@terashima_why_2001] that typically characterize sun-shade adaptation; and 2) since ʻilima shrubs are typically $<1$m tall, they are shaded by trees in montane, but not coastal habitats (\autoref{fig:study-system}d).
-While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.
+While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in such environments.
 
@@
 Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves.
-The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].
+The negative relationship, albeit nonsignificant in that 95\% confidence intervals encompassed 0, between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].
 
@@
 Under these conditions, amphistomy may substantially increase photosynthesis, as we observe in coastal ʻilima leaves.
-Selection on increased photosynthesis under similar conditions may explain why crops leaves tend to increase stomatal density ratio during domestication [@milla_shifts_2013].
+Selection on increased photosynthesis under similar conditions may explain why crop leaves tend to increase stomatal density ratio during domestication [@milla_shifts_2013].
 
@@
 However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton.
-Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.
+Experimental studies in controlled environments will be necessary to isolate the effects of light quantity and quality on $\mathrm{AA}$.
 

### 2023-11-03 ms/ms.tex
revised ms

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
+\textbf{Key results} Coastal ʻilima leaves benefit 4.04 times more from
 amphistomy compared to their montane counterparts. Our evidence was
@@
 amphistomatous leaves in open, sunny habitats can partially be explained
-the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic
+the increased benefit of amphistomy in `sun' leaves, but the mechanistic
 basis of this observation is an area for future research.
@@
 for order effects, we alternated between starting with amphi or
-pseudohypo leaf measurements. In the field, we acclimated the focal leaf
-to saturating light and high relative humidity (\(\mathrm{RH} = 70\%\)),
-as described above, until \(A\) and \(g_\text{sw}\) reach their maximum.
-We used these data as our estimates of maximum \(A\) and
-\(g_\text{sw}\). After that, we decreased \(\mathrm{RH}\) to
-\(\approx 10\%\) to induce rapid stomatal closure without biochemical
-downregulation. Hence, \(A_\text{amphi}\) and \(A_\text{hypo}\) were
-both measured at low chamber humidity after the leaf had acclimated to
-high humidity. All other environmental conditions in the leaf chamber
-remained the same. We logged data until \(g_\text{sw}\) reached its
-nadir. We then repeated the process of acclimating the leaf to 70\%
-\(\mathrm{RH}\) and inducing stomatal closure with low \(\mathrm{RH}\)
-with the other treatment (amphi or pseudohypo). We did not detect an
-effect of treatment order on \(\mathrm{AA}\) (results not shown).
+pseudohypo leaf measurements, though we did not detect an effect of
+treatment order on \(\mathrm{AA}\) (results not shown). In the field, we
+acclimated the focal leaf to saturating light and high relative humidity
+(\(\mathrm{RH} = 70\%\)), as described above, until \(A\) and
+\(g_\text{sw}\) reach their maximum. We used these data as our estimates
+of maximum \(A\) and \(g_\text{sw}\). After that, we decreased
+\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
+without biochemical downregulation. Hence, \(A_\text{amphi}\) and
+\(A_\text{hypo}\) were both measured at low chamber humidity after the
+leaf had acclimated to high humidity. All other environmental conditions
+in the leaf chamber remained the same. We logged data until
+\(g_\text{sw}\) reached its nadir. We then repeated the process of
+acclimating the leaf to 70\% \(\mathrm{RH}\) and inducing stomatal
+closure with low \(\mathrm{RH}\) with the other treatment (amphi or
+pseudohypo).
 
@@
 curves. The vertical lines in \autoref{fig:licor} show the
-\(g_\text{sw}\) for each leaf. \(\mathrm{AA}\) for each leaf was
-calculated \(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) using
+\(g_\text{sw}\) for each leaf. We estimated \(\mathrm{AA}\) from
+\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) for each leaf using
 the log-response ratio shown above.
@@
 curve for each leaf, we fit Bayesian regressions using the \emph{R}
-package \textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling
+package \textbf{brms} version 2.20.4 (Bürkner, 2017) with MCMC sampling
 in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
-2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
+2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
 interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
@@
 individual plant and site as random effects. We used the \emph{R}
-package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
-\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
+package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit the model in
+\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.1
 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
@@
 individual plant, individual plant, and site as random effects. We used
-the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
+the \emph{R} package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit
 the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
-version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
+version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
 2023). We interpolated missing adaxial guard cell lengths from 6 out of
@@
 counterparts (\autoref{fig:study-system}d; Welch Two Sample
-\emph{t}-test, \(t_{5.41} = 4.29\), \(P = 0.007\)). The montane site
+\emph{t}-test, \(t_{6.67} = 5.13\), \(P = 0.002\)). The montane site
 with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
@@
 the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample
-\emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)); coastal sites are
+\emph{t}-test, \(t_{10.86} = -2.22\), \(P = 0.049\)); coastal sites are
 significantly warmer (\autoref{fig:study-system}d; Welch Two Sample
-\emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)); and coastal sites
+\emph{t}-test, \(t_{6.01} = -2.96\), \(P = 0.025\)); and coastal sites
 receive less precipitation (\autoref{fig:study-system}d; Welch Two
-Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).
+Sample \emph{t}-test, \(t_{7.45} = 2.73\), \(P = 0.028\)).
 
@@
 leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
-\(0.027~[\numrange{-0.0044}{0.057}]\), respectively; the difference in
+\(0.027~[\numrange{-0.0034}{0.057}]\), respectively; the difference in
 average \(\mathrm{AA}\) between habitat types is
-\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.089~[\numrange{0.04}{0.14}]\).
+\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.039}{0.14}]\).
 Posterior predictions closely match observed values of \(A\)
@@
 slight extrapolation beyond the data should be reliable, but this is
-less certain. When we remove two leaves where extrapolated slightly
+less certain. When we remove two leaves where we extrapolated slightly
 beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
 \(\mathrm{AA}_\text{coastal}\) is still positive,
-\(0.081~[\numrange{0.022}{0.14}]\), but the difference between coastal
-and montane leaves is smaller, \(0.053~[\numrange{-0.011}{0.12}]\), and
+\(0.081~[\numrange{0.023}{0.13}]\), but the difference between coastal
+and montane leaves is smaller, \(0.053~[\numrange{-0.012}{0.12}]\), and
 confidence intervals slightly overlap 0. Maximum photosynthetic rate was
@@
 (\autoref{fig:traits-aa}A;
-\(\text{slope}~= -0.11~[\numrange{-0.27}{0.036}]\)).
+\(\text{slope}~= -0.11~[\numrange{-0.28}{0.035}]\)).
 
@@
 (\autoref{fig:traits-aa}B;
-\(\text{slope}~= 0.14~[\numrange{-0.06}{0.35}]\)) in our sample.
+\(\text{slope}~= 0.14~[\numrange{-0.057}{0.34}]\)) in our sample.
 
@@
 more than shade leaves, but progress has been limited by the lack of
-evidence that amphiostomy \emph{per se} affects improves photosynthesis
-in a given leaf. By experimentally blocking gas exchange through the
-upper surface in a controlled environment, we directly compared an
+evidence that amphiostomy \emph{per se} improves photosynthesis in a
+given leaf. By experimentally blocking gas exchange through the upper
+surface in a controlled environment, we directly compared an
 amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
@@
 gradients in the Hawaiian archipelago, we applied this new method to
-show for the first time that sun leaves benefit 4 times more from
+show for the first time that sun leaves benefit 4.04 times more from
 amphistomy than shade leaves on ʻilima (\emph{Sida fallax}) plants
@@
 commonly in high light habitats is that the adaptive benefit is greater
-in this environment.
+in such environments.
 
@@
 why \(\mathrm{AA}\) is greater in coastal leaves. The negative
-relationship, (albeit nonsignificant in that 95\% confidence intervals
-encompassed 0), between leaf thickness and \(\mathrm{AA}\) could be
+relationship, albeit nonsignificant in that 95\% confidence intervals
+encompassed 0, between leaf thickness and \(\mathrm{AA}\) could be
 explained if thicker leaves compensated by having a more porous
@@
 leaves. Selection on increased photosynthesis under similar conditions
-may explain why crops leaves tend to increase stomatal density ratio
+may explain why crop leaves tend to increase stomatal density ratio
 during domestication (Milla et al., 2013).
@@
 temperature and precipitaiton. Experimental studies in controlled
-environments will be necessary to isolate the effect light quantity and
-quality on \(\mathrm{AA}\).
+environments will be necessary to isolate the effects of light quantity
+and quality on \(\mathrm{AA}\).
 

### 2023-11-03 ms/si.tex
revised ms

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 

### 2023-11-03 r/04_fit-licor.R

@@ -2,17 +2,6 @@ source("r/header.R")
 
 licor = read_rds("processed-data/licor.rds")
 
-# fit_licor = brm(bf(
-#   A ~ site_code_aperture + s(gsw) + s(gsw, by = site_code_aperture)),
-#   data = licor,
-#   chains = 4L,
-#   cores = 4L,
-#   seed = 744874146,
-#   backend = "cmdstanr",
-#   control = list(max_treedepth = 11),
-#   silent = 0
-# )
-
 fit_licor = brm(bf(
   A ~ log(gsw) * site_code_aperture, sigma ~ 1 + (1|site_code_aperture)),
   data = licor,

### 2023-11-03 r/06_estimate-aa.R

@@ -18,9 +18,8 @@ licor2 = licor |>
   select(-matches("^(max|min)_(one|two)-sided$")) |>
   mutate(i = row_number())
 
-# fit_licor_draws = prepare_predictions(fit_licor)
-# ndraw = fit_licor_draws$ndraws
-ndraw = 4000
+fit_licor_draws = prepare_predictions(fit_licor)
+ndraw = fit_licor_draws$ndraws
 
 pred_licor2 = predict(fit_licor, new = licor2, summary = FALSE, ndraws = ndraw)
 

### 2023-11-03 r/07_plot-licor.R

@@ -111,8 +111,6 @@ ggplot(licor,
   ylab(expression(paste(
     italic(A), " [", mu, "mol ", m ^ -2 ~ s ^ -1, "]"
   ))) +
-  # xlim(0, 0.5) +
-  # ylim(0, 40) +
   theme_classic() +
   theme(
     axis.text = element_text(size = 10),

### 2023-12-13 ms/ms.qmd
preparing final submission

@@
 A widespread association between amphistomy and open, sunny habitats suggests the adaptive benefit of amphistomy may be greatest in these contexts, but this hypothesis has not been tested experimentally.
-Understanding why amphistomy evolves can inform  its potential as a target for crop improvement and paleoenvironment reconstruction.
+Understanding amphistomy informs its potential as a target for crop improvement and paleoenvironment reconstruction.
 
-\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy".
-We used humidity to modulate stomatal conductance and thus compare photosynthetic rates at the same total stomatal conductance.
-We estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).
+\textbf{Methods}  We developed a method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper surface (pseudohypostomy).
+Humidity modulated stomatal conductance and thus enabled comparing photosynthesis at the same total stomatal conductance.
+We estimated $\mathrm{AA}$ and leaf traits in six coastal (open, sunny) and six montane (closed, shaded) populations of the indigenous Hawaiian species ʻilima (*Sida fallax*).
 
-\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts.
-Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most of the conductance being on the lower (abaxial) surface.
+\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than montane leaves.
+Evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because 1) they are thicker and have lower CO$_2$ conductance through the internal airspace, and 2) that they benefit more because they have similar conductance on each surface, as opposed to most conductance being through the lower surface.
 
-\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts.
-The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis of this observation is an area for future research.
+\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase CO$_2$ supply to chloroplasts.
+The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis remains uncertain.
 
-**Keywords:** amphstomy, leaf, light, Hawaiʻi, *Sida fallax*, stomata
+**Keywords:** amphistomy, Hawaiʻi, leaf, light, Malvaceae, photosynthesis, *Sida fallax*, stomata
 
@@
 This montane morphology is not found in *S.
-fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022].
+fallax* populations on other Pacific Islands [@pejhanmehr_low_2023].
 
@@
 Specifically, the objectives of our study are to test whether 1) sun leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than shade leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?
-
-```{=tex}
-
-\begin{figure}[H]
-  \includegraphics[height=3.25in]{../figures/study-system.pdf}
-  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
-  \caption{A.
-Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
-leaves.
-C.
-Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
-D.
-Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
-Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
-The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right).
-MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
-  \label{fig:study-system}
-\end{figure}
-
-```
 
@@
 We tested whether climatic variables differed among our coastal and montane populations using Welch's two-sample $t$-test.
-
-```{r, locations, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-locations
-#| tbl-cap: "ʻIlima study site location information."
-#| 
-read_rds("../processed-data/site.rds") |>
-  filter(site_code != "nnpl") |>
-  mutate(Island = case_when(
-    island == "oahu" ~ "Oʻahu",
-    island == "hawaii" ~ "Hawaiʻi"
-  )) |>
-  arrange(site_type, site) |>
-  select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
-         Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
-  knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
-               booktabs = TRUE) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "4cm") |>
-  kableExtra::column_spec(2, width = "1.5cm") |>
-  kableExtra::column_spec(3, width = "1.5cm") |>
-  kableExtra::column_spec(4, width = "1.5cm") |>
-  kableExtra::column_spec(5, width = "1.5cm") |>
-  kableExtra::column_spec(6, width = "1.5cm")
-
-```
 
@@
 
-At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA).
+At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
 We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$.
@@
 
-\autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves.
+Appendix S1 (see the Supplementary Data with this article), \autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves.
 We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve.
@@
 In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves.
-The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf.
+The vertical lines in Appendix S1, \autoref{fig:licor} show the $g_\text{sw}$ for each leaf.
 We estimated $\mathrm{AA}$ from $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ for each leaf using the log-response ratio shown above.
@@
 
-<!-- \begin{align*} -->
-
-<!--   b = & D_\text{wv} / v \\ -->
-
-<!--   m = & \frac{\pi c ^ 2}{j ^ {0.5} (4 h j + \pi c)} -->
-
-<!-- \end{align*} -->
-
-<!-- $b$ is the diffusion coefficient of water vapor in air ($D_\text{wv}$) divided by the kinematic viscosity of dry air ($v$).
-$D_\text{wv} = 2.49 \times 10^{-5} ~ \text{m}^2 ~ \text{s}^{-1}$ and $v = 2.24 \times 10 ^ {-2} ~ \text{m}^3 ~ \text{mol} ^ {-1}$ at $25 ^ \circ$ [@monteith_principles_2013].
-For kidney-shaped guard cells, $c = h = j = 0.5$.
--->
-
 We estimated the relationship between leaf $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ measured from a single leaf per population.
@@
 Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants.
-$\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual curves).
+$\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see Appendix S1, \autoref{fig:licor} for individual curves).
 Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$.
-Posterior predictions closely match observed values of $A$ (\autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably.
+Posterior predictions closely match observed values of $A$ (Appendix S1, \autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably.
 It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain.
 When we remove two leaves where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.
-Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}).
+Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (Appendix S1, \autoref{fig:habitat-Ags}).
 Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).
-
-```{=tex}
-\begin{figure}[H]
-  \includegraphics{../figures/habitat-aa.pdf}
-  \caption{Coastal leaves benefit more from amphistomy than montane leaves.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf.
-Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat.
-The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves.
-$g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:habitat-aa}
-\end{figure}
-```
 
@@
 
-Coastal and montane leaves have similar average $g_\mathrm{smax,ratio}$, the ratio of adaxial (upper) to total $g_\mathrm{smax}$, the anatomical maximum stomatal conductance to water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have $`r gmaxratio_sitetype`$ higher $g_\mathrm{smax,ratio}$ than montane leaves, but the 95% confidence intervals overlap 0 difference.
+Coastal and montane leaves have similar average $g_\mathrm{smax,ratio}$, the ratio of adaxial (upper) to total $g_\mathrm{smax}$, the anatomical maximum stomatal conductance to water vapor (Appendix S1, \autoref{fig:habitat-gmaxratio}); coastal leaves have $`r gmaxratio_sitetype`$ higher $g_\mathrm{smax,ratio}$ than montane leaves, but the 95% confidence intervals overlap 0 difference.
 The $g_\mathrm{smax,ratio}$ is somewhat bimodal among sites.
-Some sites in both habitats have leaves with $g_\mathrm{smax,ratio} < 0.07$ and others with $g_\mathrm{smax,ratio} > 0.2$ (\autoref{fig:habitat-gmaxratio}).
+Some sites in both habitats have leaves with $g_\mathrm{smax,ratio} < 0.07$ and others with $g_\mathrm{smax,ratio} > 0.2$ (Appendix S1, \autoref{fig:habitat-gmaxratio}).
 This is particularly noticeable in montane sites where those on the Big Island of Hawaiʻi all have low $g_\mathrm{smax,ratio}$ whereas those on Oʻahu have relatively high $g_\mathrm{smax,ratio}$.
 There is no relationship between $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ in either habitat (\autoref{fig:traits-aa}B; $\text{slope}~= `r gmaxratio_aa`$) in our sample.
-
-```{=tex}
-\begin{figure}[H]
-  \includegraphics{../figures/traits-aa.pdf}
-  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value.
-Each leaf is from a different montane (blue) or coastal (orange) site.
-Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
-Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:traits-aa}
-\end{figure}
-```
 
@@
 
-The authors thank Kasey Barton and two anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
+The authors thank Kasey Barton and three anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.
+This is publication \#213 from the School of Life Sciences, University of Hawaiʻi at Mānoa.
 
@@
 
-Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo with a DOI and stable URL upon publication.
-Raw data will be deposited on Dryad with a DOI and stable URL upon publication.
-[THE GITHUB REPO AND DRYAD DATA ARE AVAILABLE TO REVIEWERS]
+Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw
+).
 
@@
 * Appendix S1: Supplemental figures and table
+
+\newpage
+
+# Literature Cited --- {#literature-cited}
+
+::: {#refs}
+:::
+
+\clearpage
+
+```{r, locations, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-locations
+#| tbl-cap: "ʻIlima study site location information."
+#| 
+read_rds("../processed-data/site.rds") |>
+  filter(site_code != "nnpl") |>
+  mutate(Island = case_when(
+    island == "oahu" ~ "Oʻahu",
+    island == "hawaii" ~ "Hawaiʻi"
+  )) |>
+  arrange(site_type, site) |>
+  select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
+         Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
+  knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
+               booktabs = TRUE) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "4cm") |>
+  kableExtra::column_spec(2, width = "1.5cm") |>
+  kableExtra::column_spec(3, width = "1.5cm") |>
+  kableExtra::column_spec(4, width = "1.5cm") |>
+  kableExtra::column_spec(5, width = "1.5cm") |>
+  kableExtra::column_spec(6, width = "1.5cm")
+
+```
+
+# Figure Legends --- {#figure-legends}
+
+```{=tex}
+
+\begin{figure}[H]
+  % \includegraphics[height=3.25in]{../figures/study-system.pdf}
+  % \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
+  \caption{A.
+Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
+leaves.
+C.
+Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
+D.
+Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
+Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
+The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right).
+MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
+  \label{fig:study-system}
+\end{figure}
+
+\begin{figure}[H]
+  % \includegraphics{../figures/habitat-aa.pdf}
+  \caption{Coastal leaves benefit more from amphistomy than montane leaves.
+A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf.
+Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat.
+The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves.
+$g_\mathrm{sw}$, stomatal conductance to water vapor.}
+  \label{fig:habitat-aa}
+\end{figure}
+
+\begin{figure}[H]
+  % \includegraphics{../figures/traits-aa.pdf}
+  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
+A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value.
+Each leaf is from a different montane (blue) or coastal (orange) site.
+Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
+Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
+  \label{fig:traits-aa}
+\end{figure}
+```
 
@@
 ```
-
-\newpage
-
-# Literature Cited --- {#literature-cited}

### 2023-12-13 ms/ms.tex
preparing final submission

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 these contexts, but this hypothesis has not been tested experimentally.
-Understanding why amphistomy evolves can inform its potential as a
-target for crop improvement and paleoenvironment reconstruction.
-
-\textbf{Methods} We developed a new method to quantify ``amphistomy
+Understanding amphistomy informs its potential as a target for crop
+improvement and paleoenvironment reconstruction.
+
+\textbf{Methods} We developed a method to quantify ``amphistomy
 advantage'', \(\mathrm{AA}\), as the log-ratio of photosynthesis in an
 amphistomatous leaf to that of the same leaf but with gas exchange
-blocked through the upper (adaxial) surface, which we term
-``pseudohypostomy''. We used humidity to modulate stomatal conductance
-and thus compare photosynthetic rates at the same total stomatal
-conductance. We estimated \(\mathrm{AA}\) and related physiological and
-anatomical traits in 12 populations, six coastal (open, sunny) and six
-montane (closed, shaded), of the indigenous Hawaiian species ʻilima
-(\emph{Sida fallax}).
+blocked through the upper surface (pseudohypostomy). Humidity modulated
+stomatal conductance and thus enabled comparing photosynthesis at the
+same total stomatal conductance. We estimated \(\mathrm{AA}\) and leaf
+traits in six coastal (open, sunny) and six montane (closed, shaded)
+populations of the indigenous Hawaiian species ʻilima (\emph{Sida
+fallax}).
 
 \textbf{Key results} Coastal ʻilima leaves benefit 4.04 times more from
-amphistomy compared to their montane counterparts. Our evidence was
-equivocal with respect to two hypotheses -- that coastal leaves benefit
-more because 1) they are thicker and therefore have lower CO\(_2\)
-conductance through the internal airspace, and 2) that they benefit more
-because they have similar conductance on each surface, as opposed to
-most of the conductance being on the lower (abaxial) surface.
+amphistomy than montane leaves. Evidence was equivocal with respect to
+two hypotheses -- that coastal leaves benefit more because 1) they are
+thicker and have lower CO\(_2\) conductance through the internal
+airspace, and 2) that they benefit more because they have similar
+conductance on each surface, as opposed to most conductance being
+through the lower surface.
 
 \textbf{Conclusions} This is the first direct experimental evidence that
-amphistomy \emph{per se} increases photosynthesis, consistent with the
-hypothesis that parallel pathways through upper and lower mesophyll
-increase the supply of CO\(_2\) to chloroplasts. The prevalence of
-amphistomatous leaves in open, sunny habitats can partially be explained
-the increased benefit of amphistomy in `sun' leaves, but the mechanistic
-basis of this observation is an area for future research.
-
-\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, \emph{Sida fallax},
-stomata
+amphistomy increases photosynthesis, consistent with the hypothesis that
+parallel pathways through upper and lower mesophyll increase CO\(_2\)
+supply to chloroplasts. The prevalence of amphistomatous leaves in open,
+sunny habitats can partially be explained the increased benefit of
+amphistomy in `sun' leaves, but the mechanistic basis remains uncertain.
+
+\textbf{Keywords:} amphistomy, Hawaiʻi, leaf, light, Malvaceae,
+photosynthesis, \emph{Sida fallax}, stomata
 
@@
 not found in \emph{S. fallax} populations on other Pacific Islands
-(Pejhanmehr, 2022).
+(Pejhanmehr et al., 2023).
 
@@
 have a \(g_{\mathrm{smax,ratio}}\) closer to 0.5?
-
-\begin{figure}[H]
-  \includegraphics[height=3.25in]{../figures/study-system.pdf}
-  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
-  \caption{A.
-Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
-leaves.
-C.
-Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
-D.
-Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
-Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
-The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right).
-MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
-  \label{fig:study-system}
-\end{figure}
 
@@
 
-\hypertarget{tbl-locations}{}
-\begin{table}
-\caption{\label{tbl-locations}ʻIlima study site location information. }\tabularnewline
-
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
-\toprule
-Site & Island & Habitat & Latitude & Longitude & Elevation (mas)\\
-\midrule
-\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.710} & \cellcolor{gray!6}{-157.982} & \cellcolor{gray!6}{4}\\
-Kaloko beach & Oʻahu & coastal & 21.293 & -157.661 & 4\\
-\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{19.676} & \cellcolor{gray!6}{-156.024} & \cellcolor{gray!6}{6}\\
-Kaʻena Point & Oʻahu & coastal & 21.574 & -158.278 & 4\\
-\cellcolor{gray!6}{Makapuʻu beach} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.313} & \cellcolor{gray!6}{-157.661} & \cellcolor{gray!6}{3}\\
-\addlinespace
-Puakō petroglyph park & Hawaiʻi & coastal & 19.957 & -155.858 & 8\\
-\cellcolor{gray!6}{Hawaiʻi loa ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.294} & \cellcolor{gray!6}{-157.727} & \cellcolor{gray!6}{352}\\
-Hāloa ʻĀina & Hawaiʻi & montane & 19.552 & -155.793 & 1567\\
-\cellcolor{gray!6}{Kaʻohe game management area} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{19.817} & \cellcolor{gray!6}{-155.616} & \cellcolor{gray!6}{1946}\\
-Koaiʻa tree sanctuary & Hawaiʻi & montane & 20.048 & -155.737 & 970\\
-\addlinespace
-\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.305} & \cellcolor{gray!6}{-157.779} & \cellcolor{gray!6}{248}\\
-Waʻahila ridge & Oʻahu & montane & 21.314 & -157.793 & 357\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \hypertarget{leaf-traits}{%
@@
 portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
-NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
-conductance to water vapor (\(g_\text{sw}\)) at saturating light
+Nebraska, USA). We estimated the photosynthetic rate (\(A\)) and
+stomatal conductance to water vapor (\(g_\text{sw}\)) at saturating
+light
 (\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
@@
 
+Appendix S1 (see the Supplementary Data with this article),
 \autoref{fig:ags-curve} illustrates our method for collecting
@@
 beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
-curves. The vertical lines in \autoref{fig:licor} show the
+curves. The vertical lines in Appendix S1, \autoref{fig:licor} show the
 \(g_\text{sw}\) for each leaf. We estimated \(\mathrm{AA}\) from
@@
 in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
-2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
+2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023) to
 interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
@@
 \emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.1
-and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
-sampled the posterior distribution from 4 chains with 1000 iterations
-each after 1000 warmup iterations per chain. We estimated the
-relationship between population average leaf thickness and
-\(\mathrm{AA}\) measured from a single individual per population. We
-used this approach because most of the variation in leaf thickness was
-among sites and the plant selected for gas exchange measurements was not
-always among the plants randomly selected for leaf thickness, precluding
-individual level correlation. We propagated uncertainty about in
-\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
-entire posterior distribution sample for each variable. The key
-prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
-positive, meaning the 95\% confidence interval of the slope should be
-positive and not encompass 0.
+and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023). We sampled the
+posterior distribution from 4 chains with 1000 iterations each after
+1000 warmup iterations per chain. We estimated the relationship between
+population average leaf thickness and \(\mathrm{AA}\) measured from a
+single individual per population. We used this approach because most of
+the variation in leaf thickness was among sites and the plant selected
+for gas exchange measurements was not always among the plants randomly
+selected for leaf thickness, precluding individual level correlation. We
+propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness
+estimates by integrating over the entire posterior distribution sample
+for each variable. The key prediction is that the effect of leaf
+thickness on \(\mathrm{AA}\) is positive, meaning the 95\% confidence
+interval of the slope should be positive and not encompass 0.
 
@@
 the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
-version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
-2023). We interpolated missing adaxial guard cell lengths from 6 out of
-185 samples with zero adaxial stomata using the ``mi'' function in
+version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023).
+We interpolated missing adaxial guard cell lengths from 6 out of 185
+samples with zero adaxial stomata using the ``mi'' function in
 \textbf{brms} package. We sampled the posterior distribution from 4
@@
 coastal leaves, but only 1 of 6 montane leaves
-(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
-curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
-leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
+(\autoref{fig:habitat-aa}; see Appendix S1, \autoref{fig:licor} for
+individual curves). Overall, the average \(\mathrm{AA}\) among coastal
+and montane leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
 \(0.027~[\numrange{-0.0034}{0.057}]\), respectively; the difference in
@@
 \(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.039}{0.14}]\).
-Posterior predictions closely match observed values of \(A\)
-(\autoref{fig:pp-licor}), indicating an adequate model fit from which we
-can interpolate between measurements reliably. It also suggests that
+Posterior predictions closely match observed values of \(A\) (Appendix
+S1, \autoref{fig:pp-licor}), indicating an adequate model fit from which
+we can interpolate between measurements reliably. It also suggests that
 slight extrapolation beyond the data should be reliable, but this is
@@
 \(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
-(\autoref{fig:habitat-Ags}). Water-use efficiency
+(Appendix S1, \autoref{fig:habitat-Ags}). Water-use efficiency
 (\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
 (Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).
-
-\begin{figure}[H]
-  \includegraphics{../figures/habitat-aa.pdf}
-  \caption{Coastal leaves benefit more from amphistomy than montane leaves.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf.
-Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat.
-The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves.
-$g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:habitat-aa}
-\end{figure}
 
@@
 \(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
-water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
-\(0.059~[\numrange{-0.14}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
-montane leaves, but the 95\% confidence intervals overlap 0 difference.
-The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
-sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
-and others with \(g_\mathrm{smax,ratio} > 0.2\)
-(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
+water vapor (Appendix S1, \autoref{fig:habitat-gmaxratio}); coastal
+leaves have \(0.059~[\numrange{-0.14}{0.28}]\) higher
+\(g_\mathrm{smax,ratio}\) than montane leaves, but the 95\% confidence
+intervals overlap 0 difference. The \(g_\mathrm{smax,ratio}\) is
+somewhat bimodal among sites. Some sites in both habitats have leaves
+with \(g_\mathrm{smax,ratio} < 0.07\) and others with
+\(g_\mathrm{smax,ratio} > 0.2\) (Appendix S1,
+\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
 montane sites where those on the Big Island of Hawaiʻi all have low
@@
 \(\text{slope}~= 0.14~[\numrange{-0.057}{0.34}]\)) in our sample.
-
-\begin{figure}[H]
-  \includegraphics{../figures/traits-aa.pdf}
-  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
-A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
-Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value.
-Each leaf is from a different montane (blue) or coastal (orange) site.
-Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
-Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
-  \label{fig:traits-aa}
-\end{figure}
 
@@
 
-The authors thank Kasey Barton and two anonymous reviewers for feedback
-on an earlier version of this manuscript, the University of Hawaiʻi
-honors council for guidance to GT, Tawn Speetjens for access to one
-montane site, TM Perez for advice on leaf sectioning, startup funds from
-the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
-2307341 to TNB.
+The authors thank Kasey Barton and three anonymous reviewers for
+feedback on an earlier version of this manuscript, the University of
+Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to
+one montane site, TM Perez for advice on leaf sectioning, startup funds
+from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award
+2307341 to TNB. This is publication \#213 from the School of Life
+Sciences, University of Hawaiʻi at Mānoa.
 
@@
 Custom scripts are available on a GitHub repository
-(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
-with a DOI and stable URL upon publication. Raw data will be deposited
-on Dryad with a DOI and stable URL upon publication. {[}THE GITHUB REPO
-AND DRYAD DATA ARE AVAILABLE TO REVIEWERS{]}
+(https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw
+data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw ).
 
@@
 
-\renewcommand\thefigure{S\arabic{figure}}    
-\renewcommand\thetable{S\arabic{table}}    
-\renewcommand\theequation{S\arabic{equation}}    
-\setcounter{figure}{0}    
-\setcounter{table}{0}    
-\setcounter{equation}{0}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:ags-curve}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:licor}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:pp-licor}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:habitat-Ags}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{labelformat=empty}
-  \caption{}
-  \label{fig:habitat-gmaxratio}
-\end{figure}
-
-\begin{table}
-  \captionsetup{labelformat=empty}
-  \caption{}
-
-\end{table}
-
 \newpage
@@
 \hypertarget{literature-cited}{%
-\section*{Literature Cited ---}\label{literature-cited}}
-\addcontentsline{toc}{section}{Literature Cited ---}
+\section{Literature Cited ---}\label{literature-cited}}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
-Gabry, J., and R. Češnovar. 2023.
+Gabry, J., R. Češnovar, and A. Johnson. 2023.
 \href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_phylogenetic_2022}{}}%
-Pejhanmehr, M. 2022. Phylogenetic affinities among
-{\textless{}}i{\textgreater{}}{Sida}{\textless{}}{\textbackslash{}}i{\textgreater{}}
-species and allied genera ({Malvaceae}: {Malveae}), and examination of
-{\textless{}}i{\textgreater{}}{Sida}
-fallax{\textless{}}{\textbackslash{}}i{\textgreater{}} within the
-{Hawaiian} {Islands} and throughout the {Pacific}. PhD thesis.
-University of Hawai'i at Mānoa.
+\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_low_2023}{}}%
+Pejhanmehr, M., M. Yorkston, and C. W. Morden. 2023.
+\href{https://doi.org/10.2984/76.4.2}{Low {Genetic} {Diversity} in the
+{Highly} {Morphologically} {Diverse} {Sida} fallax {Walp}. ({Malvaceae})
+{Throughout} the {Pacific1}}. \emph{Pacific Science} 76.
 
@@
 
+\clearpage
+
+\hypertarget{tbl-locations}{}
+\begin{table}
+\caption{\label{tbl-locations}ʻIlima study site location information. }\tabularnewline
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
+\toprule
+Site & Island & Habitat & Latitude & Longitude & Elevation (mas)\\
+\midrule
+\cellcolor{gray!6}{Kahuku Point} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.710} & \cellcolor{gray!6}{-157.982} & \cellcolor{gray!6}{4}\\
+Kaloko beach & Oʻahu & coastal & 21.293 & -157.661 & 4\\
+\cellcolor{gray!6}{Kaloko-Honokōhau national historical park} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{19.676} & \cellcolor{gray!6}{-156.024} & \cellcolor{gray!6}{6}\\
+Kaʻena Point & Oʻahu & coastal & 21.574 & -158.278 & 4\\
+\cellcolor{gray!6}{Makapuʻu beach} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{coastal} & \cellcolor{gray!6}{21.313} & \cellcolor{gray!6}{-157.661} & \cellcolor{gray!6}{3}\\
+\addlinespace
+Puakō petroglyph park & Hawaiʻi & coastal & 19.957 & -155.858 & 8\\
+\cellcolor{gray!6}{Hawaiʻi loa ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.294} & \cellcolor{gray!6}{-157.727} & \cellcolor{gray!6}{352}\\
+Hāloa ʻĀina & Hawaiʻi & montane & 19.552 & -155.793 & 1567\\
+\cellcolor{gray!6}{Kaʻohe game management area} & \cellcolor{gray!6}{Hawaiʻi} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{19.817} & \cellcolor{gray!6}{-155.616} & \cellcolor{gray!6}{1946}\\
+Koaiʻa tree sanctuary & Hawaiʻi & montane & 20.048 & -155.737 & 970\\
+\addlinespace
+\cellcolor{gray!6}{Mauʻumae Ridge} & \cellcolor{gray!6}{Oʻahu} & \cellcolor{gray!6}{montane} & \cellcolor{gray!6}{21.305} & \cellcolor{gray!6}{-157.779} & \cellcolor{gray!6}{248}\\
+Waʻahila ridge & Oʻahu & montane & 21.314 & -157.793 & 357\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\hypertarget{figure-legends}{%
+\section{Figure Legends ---}\label{figure-legends}}
+
+\begin{figure}[H]
+  % \includegraphics[height=3.25in]{../figures/study-system.pdf}
+  % \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
+  \caption{A.
+Typical growth form of montane (left) and coastal (right) ‘ilima plants and B.
+leaves.
+C.
+Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).
+D.
+Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study.
+Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error.
+The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right).
+MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
+  \label{fig:study-system}
+\end{figure}
+
+\begin{figure}[H]
+  % \includegraphics{../figures/habitat-aa.pdf}
+  \caption{Coastal leaves benefit more from amphistomy than montane leaves.
+A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf.
+Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat.
+The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves.
+$g_\mathrm{sw}$, stomatal conductance to water vapor.}
+  \label{fig:habitat-aa}
+\end{figure}
+
+\begin{figure}[H]
+  % \includegraphics{../figures/traits-aa.pdf}
+  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi.
+A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$.
+Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value.
+Each leaf is from a different montane (blue) or coastal (orange) site.
+Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression.
+Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
+  \label{fig:traits-aa}
+\end{figure}
+
+\renewcommand\thefigure{S\arabic{figure}}    
+\renewcommand\thetable{S\arabic{table}}    
+\renewcommand\theequation{S\arabic{equation}}    
+\setcounter{figure}{0}    
+\setcounter{table}{0}    
+\setcounter{equation}{0}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:ags-curve}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:licor}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:pp-licor}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:habitat-Ags}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{labelformat=empty}
+  \caption{}
+  \label{fig:habitat-gmaxratio}
+\end{figure}
+
+\begin{table}
+  \captionsetup{labelformat=empty}
+  \caption{}
+
+\end{table}
+
 \end{document}

### 2023-12-13 ms/si.qmd
preparing final submission

@@
 ---
+title: "Appendix S1: Supplemental figures and table"
 format:
@@
   \pagestyle{fancy}
-  \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
+  \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2024 – Appendix S1}
+  \addtokomafont{disposition}{\rmfamily}
+  \RedeclareSectionCommand[
+    font=\normalfont\Large]{section}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize\bfseries]{subsection}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize\itshape]{subsubsection}
+  \RedeclareSectionCommand[
+    font=\normalfont\normalsize]{paragraph}
 mainfont: Times New Roman

### 2023-12-13 ms/si.tex
preparing final submission

@@
 \pagestyle{fancy}
-\fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
+\fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2024 – Appendix S1}
+\addtokomafont{disposition}{\rmfamily}
+\RedeclareSectionCommand[
+  font=\normalfont\Large]{section}
+\RedeclareSectionCommand[
+  font=\normalfont\normalsize\bfseries]{subsection}
+\RedeclareSectionCommand[
+  font=\normalfont\normalsize\itshape]{subsubsection}
+\RedeclareSectionCommand[
+  font=\normalfont\normalsize]{paragraph}
 \makeatletter
@@
 \hypersetup{
+  pdftitle={Appendix S1: Supplemental figures and table},
   colorlinks=true,
@@
 
+\title{Appendix S1: Supplemental figures and table}
 \author{}
@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+\maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 

### 2023-12-13 r/17_archive-data.R

@@ -2,12 +2,16 @@ source("r/header.R")
 
 trait = read_rds("processed-data/trait.rds")
 site_Ags = read_rds("objects/site_Ags.rds")
+site = read_rds("processed-data/site.rds")
 
 # Individual level data
-ind_data = trait |> 
+ind_data = trait |>
   mutate(
-    genus = "Sida", species = "fallax", authority = "Walp.",
-    exposition = "natural environment", plant_maturity = "mature",
+    genus = "Sida",
+    species = "fallax",
+    authority = "Walp.",
+    exposition = "natural environment",
+    plant_maturity = "mature",
     leaf_age = "mature"
   ) |>
   select(
@@ -37,14 +41,40 @@ ind_data = trait |>
     leaf_thickness_um,
     licor_leaf
   ) |>
-    left_join(
-      site_Ags |>
-        select(site_code, Tleaf, name, value) |>
-        pivot_wider() |>
-        mutate(licor_leaf = TRUE),
-      by = join_by(site_code, licor_leaf)
-    ) |>
-  select(-licor_leaf)
+  left_join(
+    site_Ags |>
+      select(site_code, Tleaf, name, value) |>
+      pivot_wider() |>
+      mutate(licor_leaf = TRUE),
+    by = join_by(site_code, licor_leaf)
+  ) |>
+  select(-licor_leaf) |>
+  left_join(
+    site |>
+      select(
+        site,
+        site_code,
+        site_type,
+        island,
+        latitude_degree,
+        longitude_degree,
+        date_sampled,
+        tair_ann,
+        sl_mst_ann,
+        cl_sw_ann,
+        veg_ht_ann,
+        rf_ann
+      ),
+    by = join_by(
+      site,
+      site_code,
+      site_type,
+      island,
+      latitude_degree,
+      longitude_degree,
+      date_sampled
+    )
+  )
 
 # Summary table
 ind_data |>

### 2023-12-13 ms/ms.tex
added Zenodo link to ms

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, sharp corners, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 Custom scripts are available on a GitHub repository
-(https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw
-data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw ).
+(https://github.com/cdmuir/stomata-ilima) and archived on Zenodo
+(https://doi.org/10.5281/zenodo.10369114). Raw data are deposited on
+Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw ).
 

### 2023-12-19 ms/ms.qmd
added Zenodo link to ms
fixed a couple typos

@@
 
-where $g_\mathrm{smax,lower}$ and $g_\mathrm{smax,upper}$ are maximum stomatal conductance to water vapor at $T_\mathrm{leaf} = 25 ^ \circ$ C on the lower and upper surface, respectively.
+where $g_\mathrm{smax,lower}$ and $g_\mathrm{smax,upper}$ are the maximum stomatal conductance to water vapor at $T_\mathrm{leaf} = 25 ^ \circ$ C on the lower and upper surface, respectively.
 The maximum stomatal conductance was calculated from stomatal density and length, assuming that stomata are fully open, following @sack_developmental_2016:
@@
 A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.
-However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton.
+However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitation.
 Experimental studies in controlled environments will be necessary to isolate the effects of light quantity and quality on $\mathrm{AA}$.
@@
 
-Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw
+Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (https://doi.org/10.5281/zenodo.10369114).
+Raw data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw
 ).

### 2023-12-19 ms/ms.tex
fixed a couple typos

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, sharp corners, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023) to
-interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
+interface with \emph{R} version 4.3.2 (R Core Team, 2023). We sampled
 the posterior distribution from 4 chains with 1000 iterations each after
@@
 
-where \(g_\mathrm{smax,lower}\) and \(g_\mathrm{smax,upper}\) are
+where \(g_\mathrm{smax,lower}\) and \(g_\mathrm{smax,upper}\) are the
 maximum stomatal conductance to water vapor at
@@
 coastal and montane habitats besides vegetation height, such as
-temperature and precipitaiton. Experimental studies in controlled
+temperature and precipitation. Experimental studies in controlled
 environments will be necessary to isolate the effects of light quantity

### 2024-01-22 ms/si.qmd
copyediting edits for si

@@
 header-includes: |
+  \usepackage{textgreek}
   \usepackage{fancyhdr}
@@
 #| tbl-cap: "Average traits values among ʻilima leaves at each site.
-$\\mathrm{SD}_\\text{abaxial}$ is the stomatal density per mm$^2$ on the abaxial (lower) surface; $\\mathrm{SD}_\\text{adaxial}$ is the stomatal density per mm$^2$ on the adaxial (upper) surface; $\\mathrm{GCL}_\\text{abaxial}$ is the guard cell length in $\\mu$m on the abaxial (lower) surface; $\\mathrm{GCL}_\\text{adaxial}$ is the guard cell length in $\\mu$m on the adaxial (upper) surface; Leaf thickness is the length from upper cuticle to lower cuticle in $\\mu$m; $A$ is the photosynthetic rate in $\\mu \\text{mol CO}_2~\\text{m}^{-2}~\\text{s}^{-1}$; $g_\\mathrm{sw}$ is the stomatal conductance to water vapor in $\\text{mol}~\\text{m}^{-2}~\\text{s}^{-1}$."
+$\\mathrm{SD}_\\text{abaxial}$ is the stomatal density per mm$^2$ on the abaxial (lower) surface; $\\mathrm{SD}_\\text{adaxial}$ is the stomatal density per mm$^2$ on the adaxial (upper) surface; $\\mathrm{GCL}_\\text{abaxial}$ is the guard cell length in \\textmu m on the abaxial (lower) surface; $\\mathrm{GCL}_\\text{adaxial}$ is the guard cell length in \\textmu m on the adaxial (upper) surface; Leaf thickness is the length from upper cuticle to lower cuticle in \\textmu m; $A$ is the photosynthetic rate in \\textmu mol CO$_2~\\text{m}^{-2}~\\text{s}^{-1}$; $g_\\mathrm{sw}$ is the stomatal conductance to water vapor in $\\text{mol}~\\text{m}^{-2}~\\text{s}^{-1}$."
 #| 
@@
   \caption{Idealized method for collecting $A \textendash g_\text{sw}$ curves on either amphi or pseudohypo leaves.
-A) After clamping the leaf into the LI-6800 chamber, it acclimates to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) and humidity ($\mathrm{RH} = 70\%$).
+(A) After clamping the leaf into the LI-6800 chamber, it acclimates to high light ($\mathrm{PPFD} = 2000~\textmu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) and humidity ($\mathrm{RH} = 70\%$).
 This induces stomata to open, increasing $g_\mathrm{sw}$ and $A$ until they reach a maximum.
 We abruptly lower the chamber humidity to $\approx 10\%$ to close stomata and log data (black points) until $g_\mathrm{sw}$ and $A$ reach their nadir.
-B) We fit $A \textendash g_\text{sw}$ curves to logged data points.
+(B) We fit $A \textendash g_\text{sw}$ curves to logged data points.
 The asterisk in both panels indicates the data point used for maximum $A$ and $g_\mathrm{sw}$.
-C) $\mathrm{AA}$ is low (left panel) when the photosynthetic rate of an amphi leaf is similar to a pseudohypo leaf at the same total $g_\mathrm{sw}$ ($x$-axes); large $\mathrm{AA}$ (right panel) is indicated when an amphi leaf has a higher photosynthetic rate than a pseudohypo leaf.
+(C) $\mathrm{AA}$ is low (left panel) when the photosynthetic rate of an amphi leaf is similar to a pseudohypo leaf at the same total $g_\mathrm{sw}$ ($x$-axes); large $\mathrm{AA}$ (right panel) is indicated when an amphi leaf has a higher photosynthetic rate than a pseudohypo leaf.
 Abbreviations: $A$ is the photosynthetic rate; $\mathrm{AA}$ is the amphistomy advantage; $g_\mathrm{sw}$ is the stomatal conductance to water vapor; $\mathrm{PPFD}$ is photosynthetic photon flux density; $\mathrm{RH}$ is relative humidity.}
@@
 In leaves from Kaloko-Honokōhau national historical park and Puakō petroglyph park, we extrapolated sligtly beyond fitted curves because they did not quite overlap.
-Symbols: $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $\mathrm{AA}$, amphistomy advantage; $A$, photosynthetic rate in \textmu mol CO$_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:licor}
@@
   \caption{Posterior predictions (thin grey lines) from fitted $A \textendash g_\mathrm{sw}$ curves closely match the observed distribution (thick black line), indicating that the statistical model adequately captures variation in the response variable over the measured range.
-Symbols: $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $A$, photosynthetic rate in \textmu mol CO$_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:pp-licor}
@@
   \includegraphics{../figures/habitat-Ags.pdf}
-  \caption{The photsynthetic rate (left facet) and stomatal conductance to water vapor (right facet) of montane (blue) and coastal (orange) ʻilima leaves.
+  \caption{The photsynthetic rate (left panel) and stomatal conductance to water vapor (right panel) of montane (blue) and coastal (orange) ʻilima leaves.
 Each point-interval is the median posterior estimate plus 95\% confidence interval of trait value for that habitat.
 Smaller points next to each point-interval are the $g_\mathrm{smax,ratio}$ of individual plants, one per site.
-Symbols: $A$, photosynthetic rate in $\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
+Symbols: $A$, photosynthetic rate in \textmu mol CO$_2~\text{m}^{-2}~\text{s}^{-1}$; $g_\mathrm{sw}$, stomatal conductance to water vapor in $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$.}
   \label{fig:habitat-Ags}

### 2024-01-22 ms/si.tex
copyediting edits for si

@@
 \KOMAoption{captions}{tableheading}
+\usepackage{textgreek}
 \usepackage{fancyhdr}
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 stomatal density per mm\(^2\) on the adaxial (upper) surface;
-\(\mathrm{GCL}_\text{abaxial}\) is the guard cell length in \(\mu\)m on
+\(\mathrm{GCL}_\text{abaxial}\) is the guard cell length in \textmu m on
 the abaxial (lower) surface; \(\mathrm{GCL}_\text{adaxial}\) is the
-guard cell length in \(\mu\)m on the adaxial (upper) surface; Leaf
-thickness is the length from upper cuticle to lower cuticle in \(\mu\)m;
-\(A\) is the photosynthetic rate in
-\(\mu \text{mol CO}_2~\text{m}^{-2}~\text{s}^{-1}\); \(g_\mathrm{sw}\)
-is the stomatal conductance to water vapor in
+guard cell length in \textmu m on the adaxial (upper) surface; Leaf
+thickness is the length from upper cuticle to lower cuticle in
+\textmu m; \(A\) is the photosynthetic rate in \textmu mol
+CO\(_2~\text{m}^{-2}~\text{s}^{-1}\); \(g_\mathrm{sw}\) is the stomatal
+conductance to water vapor in
 \(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\).}\tabularnewline

### 2024-01-22 r/18_plot-ags-curve.R

@@ -5,7 +5,7 @@ x1 = seq(-5, 5, 0.01)
 x2 = seq(5, 15, 0.01)
 
 df = tibble(
-  x = c(x1, x2), 
+  x = c(x1, x2),
   y = c(plogis(x1), 2 * (plogis(rev(x2 - 10))) - 1)
 )
 
@@ -13,7 +13,7 @@ df = tibble(
 gp1 = ggplot(df, aes(x, y)) +
   annotate("rect", xmin = 5, xmax = 15, ymin = -Inf, ymax = Inf, color = NA, fill = "grey") +
   geom_segment(
-    data = filter(df, x %in% 5:15), 
+    data = filter(df, x %in% 5:15),
     mapping = aes(x = x, xend = x, y = y, yend = -Inf),
     linetype = "dashed"
   ) +
@@ -41,7 +41,7 @@ gp3 = df |>
   mutate(y1 = y + rnorm(nrow(.), 0, 0.01)) |>
   ggplot(aes(y, y1)) +
   geom_smooth(color = "grey", method = "lm", formula = 'y ~ x') +
-  geom_point(size = 2) +  
+  geom_point(size = 2) +
   annotate("point", x = 1, y = 1, size = 15, shape = "*") +
   xlab(expression(paste(italic(g)[sw], " [mol ", m ^ -2 ~ s ^ -1, "]"))) +
   ylab(expression(paste(italic(A), " [", mu, "mol ", m ^ -2 ~ s ^ -1, "]"))) +
@@ -56,9 +56,9 @@ plot_row = plot_grid(gp1, gp2, gp3, nrow = 1, rel_widths = c(0.6, 0.05, 0.35),
                      labels = c("A", "", "B"))
 
 ## Add title
-title = ggdraw() + 
+title = ggdraw() +
   draw_label(
-    expression(bold(Idealized~example~of~bolditalic(A)-bolditalic(g)[sw]~curve)),
+    expression(paste(bold(Idealized~example~of~bolditalic(A)), "-", bold(bolditalic(g)[sw]~curve))),
     fontface = 'bold', x = 0, hjust = 0
   ) +
   theme(plot.margin = margin(0, 0, 0, 7))
@@ -102,9 +102,9 @@ gp_aa = ggplot(df_aa, aes(x, y, xend = xend, yend = yend, color = `Gas exchange\
   geom_segment(size = 2, lineend = "round") +
   xlim(-1, 2.5) +
   geom_segment(data = df_aa1, lineend = "round", size = 1.5) +
[…]
